---
title: "Celiac_phage_orf_noCluster_PA_abundance_analysis"
output: html_document
date: "2025-10-20"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(edgeR)
library(glmmTMB)
library(broom.mixed)  # tidy() for mixed models
library(dplyr)
library(purrr)
library(tibble)
```


# Functions to extract timepoint-specific differential abundance/PA results with padjust values
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(broom.mixed)



# Extract timepoint-specific interaction rows and add per-gene FDR
extract_timepoint_results <- function(model_results,
                                      model_name,
                                      gene_col = "orf_id",
                                      adjust_method = "BH") {
  if (is.null(model_results) || nrow(model_results) == 0) {
    return(data.frame())
  }

  timepoint_results <- model_results %>%
    # keep interaction terms of interest
    filter(grepl("Dx.StatusCELIAC:onset_timeline_combined", term)) %>%
    mutate(
      timepoint = gsub("Dx.StatusCELIAC:onset_timeline_combined", "", term),
      model     = model_name,
      # safer than get(): pull from the data frame by name
      gene      = .data[[gene_col]]
    ) %>%
    select(gene, timepoint, estimate, std.error, p.value, conf.low, conf.high, model) %>%
    # per gene x model, adjust p across that gene's timepoints
    group_by(gene, model) %>%
    mutate(
      p.adj = {
        pv <- p.value
        ok <- !is.na(pv)
        out <- rep(NA_real_, length(pv))
        if (any(ok)) out[ok] <- p.adjust(pv[ok], method = adjust_method)
        out
      },
      n_tests_gene = n()  # handy to audit how many timepoints were tested for this gene
    ) %>%
    ungroup()

  timepoint_results
}

# Apply to all result objects you keep in the workspace for a cohort
extract_all_timepoints <- function(cohort_name, adjust_method = "BH") {
  results_list <- list()

  # PA models
  try({
    obj <- paste0(cohort_name, "_glmmTMB_logit_results")
    if (exists(obj)) {
      results_list[["PA_model1"]] <- extract_timepoint_results(
        get(obj), "PA_glmmTMB_logit", gene_col = "orf_id", adjust_method = adjust_method
      )
    }
  }, silent = TRUE)

  # Abundance models
  try({
    obj <- paste0(cohort_name, "_nb_results")
    if (exists(obj)) {
      results_list[["abundance_model1"]] <- extract_timepoint_results(
        get(obj), "abundance_nbinom", gene_col = "orf_id", adjust_method = adjust_method
      )
    }
  }, silent = TRUE)

  if (length(results_list) > 0) {
    do.call(rbind, results_list)
  } else {
    data.frame()
  }
}


```



# load clean metadata
```{r load and clean metadata}

# check if column names of abundance table is consistent with row names of the metadata table
stopifnot(identical(colnames(total.orf.abundance.clean_0.75_0.03), rownames(total.orf.metadata.clean_0.75_0.03)))

# clean metadata
# prepare correct variable format
total.orf.metadata.clean_0.75_0.03$feeding_first_year <- factor(total.orf.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
table(total.orf.metadata.clean_0.75_0.03$feeding_first_year)
# Breast_fed  Formula       Breastmilk_and_formula 
#  100            40                    103
total.orf.metadata.clean_0.75_0.03$HLA.Category <- factor(total.orf.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
table(total.orf.metadata.clean_0.75_0.03$HLA.Category)
# Standard Risk     High Risk   Low/No Risk 
#      123            82            38
total.orf.metadata.clean_0.75_0.03$Sex <- factor(total.orf.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
table(total.orf.metadata.clean_0.75_0.03$Sex)
# Female   Male 
#    194     49 
total.orf.metadata.clean_0.75_0.03$Delivery.Mode <- factor(total.orf.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
table(total.orf.metadata.clean_0.75_0.03$Delivery.Mode)
  # Vaginal C-Section 
  #     180     63 
total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
table(total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
 # 4  5  6  7  8  9 10 11 12 15 36 
 # 4  9 46 64 32 27 29  7  1 15  9
total.orf.metadata.clean_0.75_0.03$Dx.Status <- factor(total.orf.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
table(total.orf.metadata.clean_0.75_0.03$Dx.Status)
 # CELIAC CONTROL 
 #    114    129 
total.orf.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(total.orf.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))
table(total.orf.metadata.clean_0.75_0.03$onset_timeline_combined)
# t0      t0-6     t0-12     t0-18     t0-24     t0-30     t0-36 t0-over42 
# 53        20        40        19        23        25        27        36 


```


# load clean abundant table
```{r create clean abundance tables for three cohorts}

# total cohort
total.orf.abundance.clean_0.75_0.03 <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf/data/total/total.orf.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(total.orf.abundance.clean_0.75_0.03) <- gsub("X","",colnames(total.orf.abundance.clean_0.75_0.03))

total.orf.metadata.clean_0.75_0.03 <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf/data/total/total.orf.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")

# 2419  306


# Set factor levels for US cohort to match total cohort
total.orf.metadata.clean_0.75_0.03$feeding_first_year <- factor(total.orf.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
total.orf.metadata.clean_0.75_0.03$HLA.Category <- factor(total.orf.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
total.orf.metadata.clean_0.75_0.03$Sex <- factor(total.orf.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
total.orf.metadata.clean_0.75_0.03$Delivery.Mode <- factor(total.orf.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
total.orf.metadata.clean_0.75_0.03$Dx.Status <- factor(total.orf.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
total.orf.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(total.orf.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))





# US cohort
US.orf.abundance.clean_0.75_0.03 <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf/data/US/US.orf.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(US.orf.abundance.clean_0.75_0.03) <- gsub("X","",colnames(US.orf.abundance.clean_0.75_0.03))


US.orf.metadata.clean_0.75_0.03 <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf/data/US/US.orf.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")

dim(US.orf.abundance.clean_0.75_0.03)
# 4059  197

# Set factor levels for US cohort to match total cohort
US.orf.metadata.clean_0.75_0.03$feeding_first_year <- factor(US.orf.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
US.orf.metadata.clean_0.75_0.03$HLA.Category <- factor(US.orf.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
US.orf.metadata.clean_0.75_0.03$Sex <- factor(US.orf.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
US.orf.metadata.clean_0.75_0.03$Delivery.Mode <- factor(US.orf.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
US.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(US.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
US.orf.metadata.clean_0.75_0.03$Dx.Status <- factor(US.orf.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
US.orf.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(US.orf.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))

# Italy cohort
Italy.orf.abundance.clean_0.75_0.03 <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf/data/Italy/Italy.orf.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(Italy.orf.abundance.clean_0.75_0.03) <- gsub("X","",colnames(Italy.orf.abundance.clean_0.75_0.03))


Italy.orf.metadata.clean_0.75_0.03 <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf/data/Italy/Italy.orf.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
dim(Italy.orf.abundance.clean_0.75_0.03)
# 2901  111

# Set factor levels for Italy cohort to match total cohort
Italy.orf.metadata.clean_0.75_0.03$feeding_first_year <- factor(Italy.orf.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
Italy.orf.metadata.clean_0.75_0.03$HLA.Category <- factor(Italy.orf.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
Italy.orf.metadata.clean_0.75_0.03$Sex <- factor(Italy.orf.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
Italy.orf.metadata.clean_0.75_0.03$Delivery.Mode <- factor(Italy.orf.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
Italy.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(Italy.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
Italy.orf.metadata.clean_0.75_0.03$Dx.Status <- factor(Italy.orf.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
Italy.orf.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(Italy.orf.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))


```


# create total PA table
```{r generate total PA table}

total.target_cpm      <- 0.5   # e.g., 0.25–1.0 for very sparse data
total.min_reads_floor <- 3L    # guard against singletons

## ---- 1) Align samples ------------------------------------------------------
stopifnot(all(colnames(total.orf.abundance.clean_0.75_0.03) %in% rownames(total.orf.metadata.clean_0.75_0.03)))
# Reorder metadata to match counts columns
total.orf.metadata.clean_0.75_0.03 <- total.orf.metadata.clean_0.75_0.03[colnames(total.orf.abundance.clean_0.75_0.03), , drop = FALSE]

## ---- 2) Build DGEList & normalize (TMMwsp) --------------------------------

total.y <- DGEList(counts = total.orf.abundance.clean_0.75_0.03, samples = total.orf.metadata.clean_0.75_0.03)
# Optional: drop genes with all zeros (purely empty rows)
total.y <- total.y[rowSums(total.y$counts) > 0, , keep.lib.sizes = FALSE]

# Library-size normalization (robust for sparse data)
total.y <- calcNormFactors(total.y, method = "TMMwsp")

## ---- 3) CPM-anchored presence/absence -------------------------------------
# Normalized CPMs (uses effective library sizes = lib.size * norm.factors)
total.cpm_mat <- edgeR::cpm(total.y, normalized.lib.sizes = TRUE)

# Presence rule: CPM >= target_cpm AND raw count >= min_reads_floor
total.present_logical <- (total.cpm_mat >= total.target_cpm) & (total.y$counts >= total.min_reads_floor)

# Convert to 0/1 matrix with dimnames preserved
total.PA <- matrix(
  as.integer(total.present_logical),
  nrow = nrow(total.present_logical),
  ncol = ncol(total.present_logical),
  dimnames = dimnames(total.present_logical)
)

dim(total.PA)
# 2419  306

```


# create US PA table
```{r generate US PA table}

US.target_cpm  <- 0.5   # e.g., 0.25–1.0 for very sparse data
US.min_reads_floor <- 3L    # guard against singletons

## ---- 1) Align samples ------------------------------------------------------
stopifnot(all(colnames(US.orf.abundance.clean_0.75_0.03) %in% rownames(US.orf.metadata.clean_0.75_0.03)))
# Reorder metadata to match counts columns
US.orf.metadata.clean_0.75_0.03 <- US.orf.metadata.clean_0.75_0.03[colnames(US.orf.abundance.clean_0.75_0.03), , drop = FALSE]


## ---- 2) Build DGEList & normalize (TMMwsp) --------------------------------

US.y <- DGEList(counts = US.orf.abundance.clean_0.75_0.03, samples = US.orf.metadata.clean_0.75_0.03)
# Optional: drop genes with all zeros (purely empty rows)
US.y <- US.y[rowSums(US.y$counts) > 0, , keep.lib.sizes = FALSE]


# Library-size normalization (robust for sparse data)
US.y <- calcNormFactors(US.y, method = "TMMwsp")


## ---- 3) CPM-anchored presence/absence -------------------------------------
# Normalized CPMs (uses effective library sizes = lib.size * norm.factors)
US.cpm_mat <- edgeR::cpm(US.y, normalized.lib.sizes = TRUE)


# Presence rule: CPM >= target_cpm AND raw count >= min_reads_floor
US.present_logical <- (US.cpm_mat >= US.target_cpm) & (US.y$counts >= US.min_reads_floor)
table(US.present_logical)
#  FALSE   TRUE 
# 763387  36236

# every non-zero count passed

# Convert to 0/1 matrix with dimnames preserved
US.PA <- matrix(
  as.integer(US.present_logical),
  nrow = nrow(US.present_logical),
  ncol = ncol(US.present_logical),
  dimnames = dimnames(US.present_logical)
)

dim(US.PA)
# 4059  197


```


# create Italy PA table
```{r generate Italy PA table}


Italy.target_cpm      <- 0.5   # e.g., 0.25–1.0 for very sparse data
Italy.min_reads_floor <- 3L    # guard against singletons

## ---- 1) Align samples ------------------------------------------------------
stopifnot(all(colnames(Italy.orf.abundance.clean_0.75_0.03) %in% rownames(Italy.orf.metadata.clean_0.75_0.03)))
# Reorder metadata to match counts columns
Italy.orf.metadata.clean_0.75_0.03 <- Italy.orf.metadata.clean_0.75_0.03[colnames(Italy.orf.abundance.clean_0.75_0.03), , drop = FALSE]

## ---- 2) Build DGEList & normalize (TMMwsp) --------------------------------

Italy.y <- DGEList(counts = Italy.orf.abundance.clean_0.75_0.03, samples = Italy.orf.metadata.clean_0.75_0.03)
# Optional: drop genes with all zeros (purely empty rows)
Italy.y <- Italy.y[rowSums(Italy.y$counts) > 0, , keep.lib.sizes = FALSE]

# Library-size normalization (robItalyt for sparse data)
Italy.y <- calcNormFactors(Italy.y, method = "TMMwsp")

## ---- 3) CPM-anchored presence/absence -------------------------------------
# Normalized CPMs (Italyes effective library sizes = lib.size * norm.factors)
Italy.cpm_mat <- edgeR::cpm(Italy.y, normalized.lib.sizes = TRUE)

# Presence rule: CPM >= target_cpm AND raw count >= min_reads_floor
Italy.present_logical <- (Italy.cpm_mat >= Italy.target_cpm) & (Italy.y$counts >= Italy.min_reads_floor)

# Convert to 0/1 matrix with dimnames preserved
Italy.PA <- matrix(
  as.integer(Italy.present_logical),
  nrow = nrow(Italy.present_logical),
  ncol = ncol(Italy.present_logical),
  dimnames = dimnames(Italy.present_logical)
)

dim(Italy.PA)
# 2901  111

```



######################################################### PA ##################################################


# PA model - glmmTMB (Mixed-effects logistic) - logit link
When: Default choice for pooled-over-time models with within-patient correlation.
Why: True random intercept (and optional random slope), fast, flexible links.
Pros: Handles complex fixed effects; supports cloglog link (family = binomial("cloglog")) if events are very rare.
Cons: Can struggle with extreme sparsity/separation—try GEE or Bayesian if convergence issues persist.

## Total cohort - glmmTMB logit
```{r PA_glmmTMB_logit_total}
library(glmmTMB)
library(broom.mixed)

# Create long-format data for total cohort
total_pa_long <- total.PA %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(total.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

# Fit model for each orf
total_glmmTMB_logit_results <- total_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)


total.adjusted.results <- total_glmmTMB_logit_results %>%
  filter(!is.na(p.value)) %>%
  group_by(term) %>%   # often you adjust per term of interest
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


# Save results
#write.csv(total_glmmTMB_logit_results, "total_PA_glmmTMB_logit_results.csv", row.names = FALSE)

write.csv(total.adjusted.results, "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_Contig_Phrog_compositional/orf_contig_results_figures/total_orf_noCluster_PA_model1_glmmTMB_logit_adjusted.csv", row.names = FALSE)
```


## get differences between control and case at each timepoint
```{r}

library(emmeans)
library(tidyverse)
library(purrr)

# 1) Build models (you already have this as US_models)
total_orf_PA_models <- total_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ tryCatch(
      glmmTMB(
        PA ~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. +
          HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
        family = binomial(link = "logit"), data = .x
      ),
      error = function(e) NULL
    ))
  )

# 2) Extract contrasts WITHOUT adding orf_id inside
extract_contrasts <- function(fit) {
  if (is.null(fit)) return(tibble())

  emm <- tryCatch(
    emmeans(fit, ~ Dx.Status | onset_timeline_combined),
    error = function(e) NULL
  )
  if (is.null(emm)) return(tibble())

  contr <- tryCatch(
    contrast(emm, method = "revpairwise", by = "onset_timeline_combined", adjust = "none"),
    error = function(e) NULL
  )
  if (is.null(contr)) return(tibble())

  s  <- as_tibble(summary(contr))         # includes estimate, SE, df (maybe), p.value
  ci <- as_tibble(confint(contr)) %>%
        select(any_of(c("contrast","onset_timeline_combined","lower.CL","upper.CL")))

  # choose join keys that exist in both
  keys <- intersect(names(s), names(ci))
  if (!all(c("contrast","onset_timeline_combined") %in% keys)) {
    # fall back to safest return (no CI)
    out <- s
  } else {
    out <- left_join(s, ci, by = keys)
  }

  # add odds ratios (on link = logit scale)
  out %>%
    mutate(
      odds_ratio  = exp(estimate),
      conf.low.OR = if ("lower.CL" %in% names(.)) exp(lower.CL) else NA_real_,
      conf.high.OR= if ("upper.CL" %in% names(.)) exp(upper.CL) else NA_real_
    )
}

# 2) Map across your models and unnest
total_orf_PA_contrast_results <- total_orf_PA_models %>%
  mutate(contrasts = map(model, extract_contrasts)) %>%
  select(orf_id, contrasts) %>%
  tidyr::unnest(contrasts)


# 3) BH adjust p-values per timepoint
total_orf_PA_contrast_results <- total_orf_PA_contrast_results %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


total_orf_PA_contrast_results <- total_orf_PA_contrast_results %>%
  filter(!is.na(p.value)) %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

View(total_orf_PA_contrast_results %>% filter(p.adj < 0.05))
# 0 significant orfs

```



## US cohort - glmmTMB logit
```{r PA_glmmTMB_logit_US}
# Create long-format data for US cohort
US_pa_long <- US.PA %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(US.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

# Fit model for each orf
US_glmmTMB_logit_results <- US_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)


US.adjusted.results <- US_glmmTMB_logit_results %>%
  filter(!is.na(p.value)) %>%
  group_by(term) %>%   # often you adjust per term of interest
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


# Save results
#write.csv(US_glmmTMB_logit_results, "US_PA_glmmTMB_logit_results.csv", row.names = FALSE)

write.csv(US.adjusted.results, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_PA_model1_glmmTMB_logit_adjusted.csv", row.names = FALSE)
```


## get differences between control and case at each timepoint
```{r}


library(emmeans)
library(tidyverse)
library(purrr)

# 1) Build models (you already have this as US_models)
US_orf_PA_models <- US_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ tryCatch(
      glmmTMB(
        PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
          HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
        family = binomial(link = "logit"), data = .x
      ),
      error = function(e) NULL
    ))
  )

# 2) Extract contrasts WITHOUT adding orf_id inside
extract_contrasts <- function(fit) {
  if (is.null(fit)) return(tibble())

  emm <- tryCatch(
    emmeans(fit, ~ Dx.Status | onset_timeline_combined),
    error = function(e) NULL
  )
  if (is.null(emm)) return(tibble())

  contr <- tryCatch(
    contrast(emm, method = "revpairwise", by = "onset_timeline_combined", adjust = "none"),
    error = function(e) NULL
  )
  if (is.null(contr)) return(tibble())

  s  <- as_tibble(summary(contr))         # includes estimate, SE, df (maybe), p.value
  ci <- as_tibble(confint(contr)) %>%
        select(any_of(c("contrast","onset_timeline_combined","lower.CL","upper.CL")))

  # choose join keys that exist in both
  keys <- intersect(names(s), names(ci))
  if (!all(c("contrast","onset_timeline_combined") %in% keys)) {
    # fall back to safest return (no CI)
    out <- s
  } else {
    out <- left_join(s, ci, by = keys)
  }

  # add odds ratios (on link = logit scale)
  out %>%
    mutate(
      odds_ratio  = exp(estimate),
      conf.low.OR = if ("lower.CL" %in% names(.)) exp(lower.CL) else NA_real_,
      conf.high.OR= if ("upper.CL" %in% names(.)) exp(upper.CL) else NA_real_
    )
}

# 2) Map across your models and unnest
US_orf_PA_contrast_results <- US_orf_PA_models %>%
  mutate(contrasts = map(model, extract_contrasts)) %>%
  select(orf_id, contrasts) %>%
  tidyr::unnest(contrasts)


# 3) BH adjust p-values per timepoint
US_orf_PA_contrast_results <- US_orf_PA_contrast_results %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


US_orf_PA_contrast_results <- US_orf_PA_contrast_results %>%
  filter(!is.na(p.value)) %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

View(US_orf_PA_contrast_results %>% filter(p.adj < 0.05))
# 0 significant orf

```



## Italy cohort - glmmTMB logit
```{r PA_glmmTMB_logit_Italy}
# Create long-format data for Italy cohort
Italy_pa_long <- Italy.PA %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(Italy.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

# Fit model for each orf
Italy_glmmTMB_logit_results <- Italy_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

Italy.adjusted.results <- Italy_glmmTMB_logit_results %>%
  filter(!is.na(p.value)) %>%
  group_by(term) %>%   # often you adjust per term of interest
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()



# Save results
#write.csv(Italy_glmmTMB_logit_results, "Italy_PA_glmmTMB_logit_results.csv", row.names = FALSE)

write.csv(Italy.adjusted.results, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/Italy_orf_PA_model1_glmmTMB_logit_adjusted.csv", row.names = FALSE)
```


## get differences between control and case at each timepoint
```{r}


library(emmeans)
library(tidyverse)
library(purrr)

# 1) Build models (you already have this as US_models)
Italy_orf_PA_models <- Italy_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ tryCatch(
      glmmTMB(
        PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
          HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
        family = binomial(link = "logit"), data = .x
      ),
      error = function(e) NULL
    ))
  )

# 2) Extract contrasts WITHOUT adding orf_id inside
extract_contrasts <- function(fit) {
  if (is.null(fit)) return(tibble())

  emm <- tryCatch(
    emmeans(fit, ~ Dx.Status | onset_timeline_combined),
    error = function(e) NULL
  )
  if (is.null(emm)) return(tibble())

  contr <- tryCatch(
    contrast(emm, method = "revpairwise", by = "onset_timeline_combined", adjust = "none"),
    error = function(e) NULL
  )
  if (is.null(contr)) return(tibble())

  s  <- as_tibble(summary(contr))         # includes estimate, SE, df (maybe), p.value
  ci <- as_tibble(confint(contr)) %>%
        select(any_of(c("contrast","onset_timeline_combined","lower.CL","upper.CL")))

  # choose join keys that exist in both
  keys <- intersect(names(s), names(ci))
  if (!all(c("contrast","onset_timeline_combined") %in% keys)) {
    # fall back to safest return (no CI)
    out <- s
  } else {
    out <- left_join(s, ci, by = keys)
  }

  # add odds ratios (on link = logit scale)
  out %>%
    mutate(
      odds_ratio  = exp(estimate),
      conf.low.OR = if ("lower.CL" %in% names(.)) exp(lower.CL) else NA_real_,
      conf.high.OR= if ("upper.CL" %in% names(.)) exp(upper.CL) else NA_real_
    )
}

# 2) Map across your models and unnest
Italy_orf_PA_contrast_results <- Italy_orf_PA_models %>%
  mutate(contrasts = map(model, extract_contrasts)) %>%
  select(orf_id, contrasts) %>%
  tidyr::unnest(contrasts)


# 3) BH adjust p-values per timepoint
Italy_orf_PA_contrast_results <- Italy_orf_PA_contrast_results %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


Italy_orf_PA_contrast_results <- Italy_orf_PA_contrast_results %>%
  filter(!is.na(p.value)) %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

View(Italy_orf_PA_contrast_results %>% filter(p.adj < 0.05))
# 0 significant orf

```



## Timepoint-specific Results (every timepoint compare to t0)

```{r PA_timepoint_results}
# Extract timepoint results for all PA models


# Total cohort PA models
if(exists("total_glmmTMB_logit_results")) {

  
  total_pa_timepoints <- extract_timepoint_results(total_glmmTMB_logit_results, "PA_glmmTMB_logit")
  write.csv(total_pa_timepoints, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_PA_glmmTMB_logit_timepoint_results_padj.csv",row.names = FALSE)
  
}



# US cohort PA models
if(exists("US_glmmTMB_logit_results")) {

  
  US_pa_timepoints <- extract_timepoint_results(US_glmmTMB_logit_results, "PA_glmmTMB_logit")
  write.csv(US_pa_timepoints, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_PA_model1_glmmTMB_logit_timepoint_specific_results_padj.csv", row.names = FALSE)
  
  
}



# Italy cohort PA models
if(exists("Italy_glmmTMB_logit_results")) {

  
  Italy_pa_timepoints <- extract_timepoint_results(Italy_glmmTMB_logit_results, "PA_glmmTMB_logit")
  write.csv(Italy_pa_timepoints, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/Italy_orf_PA_model1_glmmTMB_logit_timepoint_specific_results_padj.csv", row.names = FALSE)
  
}


```


# Create heatmap for significant PA results (every timepoint compare to t0)
```{r}

create_orf_heatmap <- function(timepoint_data, cohort_name, output_dir = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/"){
  
  print("test1")
  sig_data <- timepoint_data %>%
    filter(!is.na(p.adj) & p.adj < 0.05)

    timepoint_order <- c("t0-over42", "t0-36", "t0-30", "t0-24", "t0-18", "t0-12", "t0-6", "t0")

    
  print("test2")  
    heatmap_data <- sig_data %>%
                    select(gene, timepoint, estimate) %>%
                    mutate(timepoint = factor(timepoint, levels = timepoint_order)) %>%
                    filter(timepoint %in% timepoint_order) %>%
                    tidyr::pivot_wider(names_from = timepoint, values_from = estimate, values_fill = 0) %>%
                    column_to_rownames("gene")

  print("test3") 
     mat <- as.matrix(heatmap_data)
     
     print("test4") 

     m <- max(abs(mat), na.rm = TRUE)
     
     print("test5") 
      if (!is.finite(m) || m == 0) m <- 1  # safe fallback if all zeros
     
     print("test6") 
      cols   <- colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(101)
      print("test7") 
      breaks <- seq(-m, m, length.out = length(cols) + 1)

      print("test8") 
      p <- pheatmap(
        mat,
        color        = cols,
        breaks       = breaks,      # <- key change: diverging scale centered at 0
        scale        = "none",
        cluster_cols = FALSE,       # keep chronological order
        cluster_rows = TRUE,
        fontsize     = 8,
        main         = paste(cohort_name, "Cohort - ORF PA Analysis\nLog Odds Ratios (CONTROL vs CELIAC)"),
        na_col       = "grey90"
      )

      print("test9") 

      return(p)

}


total_heatmap <- create_orf_heatmap(total_pa_timepoints, "total")
ggsave(total_heatmap,file = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_PA_model1_orf_heatmap_selfmade.pdf",dpi = 600,width = 6,height = 7)


US_heatmap <- create_orf_heatmap(US_pa_timepoints, "US")
ggsave(US_heatmap,file = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/US_PA_model1_orf_heatmap_selfmade.pdf",dpi = 600,width = 6,height = 7)

Italy_heatmap <- create_orf_heatmap(Italy_pa_timepoints, "Italy")
# no significant orf

```


######################################################### Abundance (Negative Binomial GLMM) ##################################################


## Total cohort - Negative Binomial GLMM
```{r abundance_NB_total}
library(edgeR)
library(glmmTMB)

# Prepare DGEList and normalization for total cohort
total.y <- DGEList(counts = total.orf.abundance.clean_0.75_0.03, samples = total.orf.metadata.clean_0.75_0.03)
total.y <- calcNormFactors(total.y, method = "TMMwsp")
total.lib_eff <- with(total.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
total_abundance_long <- total.orf.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "count") %>%
  left_join(total.y$samples %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  dplyr::mutate(offset = log(lib.size * norm.factors))

# Fit NB model for each orf
total_nb_results <- total_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

total.adjusted.results <- total_nb_results %>%
  filter(!is.na(p.value)) %>%
  group_by(term) %>%   # often you adjust per term of interest
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


# Save results
#write.csv(total_nb_results, "total_abundance_NB_results.csv", row.names = FALSE)

write.csv(total.adjusted.results, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_abundance_model1_nbinom_adjusted.csv", row.names = FALSE)
```

## get differences between control and case at each timepoint
```{r}

library(emmeans)
library(tidyverse)
library(purrr)

# 1) Build models (you already have this as US_models)
total_orf_Abundance_models <- total_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ tryCatch(
      glmmTMB(
        count ~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. +
          HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
        family = nbinom2, data = .x
      ),
      error = function(e) NULL
    ))
  )

# 2) Extract contrasts WITHOUT adding orf_id inside
extract_contrasts <- function(fit) {
  if (is.null(fit)) return(tibble())

  emm <- tryCatch(
    emmeans(fit, ~ Dx.Status | onset_timeline_combined),
    error = function(e) NULL
  )
  if (is.null(emm)) return(tibble())

  contr <- tryCatch(
    contrast(emm, method = "revpairwise", by = "onset_timeline_combined", adjust = "none"),
    error = function(e) NULL
  )
  if (is.null(contr)) return(tibble())

  s  <- as_tibble(summary(contr))         # includes estimate, SE, df (maybe), p.value
  ci <- as_tibble(confint(contr)) %>%
        select(any_of(c("contrast","onset_timeline_combined","lower.CL","upper.CL")))

  # choose join keys that exist in both
  keys <- intersect(names(s), names(ci))
  if (!all(c("contrast","onset_timeline_combined") %in% keys)) {
    # fall back to safest return (no CI)
    out <- s
  } else {
    out <- left_join(s, ci, by = keys)
  }

  # add odds ratios (on link = logit scale)
  out %>%
    mutate(
      odds_ratio  = exp(estimate),
      conf.low.OR = if ("lower.CL" %in% names(.)) exp(lower.CL) else NA_real_,
      conf.high.OR= if ("upper.CL" %in% names(.)) exp(upper.CL) else NA_real_
    )
}

# 2) Map across your models and unnest
total_orf_Abundance_contrast_results <- total_orf_Abundance_models %>%
  mutate(contrasts = map(model, extract_contrasts)) %>%
  select(orf_id, contrasts) %>%
  tidyr::unnest(contrasts)


# 3) BH adjust p-values per timepoint
total_orf_Abundance_contrast_results <- total_orf_Abundance_contrast_results %>%
  filter(!is.na(p.value)) %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


View(total_orf_Abundance_contrast_results %>% filter(p.adj < 0.05))
# 2 significant

write.csv(total_orf_Abundance_contrast_results,"../Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_abundance_model1_nbinom_contrast_results_padj.csv", row.names = FALSE)

```


## US cohort - Negative Binomial GLMM
```{r abundance_NB_US}
# Prepare DGEList and normalization for US cohort
US.y <- DGEList(counts = US.orf.abundance.clean_0.75_0.03, samples = US.orf.metadata.clean_0.75_0.03)
US.y <- calcNormFactors(US.y, method = "TMMwsp")
US.lib_eff <- with(US.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
US_abundance_long <- US.orf.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "count") %>%
  left_join(US.y$samples %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  dplyr::mutate(offset = log(lib.size * norm.factors))

# Fit NB model for each orf
US_nb_results <- US_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)


US.adjusted.results <- US_nb_results %>%
  filter(!is.na(p.value)) %>%
  group_by(term) %>%   # often you adjust per term of interest
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()



# Save results
#write.csv(US_nb_results, "US_abundance_NB_results.csv", row.names = FALSE)

write.csv(US.adjusted.results, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_abundance_model1_nbinom_adjusted.csv", row.names = FALSE)
```


## get differences between control and case at each timepoint
```{r}

library(emmeans)
library(tidyverse)
library(purrr)

# 1) Build models (you already have this as US_models)
US_orf_Abundance_models <- US_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ tryCatch(
      glmmTMB(
        count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
          HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
        family = nbinom2, data = .x
      ),
      error = function(e) NULL
    ))
  )

# 2) Extract contrasts WITHOUT adding orf_id inside
extract_contrasts <- function(fit) {
  if (is.null(fit)) return(tibble())

  emm <- tryCatch(
    emmeans(fit, ~ Dx.Status | onset_timeline_combined),
    error = function(e) NULL
  )
  if (is.null(emm)) return(tibble())

  contr <- tryCatch(
    contrast(emm, method = "revpairwise", by = "onset_timeline_combined", adjust = "none"),
    error = function(e) NULL
  )
  if (is.null(contr)) return(tibble())

  s  <- as_tibble(summary(contr))         # includes estimate, SE, df (maybe), p.value
  ci <- as_tibble(confint(contr)) %>%
        select(any_of(c("contrast","onset_timeline_combined","lower.CL","upper.CL")))

  # choose join keys that exist in both
  keys <- intersect(names(s), names(ci))
  if (!all(c("contrast","onset_timeline_combined") %in% keys)) {
    # fall back to safest return (no CI)
    out <- s
  } else {
    out <- left_join(s, ci, by = keys)
  }

  # add odds ratios (on link = logit scale)
  out %>%
    mutate(
      odds_ratio  = exp(estimate),
      conf.low.OR = if ("lower.CL" %in% names(.)) exp(lower.CL) else NA_real_,
      conf.high.OR= if ("upper.CL" %in% names(.)) exp(upper.CL) else NA_real_
    )
}

# 2) Map across your models and unnest
US_orf_Abundance_contrast_results <- US_orf_Abundance_models %>%
  mutate(contrasts = map(model, extract_contrasts)) %>%
  select(orf_id, contrasts) %>%
  tidyr::unnest(contrasts)


# 3) BH adjust p-values per timepoint
US_orf_Abundance_contrast_results <- US_orf_Abundance_contrast_results %>%
  filter(!is.na(p.value)) %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

View(US_orf_Abundance_contrast_results %>% filter(p.adj < 0.05))
# 0 significant

```


## Italy cohort - Negative Binomial GLMM
```{r abundance_NB_Italy}
# Prepare DGEList and normalization for Italy cohort
Italy.y <- DGEList(counts = Italy.orf.abundance.clean_0.75_0.03, samples = Italy.orf.metadata.clean_0.75_0.03)
Italy.y <- calcNormFactors(Italy.y, method = "TMMwsp")
Italy.lib_eff <- with(Italy.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
Italy_abundance_long <- Italy.orf.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "count") %>%
  left_join(Italy.y$samples %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  dplyr::mutate(offset = log(lib.size * norm.factors))

# Fit NB model for each orf
Italy_nb_results <- Italy_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

Italy.adjusted.results <- Italy_nb_results %>%
  filter(!is.na(p.value)) %>%
  group_by(term) %>%   # often you adjust per term of interest
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

# Save results
#write.csv(Italy_nb_results, "Italy_abundance_NB_results.csv", row.names = FALSE)

write.csv(Italy.adjusted.results, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/Italy_orf_abundance_model1_nbinom_adjusted.csv", row.names = FALSE)
```


## get differences between control and case at each timepoint
```{r}

library(emmeans)
library(tidyverse)
library(purrr)

# 1) Build models (you already have this as US_models)
Italy_orf_Abundance_models <- Italy_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ tryCatch(
      glmmTMB(
        count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
          HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
        family = nbinom2, data = .x
      ),
      error = function(e) NULL
    ))
  )

# 2) Extract contrasts WITHOUT adding orf_id inside
extract_contrasts <- function(fit) {
  if (is.null(fit)) return(tibble())

  emm <- tryCatch(
    emmeans(fit, ~ Dx.Status | onset_timeline_combined),
    error = function(e) NULL
  )
  if (is.null(emm)) return(tibble())

  contr <- tryCatch(
    contrast(emm, method = "revpairwise", by = "onset_timeline_combined", adjust = "none"),
    error = function(e) NULL
  )
  if (is.null(contr)) return(tibble())

  s  <- as_tibble(summary(contr))         # includes estimate, SE, df (maybe), p.value
  ci <- as_tibble(confint(contr)) %>%
        select(any_of(c("contrast","onset_timeline_combined","lower.CL","upper.CL")))

  # choose join keys that exist in both
  keys <- intersect(names(s), names(ci))
  if (!all(c("contrast","onset_timeline_combined") %in% keys)) {
    # fall back to safest return (no CI)
    out <- s
  } else {
    out <- left_join(s, ci, by = keys)
  }

  # add odds ratios (on link = logit scale)
  out %>%
    mutate(
      odds_ratio  = exp(estimate),
      conf.low.OR = if ("lower.CL" %in% names(.)) exp(lower.CL) else NA_real_,
      conf.high.OR= if ("upper.CL" %in% names(.)) exp(upper.CL) else NA_real_
    )
}

# 2) Map across your models and unnest
Italy_orf_Abundance_contrast_results <- Italy_orf_Abundance_models %>%
  mutate(contrasts = map(model, extract_contrasts)) %>%
  select(orf_id, contrasts) %>%
  tidyr::unnest(contrasts)


# 3) BH adjust p-values per timepoint
Italy_orf_Abundance_contrast_results <- Italy_orf_Abundance_contrast_results %>%
  filter(!is.na(p.value)) %>%
  group_by(onset_timeline_combined) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()

View(Italy_orf_Abundance_contrast_results %>% filter(p.adj < 0.05))
# 0 significant

```


## create heatmap for model contrast results
```{r}

create_orf_heatmap <- function(timepoint_data, cohort_name, output_dir = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/"){
  
  print("test1")
  sig_data <- timepoint_data %>%
    filter(!is.na(p.adj) & p.adj < 0.05)

    timepoint_order <- c("t0-over42", "t0-36", "t0-30", "t0-24", "t0-18", "t0-12", "t0-6", "t0")

    heatmap_data <- sig_data %>%
                    select(orf_id, onset_timeline_combined, estimate) %>%
                    mutate(onset_timeline_combined = factor(onset_timeline_combined, levels = timepoint_order)) %>%
                    filter(onset_timeline_combined %in% timepoint_order) %>%
                    tidyr::pivot_wider(names_from = onset_timeline_combined, values_from = estimate, values_fill = 0) %>%
                    column_to_rownames("orf_id")


     mat <- as.matrix(heatmap_data)

     m <- max(abs(mat), na.rm = TRUE)
     
      if (!is.finite(m) || m == 0) m <- 1  # safe fallback if all zeros
     
      cols   <- colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(101)

      breaks <- seq(-m, m, length.out = length(cols) + 1)
      
      p <- pheatmap(
        mat,
        color        = cols,
        breaks       = breaks,      # <- key change: diverging scale centered at 0
        scale        = "none",
        cluster_cols = FALSE,       # keep chronological order
        cluster_rows = TRUE,
        fontsize     = 8,
        main         = paste(cohort_name, "Cohort - ORF PA Analysis\nLog Odds Ratios (CONTROL vs CELIAC)"),
        na_col       = "grey90"
      )

      return(p)

}

total_contrast_heatmap <- create_orf_heatmap(total_orf_Abundance_contrast_results, "total")
ggsave(total_contrast_heatmap,file = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_Abundance_contrast_heatmap_selfmade.pdf",dpi = 600,width = 6,height = 7)


```



## Timepoint-specific Results (every timepoint compres to t0)
```{r abundance_timepoint_results}
# Extract timepoint results for all abundance models

# Total cohort abundance timepoints
if(exists("total_nb_results")) {
  total_abundance_timepoints <- extract_timepoint_results(total_nb_results, "abundance_nbinom")
 write.csv(total_abundance_timepoints, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_abundance_model1_nbinom_timepoint_specific_results_padj.csv", row.names = FALSE)
}
# 4 significant


# US cohort abundance timepoints
if(exists("US_nb_results")) {
  US_abundance_timepoints <- extract_timepoint_results(US_nb_results, "abundance_nbinom")
  write.csv(US_abundance_timepoints, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_abundance_model1_nbinom_timepoint_specific_results_padj.csv", row.names = FALSE)
}
# 1 significant

# Italy cohort abundance timepoints
if(exists("Italy_nb_results")) {
  Italy_abundance_timepoints <- extract_timepoint_results(Italy_nb_results, "abundance_nbinom")
  write.csv(Italy_abundance_timepoints, "../Orf_orf_Phrog_compositional/orf_orf_results_figures/Italy_orf_abundance_model1_nbinom_timepoint_specific_results_p.adj.csv", row.names = FALSE)
}
# 0 significant


```


# Create heatmap for significant Abundance results
```{r}

create_orf_heatmap <- function(timepoint_data, cohort_name, output_dir = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/"){
  
  print("test1")
  sig_data <- timepoint_data %>%
    filter(!is.na(p.adj) & p.adj < 0.05)

    timepoint_order <- c("t0-over42", "t0-36", "t0-30", "t0-24", "t0-18", "t0-12", "t0-6", "t0")

    
  print("test2")  
    heatmap_data <- sig_data %>%
                    select(gene, timepoint, estimate) %>%
                    mutate(timepoint = factor(timepoint, levels = timepoint_order)) %>%
                    filter(timepoint %in% timepoint_order) %>%
                    tidyr::pivot_wider(names_from = timepoint, values_from = estimate, values_fill = 0) %>%
                    column_to_rownames("gene")

  print("test3") 
     mat <- as.matrix(heatmap_data)
     
     print("test4") 

     m <- max(abs(mat), na.rm = TRUE)
     
     print("test5") 
      if (!is.finite(m) || m == 0) m <- 1  # safe fallback if all zeros
     
     print("test6") 
      cols   <- colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(101)
      print("test7") 
      breaks <- seq(-m, m, length.out = length(cols) + 1)

      print("test8") 
      p <- pheatmap(
        mat,
        color        = cols,
        breaks       = breaks,      # <- key change: diverging scale centered at 0
        scale        = "none",
        cluster_cols = FALSE,       # keep chronological order
        cluster_rows = TRUE,
        fontsize     = 8,
        main         = paste(cohort_name, "Cohort - ORF PA Analysis\nLog Odds Ratios (CONTROL vs CELIAC)"),
        na_col       = "grey90"
      )

      print("test9") 

      return(p)

}


total_heatmap <- create_orf_heatmap(total_abundance_timepoints, "total")
ggsave(total_heatmap,file = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_Abundance_longitudinal_heatmap_selfmade.pdf",dpi = 600,width = 6,height = 7)


US_heatmap <- create_orf_heatmap(US_abundance_timepoints, "US")
ggsave(US_heatmap,file = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_Abundance_longitudinal_heatmap_selfmade.pdf",dpi = 600,width = 6,height = 7)

Italy_heatmap <- create_orf_heatmap(Italy_pa_timepoints, "Italy")
# no significant orf

```



######################################################### Compositional ##################################################

# Compositional analysis selfmade - Alpha

# total - per time point contrast
```{r}

# ── Packages ───────────────────────────────────────────────────────────────────
library(tidyverse)
library(phyloseq)
library(glmmTMB)
library(lme4)
library(emmeans)
library(broom.mixed)

# ── 0) Inputs: counts + metadata ───────────────────────────────────────────────
# counts: rows=orfs (features), cols=samples (counts)
count_mat <- total.orf.abundance.clean_0.75_0.03 %>% as.data.frame() %>% as.matrix()
meta_df   <- total.orf.metadata.clean_0.75_0.03 %>% as.data.frame()

stopifnot(!is.null(rownames(count_mat)), !is.null(colnames(count_mat)))
stopifnot(!is.null(rownames(meta_df)))

# align by common samples
common_samples <- intersect(colnames(count_mat), rownames(meta_df))
count_mat <- count_mat[, common_samples, drop = FALSE]
meta_df   <- meta_df[common_samples, , drop = FALSE]

# ── 1) Build phyloseq (no tree needed) ────────────────────────────────────────
OTU <- otu_table(count_mat, taxa_are_rows = TRUE)
SAM <- sample_data(meta_df)
ps  <- phyloseq(OTU, SAM)

# exposure / library size (replace with effective lib size if you have it)
lib_size <- sample_sums(ps)
sample_data(ps)$eff_libsize <- as.numeric(lib_size)
sample_data(ps)$offset_log  <- log(pmax(sample_data(ps)$eff_libsize, 1))

# ── 2) Alpha diversity per sample ─────────────────────────────────────────────
otu_samp_taxa <- t(otu_table(ps))                    # samples x taxa
observed <- vegan::specnumber(otu_samp_taxa)
shannon  <- vegan::diversity(otu_samp_taxa, index = "shannon")

alpha_tbl <- tibble(
  sample_id = names(observed),
  Observed  = as.numeric(observed),
  Shannon   = as.numeric(shannon)
)


alpha_all <- alpha_tbl %>%
  left_join(meta_df %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(
    eff_libsize = as.numeric(lib_size[sample_id]),
    offset_log  = log(pmax(eff_libsize, 1)),
    # factors to match your other models
    Dx.Status = forcats::fct_relevel(as.factor(Dx.Status), "CONTROL"),
    onset_timeline_combined = as.factor(onset_timeline_combined),
    patientID = as.factor(patientID),
    Country = forcats::fct_relevel(as.factor(Country), "ITALY"),
    Sex = as.factor(Sex),
    HLA.Category = as.factor(HLA.Category),
    feeding_first_year = as.factor(feeding_first_year),
    Delivery.Mode = as.factor(Delivery.Mode),
    Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.)
  ) %>%
  drop_na(Dx.Status, onset_timeline_combined, patientID)


# ── 3) GLOBAL models (same structure as PA/abundance) ─────────────────────────
# Observed richness ~ NB with offset + random intercept
fit_obs <- glmmTMB(
  Observed ~ Dx.Status * onset_timeline_combined +
    Country + Sex + Age.at.Gluten.Introduction..months. +
    HLA.Category + feeding_first_year + Delivery.Mode +
    (1 | patientID) + offset(offset_log),
  family = nbinom2,
  data = alpha_all
)



# Shannon ~ Gaussian LMM with random intercept
fit_shan <- glmmTMB(
  Shannon ~ Dx.Status * onset_timeline_combined +
    Country + Sex + Age.at.Gluten.Introduction..months. +
    HLA.Category + feeding_first_year + Delivery.Mode +
    (1 | patientID),
  family = gaussian(),   # <- key change
  data = alpha_all
)

library(emmeans)
library(tidyverse)

# Observed richness: hold offset at 0 so contrasts are rate ratios (IRR)
rg_obs <- ref_grid(fit_obs, at = list(offset_log = 0))
emm_obs <- emmeans(rg_obs, ~ Dx.Status | onset_timeline_combined)
ctr_obs <- contrast(emm_obs, method = "revpairwise", by = "onset_timeline_combined", adjust = "none")

# Shannon diversity: Gaussian GLMM, direct emmeans
emm_shan <- emmeans(fit_shan, ~ Dx.Status | onset_timeline_combined)
ctr_shan <- contrast(emm_shan, method = "revpairwise", by = "onset_timeline_combined", adjust = "none")



# Observed: log(IRR), plus exponentiated IRR
sum_obs <- as_tibble(summary(ctr_obs, infer = c(TRUE, TRUE))) %>%
  left_join(as_tibble(confint(ctr_obs)) %>%
              select(contrast, onset_timeline_combined),
            by = c("contrast","onset_timeline_combined")) %>%
  mutate(
    metric = "Observed",
    IRR      = exp(estimate),
    IRR.low  = exp(asymp.LCL),
    IRR.high = exp(asymp.UCL)
  )


# Shannon: difference in mean Shannon between groups
sum_shan <- as_tibble(summary(ctr_shan, infer = c(TRUE, TRUE))) %>%
  left_join(as_tibble(confint(ctr_shan)) %>%
              select(contrast, onset_timeline_combined),
            by = c("contrast","onset_timeline_combined")) %>%
  mutate(metric = "Shannon")


# Combine results and adjust p-values
alpha_contrasts <- bind_rows(sum_obs, sum_shan) %>%
  filter(grepl("CELIAC", contrast) & grepl("CONTROL", contrast)) %>%
  dplyr::rename(timepoint = onset_timeline_combined) %>%
  group_by(metric) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


# make plot for Observed
total_observed_contrast <- alpha_contrasts %>%
  filter(metric == "Observed") %>%
  mutate(
    timepoint = factor(timepoint, levels = unique(timepoint)),
    sig = !is.na(p.adj) & p.adj < 0.05
  ) %>%
  ggplot(aes(y = timepoint, x = IRR)) +
  geom_errorbarh(aes(xmin = IRR.low, xmax = IRR.high), height = 0.25) +
  geom_point(aes(color = sig), size = 2.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "IRR (CONTROL vs CELIAC)",
    y = "Timepoint",
    title = "Total Observed richness: per-timepoint IRR with 95% CI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(total_observed_contrast,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_observed_contrast.pdf",dpi = 600,width = 6,height=6)



# make plot for Observed
total_shannon_contrast <- alpha_contrasts %>%
  filter(metric == "Shannon") %>%
  mutate(
    timepoint = factor(timepoint, levels = unique(timepoint)),
    sig = !is.na(p.adj) & p.adj < 0.05
  ) %>%
  ggplot(aes(y = timepoint, x = estimate)) +
  geom_errorbarh(aes(xmin = lower.CL, xmax = upper.CL), height = 0.25) +
  geom_point(aes(color = sig), size = 2.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Shannon difference (CONTROL vs CELIAC)",
    y = "Timepoint",
    title = "Total Shannon diversity: per-timepoint difference with 95% CI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(total_shannon_contrast,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_shannon_contrast.pdf",dpi = 600,width = 6,height=6)


```

# total - all terms
```{r}

library(broom.mixed)
library(dplyr)
library(tibble)
library(stringr)
library(forcats)
library(ggplot2)

# --- 1) Tidy all fixed effects (exclude intercept & RE summaries) -------------
tidy_all_shan <- tidy(fit_shan, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Shannon")

tidy_all_obs <- tidy(fit_obs, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Observed (log scale)",
         IRR = exp(estimate),
         IRR.low  = exp(conf.low),
         IRR.high = exp(conf.high))

# --- 2) Combine and compute BH FDR within each model --------------------------
all_terms <- bind_rows(tidy_all_shan, tidy_all_obs) %>%
  group_by(model) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  mutate(sig = !is.na(p.adj) & p.adj < 0.05)


# --- 3) Make human-friendly term labels ---------------------------------------
pretty_label <- function(x) {
  x %>%
    # Interaction first (keep it readable)
    str_replace("^Dx\\.StatusCELIAC:onset_timeline_combined", "Dx: CELIAC × TP: ") %>%
    # Main effects for the interaction components
    str_replace("^Dx\\.StatusCELIAC$", "Dx: CELIAC (vs CONTROL)") %>%
    str_replace("^onset_timeline_combined", "TP: ") %>%
    # Other categorical covariates (adjust names if your levels differ)
    str_replace("^SexMale$", "Sex: Male (vs Female)") %>%
    str_replace("^HLA\\.CategoryHigh Risk$", "HLA: High Risk (vs Ref)") %>%
    str_replace("^HLA\\.CategoryLow/No Risk$", "HLA: Low/No Risk (vs Ref)") %>%
    str_replace("^feeding_first_yearFormula$", "Feeding: Formula (vs Breastmilk)") %>%
    str_replace("^feeding_first_yearBreastmilk_and_formula$", "Feeding: Mixed (vs Breastmilk)") %>%
    str_replace("^Delivery\\.ModeC-Section$", "Delivery: C-Section (vs Vaginal)") %>%
    # Continuous covariate
    str_replace("^Age\\.at\\.Gluten\\.Introduction\\.\\.months\\.$", "Age at gluten intro (months)")
}


all_terms <- all_terms %>%
  mutate(term_label = pretty_label(term))

# Optional: order terms in logical groups (interactions, then mains, then others)
all_terms <- all_terms %>%
  mutate(group =
           case_when(
             str_detect(term, "^Dx\\.Status.*:onset_timeline_combined") ~ "Interaction: Dx × TP",
             str_detect(term, "^onset_timeline_combined")               ~ "Main: Timepoint (vs baseline)",
             str_detect(term, "^Dx\\.Status")                            ~ "Main: Dx",
             TRUE                                                        ~ "Other covariates"
           ))



# 1) Define the desired panel order for 'group'
group_levels <- c(
  "Interaction: Dx × TP",
  "Main: Timepoint (vs baseline)",
  "Main: Dx",
  "Other covariates"
)

# 2) Build stable ordering keys first, then arrange
all_terms <- all_terms %>%
  # make sure term_label is plain character for ordering
  mutate(
    group = factor(group, levels = group_levels),
    term_label = as.character(term_label),
    .by = NULL
  ) %>%
  arrange(model, group, term_label) %>%
  group_by(model) %>%
  # now freeze the y-axis order within each model
  mutate(term_label = factor(term_label, levels = unique(term_label))) %>%
  ungroup()



# --- 4) Forest plots ----------------------------------------------------------
## A) Shannon (difference; 0 = no effect)
total_orf_shannon_allTerms <- all_terms %>%
  filter(model == "Shannon") %>%
  ggplot(aes(y = fct_rev(term_label), x = estimate)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.22) +
  geom_point(aes(color = p.adj < 0.05), size = 2.7) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Total Shannon diversity — all fixed terms",
    x = "Estimate (difference in Shannon)",
    y = NULL,
    caption = "Red: BH FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.y = element_text(face = "bold"))

ggsave(total_orf_shannon_allTerms,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_shannon_allTerms.pdf",dpi = 600,width = 6,height=12)


## B) Observed richness (log scale; 0 = no effect)
total_orf_observed_allTerms <- all_terms %>%
  filter(model == "Observed (log scale)") %>%
  ggplot(aes(y = fct_rev(term_label), x = estimate)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.22) +
  geom_point(aes(color = p.adj < 0.05), size = 2.7) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Total Observed richness — all fixed terms",
    x = "Estimate (log rate ratio)",
    y = NULL,
    caption = "Red: BH FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.y = element_text(face = "bold"))


ggsave(total_orf_observed_allTerms,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_observed_allTerms.pdf",dpi = 600,width = 6,height=12)


```


# US - per time point contrast
```{r}


# ── 0) Inputs: counts + metadata ───────────────────────────────────────────────
# counts: rows=orfs (features), cols=samples (counts)
count_mat <- US.orf.abundance.clean_0.75_0.03 %>% as.data.frame() %>% as.matrix()
meta_df   <- US.orf.metadata.clean_0.75_0.03 %>% as.data.frame()

stopifnot(!is.null(rownames(count_mat)), !is.null(colnames(count_mat)))
stopifnot(!is.null(rownames(meta_df)))

# align by common samples
common_samples <- intersect(colnames(count_mat), rownames(meta_df))
count_mat <- count_mat[, common_samples, drop = FALSE]
meta_df   <- meta_df[common_samples, , drop = FALSE]

# ── 1) Build phyloseq (no tree needed) ────────────────────────────────────────
OTU <- otu_table(count_mat, taxa_are_rows = TRUE)
SAM <- sample_data(meta_df)
ps  <- phyloseq(OTU, SAM)

# exposure / library size (replace with effective lib size if you have it)
lib_size <- sample_sums(ps)
sample_data(ps)$eff_libsize <- as.numeric(lib_size)
sample_data(ps)$offset_log  <- log(pmax(sample_data(ps)$eff_libsize, 1))

# ── 2) Alpha diversity per sample ─────────────────────────────────────────────
otu_samp_taxa <- t(otu_table(ps))                    # samples x taxa
observed <- vegan::specnumber(otu_samp_taxa)
shannon  <- vegan::diversity(otu_samp_taxa, index = "shannon")

alpha_tbl <- tibble(
  sample_id = names(observed),
  Observed  = as.numeric(observed),
  Shannon   = as.numeric(shannon)
)


alpha_all <- alpha_tbl %>%
  left_join(meta_df %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(
    eff_libsize = as.numeric(lib_size[sample_id]),
    offset_log  = log(pmax(eff_libsize, 1)),
    # factors to match your other models
    Dx.Status = forcats::fct_relevel(as.factor(Dx.Status), "CONTROL"),
    onset_timeline_combined = as.factor(onset_timeline_combined),
    patientID = as.factor(patientID),
    Sex = as.factor(Sex),
    HLA.Category = as.factor(HLA.Category),
    feeding_first_year = as.factor(feeding_first_year),
    Delivery.Mode = as.factor(Delivery.Mode),
    Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.)
  ) %>%
  drop_na(Dx.Status, onset_timeline_combined, patientID)


# ── 3) GLOBAL models (same structure as PA/abundance) ─────────────────────────
# Observed richness ~ NB with offset + random intercept
fit_obs <- glmmTMB(
  Observed ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
    HLA.Category + feeding_first_year + Delivery.Mode +
    (1 | patientID) + offset(offset_log),
  family = nbinom2,
  data = alpha_all
)

# Shannon ~ Gaussian LMM with random intercept
fit_shan <- glmmTMB(
  Shannon ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
    HLA.Category + feeding_first_year + Delivery.Mode +
    (1 | patientID),
  family = gaussian(),   # <- key change
  data = alpha_all
)

library(emmeans)
library(tidyverse)

# Observed richness: hold offset at 0 so contrasts are rate ratios (IRR)
rg_obs <- ref_grid(fit_obs, at = list(offset_log = 0))
emm_obs <- emmeans(rg_obs, ~ Dx.Status | onset_timeline_combined)
ctr_obs <- contrast(emm_obs, method = "revpairwise", by = "onset_timeline_combined", adjust = "none")

# Shannon diversity: Gaussian GLMM, direct emmeans
emm_shan <- emmeans(fit_shan, ~ Dx.Status | onset_timeline_combined)
ctr_shan <- contrast(emm_shan, method = "revpairwise", by = "onset_timeline_combined", adjust = "none")



# Observed: log(IRR), plus exponentiated IRR
sum_obs <- as_tibble(summary(ctr_obs, infer = c(TRUE, TRUE))) %>%
  left_join(as_tibble(confint(ctr_obs)) %>%
              select(contrast, onset_timeline_combined),
            by = c("contrast","onset_timeline_combined")) %>%
  mutate(
    metric = "Observed",
    IRR      = exp(estimate),
    IRR.low  = exp(asymp.LCL),
    IRR.high = exp(asymp.UCL)
  )


# Shannon: difference in mean Shannon between groups
sum_shan <- as_tibble(summary(ctr_shan, infer = c(TRUE, TRUE))) %>%
  left_join(as_tibble(confint(ctr_shan)) %>%
              select(contrast, onset_timeline_combined),
            by = c("contrast","onset_timeline_combined")) %>%
  mutate(metric = "Shannon")


# Combine results and adjust p-values
alpha_contrasts <- bind_rows(sum_obs, sum_shan) %>%
  filter(grepl("CELIAC", contrast) & grepl("CONTROL", contrast)) %>%
  dplyr::rename(timepoint = onset_timeline_combined) %>%
  group_by(metric) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


# make plot for Observed
US_observed_contrast <- alpha_contrasts %>%
  filter(metric == "Observed") %>%
  mutate(
    timepoint = factor(timepoint, levels = unique(timepoint)),
    sig = !is.na(p.adj) & p.adj < 0.05
  ) %>%
  ggplot(aes(y = timepoint, x = IRR)) +
  geom_errorbarh(aes(xmin = IRR.low, xmax = IRR.high), height = 0.25) +
  geom_point(aes(color = sig), size = 2.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "IRR (CONTROL vs CELIAC)",
    y = "Timepoint",
    title = "US Observed richness: per-timepoint IRR with 95% CI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(US_observed_contrast,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_observed_contrast.pdf",dpi = 600,width = 6,height=6)



# make plot for Observed
US_shannon_contrast <- alpha_contrasts %>%
  filter(metric == "Shannon") %>%
  mutate(
    timepoint = factor(timepoint, levels = unique(timepoint)),
    sig = !is.na(p.adj) & p.adj < 0.05
  ) %>%
  ggplot(aes(y = timepoint, x = estimate)) +
  geom_errorbarh(aes(xmin = lower.CL, xmax = upper.CL), height = 0.25) +
  geom_point(aes(color = sig), size = 2.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Shannon difference (CONTROL vs CELIAC)",
    y = "Timepoint",
    title = "US Shannon diversity: per-timepoint difference with 95% CI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(US_shannon_contrast,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_shannon_contrast.pdf",dpi = 600,width = 6,height=6)



```

# US - all terms
```{r}

# --- 1) Tidy all fixed effects (exclude intercept & RE summaries) -------------
tidy_all_shan <- tidy(fit_shan, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Shannon")

tidy_all_obs <- tidy(fit_obs, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Observed (log scale)",
         IRR = exp(estimate),
         IRR.low  = exp(conf.low),
         IRR.high = exp(conf.high))

# --- 2) Combine and compute BH FDR within each model --------------------------
all_terms <- bind_rows(tidy_all_shan, tidy_all_obs) %>%
  group_by(model) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  mutate(sig = !is.na(p.adj) & p.adj < 0.05)


# --- 3) Make human-friendly term labels ---------------------------------------
pretty_label <- function(x) {
  x %>%
    # Interaction first (keep it readable)
    str_replace("^Dx\\.StatusCELIAC:onset_timeline_combined", "Dx: CELIAC × TP: ") %>%
    # Main effects for the interaction components
    str_replace("^Dx\\.StatusCELIAC$", "Dx: CELIAC (vs CONTROL)") %>%
    str_replace("^onset_timeline_combined", "TP: ") %>%
    # Other categorical covariates (adjust names if your levels differ)
    str_replace("^SexMale$", "Sex: Male (vs Female)") %>%
    str_replace("^HLA\\.CategoryHigh Risk$", "HLA: High Risk (vs Ref)") %>%
    str_replace("^HLA\\.CategoryLow/No Risk$", "HLA: Low/No Risk (vs Ref)") %>%
    str_replace("^feeding_first_yearFormula$", "Feeding: Formula (vs Breastmilk)") %>%
    str_replace("^feeding_first_yearBreastmilk_and_formula$", "Feeding: Mixed (vs Breastmilk)") %>%
    str_replace("^Delivery\\.ModeC-Section$", "Delivery: C-Section (vs Vaginal)") %>%
    # Continuous covariate
    str_replace("^Age\\.at\\.Gluten\\.Introduction\\.\\.months\\.$", "Age at gluten intro (months)")
}


all_terms <- all_terms %>%
  mutate(term_label = pretty_label(term))

# Optional: order terms in logical groups (interactions, then mains, then others)
all_terms <- all_terms %>%
  mutate(group =
           case_when(
             str_detect(term, "^Dx\\.Status.*:onset_timeline_combined") ~ "Interaction: Dx × TP",
             str_detect(term, "^onset_timeline_combined")               ~ "Main: Timepoint (vs baseline)",
             str_detect(term, "^Dx\\.Status")                            ~ "Main: Dx",
             TRUE                                                        ~ "Other covariates"
           ))



# 1) Define the desired panel order for 'group'
group_levels <- c(
  "Interaction: Dx × TP",
  "Main: Timepoint (vs baseline)",
  "Main: Dx",
  "Other covariates"
)

# 2) Build stable ordering keys first, then arrange
all_terms <- all_terms %>%
  # make sure term_label is plain character for ordering
  mutate(
    group = factor(group, levels = group_levels),
    term_label = as.character(term_label),
    .by = NULL
  ) %>%
  arrange(model, group, term_label) %>%
  group_by(model) %>%
  # now freeze the y-axis order within each model
  mutate(term_label = factor(term_label, levels = unique(term_label))) %>%
  ungroup()



# --- 4) Forest plots ----------------------------------------------------------
## A) Shannon (difference; 0 = no effect)
US_orf_shannon_allTerms <- all_terms %>%
  filter(model == "Shannon") %>%
  ggplot(aes(y = fct_rev(term_label), x = estimate)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.22) +
  geom_point(aes(color = p.adj < 0.05), size = 2.7) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "US Shannon diversity — all fixed terms",
    x = "Estimate (difference in Shannon)",
    y = NULL,
    caption = "Red: BH FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.y = element_text(face = "bold"))

ggsave(US_orf_shannon_allTerms,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_shannon_allTerms.pdf",dpi = 600,width = 6,height=12)


## B) Observed richness (log scale; 0 = no effect)
US_orf_observed_allTerms <- all_terms %>%
  filter(model == "Observed (log scale)") %>%
  ggplot(aes(y = fct_rev(term_label), x = estimate)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.22) +
  geom_point(aes(color = p.adj < 0.05), size = 2.7) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "US Observed richness — all fixed terms",
    x = "Estimate (log rate ratio)",
    y = NULL,
    caption = "Red: BH FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.y = element_text(face = "bold"))


ggsave(US_orf_observed_allTerms,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/US_orf_observed_allTerms.pdf",dpi = 600,width = 6,height=12)


```


# Italy - per time point contrast
```{r}


# ── 0) Inputs: counts + metadata ───────────────────────────────────────────────
# counts: rows=orfs (features), cols=samples (counts)
count_mat <- Italy.orf.abundance.clean_0.75_0.03 %>% as.data.frame() %>% as.matrix()
meta_df   <- Italy.orf.metadata.clean_0.75_0.03 %>% as.data.frame()

stopifnot(!is.null(rownames(count_mat)), !is.null(colnames(count_mat)))
stopifnot(!is.null(rownames(meta_df)))

# align by common samples
common_samples <- intersect(colnames(count_mat), rownames(meta_df))
count_mat <- count_mat[, common_samples, drop = FALSE]
meta_df   <- meta_df[common_samples, , drop = FALSE]

# ── 1) Build phyloseq (no tree needed) ────────────────────────────────────────
OTU <- otu_table(count_mat, taxa_are_rows = TRUE)
SAM <- sample_data(meta_df)
ps  <- phyloseq(OTU, SAM)

# exposure / library size (replace with effective lib size if you have it)
lib_size <- sample_sums(ps)
sample_data(ps)$eff_libsize <- as.numeric(lib_size)
sample_data(ps)$offset_log  <- log(pmax(sample_data(ps)$eff_libsize, 1))

# ── 2) Alpha diversity per sample ─────────────────────────────────────────────
otu_samp_taxa <- t(otu_table(ps))                    # samples x taxa
observed <- vegan::specnumber(otu_samp_taxa)
shannon  <- vegan::diversity(otu_samp_taxa, index = "shannon")

alpha_tbl <- tibble(
  sample_id = names(observed),
  Observed  = as.numeric(observed),
  Shannon   = as.numeric(shannon)
)


alpha_all <- alpha_tbl %>%
  left_join(meta_df %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(
    eff_libsize = as.numeric(lib_size[sample_id]),
    offset_log  = log(pmax(eff_libsize, 1)),
    # factors to match your other models
    Dx.Status = forcats::fct_relevel(as.factor(Dx.Status), "CONTROL"),
    onset_timeline_combined = as.factor(onset_timeline_combined),
    patientID = as.factor(patientID),
    Sex = as.factor(Sex),
    HLA.Category = as.factor(HLA.Category),
    feeding_first_year = as.factor(feeding_first_year),
    Delivery.Mode = as.factor(Delivery.Mode),
    Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.)
  ) %>%
  drop_na(Dx.Status, onset_timeline_combined, patientID)


# ── 3) GLOBAL models (same structure as PA/abundance) ─────────────────────────
# Observed richness ~ NB with offset + random intercept
fit_obs <- glmmTMB(
  Observed ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
    HLA.Category + feeding_first_year + Delivery.Mode +
    (1 | patientID) + offset(offset_log),
  family = nbinom2,
  data = alpha_all
)

# Shannon ~ Gaussian LMM with random intercept
fit_shan <- glmmTMB(
  Shannon ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. +
    HLA.Category + feeding_first_year + Delivery.Mode +
    (1 | patientID),
  family = gaussian(),   # <- key change
  data = alpha_all
)



# Observed richness: hold offset at 0 so contrasts are rate ratios (IRR)
rg_obs <- ref_grid(fit_obs, at = list(offset_log = 0))
emm_obs <- emmeans(rg_obs, ~ Dx.Status | onset_timeline_combined)
ctr_obs <- contrast(emm_obs, method = "revpairwise", by = "onset_timeline_combined", adjust = "none")

# Shannon diversity: Gaussian GLMM, direct emmeans
emm_shan <- emmeans(fit_shan, ~ Dx.Status | onset_timeline_combined)
ctr_shan <- contrast(emm_shan, method = "revpairwise", by = "onset_timeline_combined", adjust = "none")



# Observed: log(IRR), plus exponentiated IRR
sum_obs <- as_tibble(summary(ctr_obs, infer = c(TRUE, TRUE))) %>%
  left_join(as_tibble(confint(ctr_obs)) %>%
              select(contrast, onset_timeline_combined),
            by = c("contrast","onset_timeline_combined")) %>%
  mutate(
    metric = "Observed",
    IRR      = exp(estimate),
    IRR.low  = exp(asymp.LCL),
    IRR.high = exp(asymp.UCL)
  )


# Shannon: difference in mean Shannon between groups
sum_shan <- as_tibble(summary(ctr_shan, infer = c(TRUE, TRUE))) %>%
  left_join(as_tibble(confint(ctr_shan)) %>%
              select(contrast, onset_timeline_combined),
            by = c("contrast","onset_timeline_combined")) %>%
  mutate(metric = "Shannon")


# Combine results and adjust p-values
alpha_contrasts <- bind_rows(sum_obs, sum_shan) %>%
  filter(grepl("CELIAC", contrast) & grepl("CONTROL", contrast)) %>%
  dplyr::rename(timepoint = onset_timeline_combined) %>%
  group_by(metric) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


# make plot for Observed
Italy_observed_contrast <- alpha_contrasts %>%
  filter(metric == "Observed") %>%
  mutate(
    timepoint = factor(timepoint, levels = unique(timepoint)),
    sig = !is.na(p.adj) & p.adj < 0.05
  ) %>%
  ggplot(aes(y = timepoint, x = IRR)) +
  geom_errorbarh(aes(xmin = IRR.low, xmax = IRR.high), height = 0.25) +
  geom_point(aes(color = sig), size = 2.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "IRR (CONTROL vs CELIAC)",
    y = "Timepoint",
    title = "Italy Observed richness: per-timepoint IRR with 95% CI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(Italy_observed_contrast,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/Italy_orf_observed_contrast.pdf",dpi = 600,width = 6,height=6)



# make plot for Observed
Italy_shannon_contrast <- alpha_contrasts %>%
  filter(metric == "Shannon") %>%
  mutate(
    timepoint = factor(timepoint, levels = unique(timepoint)),
    sig = !is.na(p.adj) & p.adj < 0.05
  ) %>%
  ggplot(aes(y = timepoint, x = estimate)) +
  geom_errorbarh(aes(xmin = lower.CL, xmax = upper.CL), height = 0.25) +
  geom_point(aes(color = sig), size = 2.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Shannon difference (CONTROL vs CELIAC)",
    y = "Timepoint",
    title = "Italy Shannon diversity: per-timepoint difference with 95% CI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(Italy_shannon_contrast,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/Italy_orf_shannon_contrast.pdf",dpi = 600,width = 6,height=6)



```


# Italy - all terms
```{r}

# --- 1) Tidy all fixed effects (exclude intercept & RE summaries) -------------
tidy_all_shan <- tidy(fit_shan, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Shannon")

tidy_all_obs <- tidy(fit_obs, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Observed (log scale)",
         IRR = exp(estimate),
         IRR.low  = exp(conf.low),
         IRR.high = exp(conf.high))

# --- 2) Combine and compute BH FDR within each model --------------------------
all_terms <- bind_rows(tidy_all_shan, tidy_all_obs) %>%
  group_by(model) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  mutate(sig = !is.na(p.adj) & p.adj < 0.05)


# --- 3) Make human-friendly term labels ---------------------------------------
pretty_label <- function(x) {
  x %>%
    # Interaction first (keep it readable)
    str_replace("^Dx\\.StatusCELIAC:onset_timeline_combined", "Dx: CELIAC × TP: ") %>%
    # Main effects for the interaction components
    str_replace("^Dx\\.StatusCELIAC$", "Dx: CELIAC (vs CONTROL)") %>%
    str_replace("^onset_timeline_combined", "TP: ") %>%
    # Other categorical covariates (adjust names if your levels differ)
    str_replace("^SexMale$", "Sex: Male (vs Female)") %>%
    str_replace("^HLA\\.CategoryHigh Risk$", "HLA: High Risk (vs Ref)") %>%
    str_replace("^HLA\\.CategoryLow/No Risk$", "HLA: Low/No Risk (vs Ref)") %>%
    str_replace("^feeding_first_yearFormula$", "Feeding: Formula (vs Breastmilk)") %>%
    str_replace("^feeding_first_yearBreastmilk_and_formula$", "Feeding: Mixed (vs Breastmilk)") %>%
    str_replace("^Delivery\\.ModeC-Section$", "Delivery: C-Section (vs Vaginal)") %>%
    # Continuous covariate
    str_replace("^Age\\.at\\.Gluten\\.Introduction\\.\\.months\\.$", "Age at gluten intro (months)")
}


all_terms <- all_terms %>%
  mutate(term_label = pretty_label(term))

# Optional: order terms in logical groups (interactions, then mains, then others)
all_terms <- all_terms %>%
  mutate(group =
           case_when(
             str_detect(term, "^Dx\\.Status.*:onset_timeline_combined") ~ "Interaction: Dx × TP",
             str_detect(term, "^onset_timeline_combined")               ~ "Main: Timepoint (vs baseline)",
             str_detect(term, "^Dx\\.Status")                            ~ "Main: Dx",
             TRUE                                                        ~ "Other covariates"
           ))



# 1) Define the desired panel order for 'group'
group_levels <- c(
  "Interaction: Dx × TP",
  "Main: Timepoint (vs baseline)",
  "Main: Dx",
  "Other covariates"
)

# 2) Build stable ordering keys first, then arrange
all_terms <- all_terms %>%
  # make sure term_label is plain character for ordering
  mutate(
    group = factor(group, levels = group_levels),
    term_label = as.character(term_label),
    .by = NULL
  ) %>%
  arrange(model, group, term_label) %>%
  group_by(model) %>%
  # now freeze the y-axis order within each model
  mutate(term_label = factor(term_label, levels = unique(term_label))) %>%
  ungroup()



# --- 4) Forest plots ----------------------------------------------------------
## A) Shannon (difference; 0 = no effect)
Italy_orf_shannon_allTerms <- all_terms %>%
  filter(model == "Shannon") %>%
  ggplot(aes(y = fct_rev(term_label), x = estimate)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.22) +
  geom_point(aes(color = p.adj < 0.05), size = 2.7) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Italy Shannon diversity — all fixed terms",
    x = "Estimate (difference in Shannon)",
    y = NULL,
    caption = "Red: BH FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.y = element_text(face = "bold"))

ggsave(Italy_orf_shannon_allTerms,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/Italy_orf_shannon_allTerms.pdf",dpi = 600,width = 6,height=12)


## B) Observed richness (log scale; 0 = no effect)
Itayl_orf_observed_allTerms <- all_terms %>%
  filter(model == "Observed (log scale)") %>%
  ggplot(aes(y = fct_rev(term_label), x = estimate)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.22) +
  geom_point(aes(color = p.adj < 0.05), size = 2.7) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "black")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Italy Observed richness — all fixed terms",
    x = "Estimate (log rate ratio)",
    y = NULL,
    caption = "Red: BH FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.y = element_text(face = "bold"))


ggsave(Itayl_orf_observed_allTerms,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/Itayl_orf_observed_allTerms.pdf",dpi = 600,width = 6,height=12)


```



# Compositional analysis selfmade - Beta

# total - per timepoint contrast
```{r}


# Packages
library(tidyverse)
library(vegan)
library(MDMR)

# ---- Inputs ----
counts <- total.orf.abundance.clean_0.75_0.03 %>% as.data.frame() %>% as.matrix()
meta   <- total.orf.metadata.clean_0.75_0.03 %>% as.data.frame() %>% tibble::rownames_to_column("sample_id")

# Align samples
common <- intersect(colnames(counts), meta$sample_id)
counts <- counts[, common, drop = FALSE]
meta   <- meta  %>% filter(sample_id %in% common)

# Bray–Curtis on relative abundances
X        <- t(counts)  # samples x ORFs
row_sums <- rowSums(X)
X_rel    <- sweep(X, 1, pmax(row_sums, 1e-12), "/")
D_bray   <- vegdist(X_rel, method = "bray")

# Reorder metadata to match distance matrix
ord_ids <- rownames(as.matrix(D_bray))
meta    <- meta %>% slice(match(ord_ids, sample_id))

# Coerce factors (important)
meta <- meta %>%
  mutate(
    Dx.Status = forcats::fct_relevel(as.factor(Dx.Status), "CONTROL"),
    Country = forcats::fct_relevel(as.factor(Country), "ITALY"),
    onset_timeline_combined = as.factor(onset_timeline_combined),
    patientID = as.factor(patientID),
    Sex = as.factor(Sex),
    HLA.Category = as.factor(HLA.Category),
    feeding_first_year = as.factor(feeding_first_year),
    Delivery.Mode = as.factor(Delivery.Mode),
    Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.)
  )

# ---- mixed MDMR: fixed effects + random intercept for patientID ----
fmla_mixed <- ~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. +
               HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID)

# Tip for Bray (non-Euclidean): use.ssd ~ 0.99 can speed up
fit_mmdmr <- mixed.mdmr(
  fmla = fmla_mixed,
  data = meta,
  D    = D_bray,
  use.ssd = 0.99,   # set to 1.00 if you prefer full SSD (slower)
  ncores = 1        # >1 uses forking; keep 1 on Windows
)

# Results
res_tab <- summary(fit_mmdmr)           # Statistic, Numer DF, p-value per term
res_tab <- data.frame(res_tab) %>%
  rownames_to_column("term") %>%
  mutate(p.adj = p.adjust(`p.value`, method = "BH"))



ord_ids <- rownames(as.matrix(D_bray))
stopifnot(identical(meta$sample_id, ord_ids))

# Helper that runs MDMR for ONE timepoint, adjusting for covariates
run_mdmr_tp <- function(tp, meta, D_full, ord_ids, use_perm_p = FALSE, nperm = 4999, seed = 1L) {
  print(tp)
  df <- meta %>% filter(onset_timeline_combined == tp) %>% droplevels()
  if (nrow(df) < 3 || n_distinct(df$Dx.Status) < 2) {
    return(tibble(timepoint = tp, term = "Dx.Status", Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }
  idx <- match(df$sample_id, ord_ids)
  if (any(is.na(idx))) {
    return(tibble(timepoint = tp, term = "Dx.Status", Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }
  D_tp <- as.dist(as.matrix(D_full)[idx, idx, drop = FALSE])


  # Quick degeneracy check (all pairwise distances == 0)
  M <- as.matrix(D_tp); U <- upper.tri(M, diag = FALSE)
  if (!any(U) || all(M[U] == 0)) {
    print("test")
    return(tibble(timepoint = tp, term = "Dx.Status",
                  Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }

  # Build X as a DATA FRAME with factors kept as factors
  X_df <- df %>%
    transmute(
      Dx.Status = factor(Dx.Status),
      Country = factor(Country),
      Sex = factor(Sex),
      Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.),
      HLA.Category = factor(HLA.Category),
      feeding_first_year = factor(feeding_first_year),
      Delivery.Mode = factor(Delivery.Mode)
    )

  # mdmr() expects X= (matrix or data.frame) and D=/G=
  # Use analytic p-values by default; or set use_perm_p = TRUE for permutation p's
  set.seed(seed)
  fit <- tryCatch(
    MDMR::mdmr(
      X = X_df,
      D = D_tp,
      perm.p = use_perm_p,
      nperm = if (use_perm_p) nperm else 500  # ignored if perm.p=FALSE
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    print("fit is still null")
    return(tibble(timepoint = tp, term = "Dx.Status",
                  Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }

  
  sm <- data.frame(summary(fit)) %>%
        rownames_to_column("term") %>%
    filter(term == "Dx.Status1") %>%
    mutate(timepoint = tp) %>%
    select(timepoint,Statistic,Numer.DF,Pseudo.R2,Analytic.p.value)


}


# Run across timepoints (no random effect needed within TP)
tp_levels <- levels(meta$onset_timeline_combined)

per_tp_dx <- purrr::map_dfr(
  tp_levels,
  ~ run_mdmr_tp(.x, meta = meta, D_full = D_bray, ord_ids = ord_ids, use_perm_p = FALSE)
) %>%
  dplyr::mutate(p.adj = ifelse(all(is.na(Analytic.p.value)), NA_real_, p.adjust(Analytic.p.value, method = "BH")))

per_tp_dx
#   timepoint  Statistic Numer.DF  Pseudo.R2 Analytic.p.value     p.adj
# 1        t0 0.01882650        1 0.01591231        0.6606958 0.7550809
# 2      t0-6 0.04802794        1 0.03108970        0.8658763 0.7550809
# 3     t0-12 0.04507872        1 0.03591940        0.1087494 0.7550809
# 4     t0-18 0.09820395        1 0.05283506        0.4670523 0.7550809
# 5     t0-24 0.09111564        1 0.05661518        0.2078437 0.7550809
# 6     t0-30 0.06125077        1 0.03941115        0.4701198 0.7550809
# 7     t0-36 0.06864050        1 0.04807131        0.2207682 0.7550809
# 8 t0-over42 0.06607252        1 0.04063349        0.0364879 0.7550809


# make plot
library(tidyverse)
library(vegan)
library(ggplot2)

# Assumes:
# D_bray : dist with rows/cols == meta$sample_id
ord_ids <- rownames(as.matrix(D_bray))
stopifnot(identical(meta$sample_id, ord_ids))
meta <- droplevels(meta)

cap_one_tp <- function(tp, meta, D_full, ord_ids) {
  df <- meta %>% filter(onset_timeline_combined == tp) %>% droplevels()
  if (nrow(df) < 3 || n_distinct(df$Dx.Status) < 2) return(NULL)

  idx <- match(df$sample_id, ord_ids)
  if (any(is.na(idx))) return(NULL)

  subM <- as.matrix(D_full)[idx, idx, drop = FALSE]
  U <- upper.tri(subM, diag = FALSE)
  if (!any(U) || all(subM[U] == 0)) return(NULL)  # degenerate

  D_tp <- as.dist(subM)

  # CAP with Dx constrained, covariates conditioned (partialled out)
  cap <- capscale(
    D_tp ~ Dx.Status +
      Condition(Sex + Age.at.Gluten.Introduction..months. +
                HLA.Category + feeding_first_year + Delivery.Mode),
    data = df,
    add = TRUE
  )

  # Permutation test for Dx (marginal)
  aov_dx <- tryCatch(anova(cap, by = "margin", permutations = 9999), error = function(e) NULL)
  p_dx   <- if (!is.null(aov_dx) && "Dx.Status" %in% rownames(aov_dx)) aov_dx["Dx.Status", "Pr(>F)"] else NA_real_

  # Scores (CAP1/2)
  scr <- scores(cap, display = "sites", choices = 1:2)
  CAP1 <- scr[, 1]
  CAP2 <- if (ncol(scr) >= 2) scr[, 2] else rep(0, length(CAP1))

  # % variance on constrained axes (for labels)
  eig <- cap$CCA$eig
  varex <- if (length(eig)) 100 * eig / sum(eig) else numeric(0)
  v1 <- if (length(varex) >= 1) varex[1] else NA_real_
  v2 <- if (length(varex) >= 2) varex[2] else NA_real_

  tibble(
    sample_id = df$sample_id,
    timepoint = as.character(tp),
    Dx.Status = df$Dx.Status,
    CAP1 = CAP1,
    CAP2 = CAP2,
    p_dx = p_dx,          # carry p with the rows (same value within TP)
    pct_CAP1 = v1,
    pct_CAP2 = v2
  )
}

tp_levels <- levels(meta$onset_timeline_combined)
if (is.null(tp_levels)) tp_levels <- unique(meta$onset_timeline_combined)

cap_list <- lapply(tp_levels, cap_one_tp, meta = meta, D_full = D_bray, ord_ids = ord_ids)
cap_df   <- bind_rows(cap_list)   # drops NULLs safely

# Make sure we actually have rows
stopifnot(nrow(cap_df) > 0)

tp_order <- c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42")

# 1) Harmonize key type
## Rebuild per-TP FDR straight from cap_df (cap_df has one p_dx per TP carried on rows)
cap_df <- cap_df %>%
  mutate(timepoint = factor(timepoint, levels = tp_order))

p_tbl <- cap_df %>%
  dplyr::distinct(timepoint, p_dx) %>%
  dplyr::mutate(p.adj = ifelse(all(is.na(p_dx)), NA_real_, p.adjust(p_dx, method = "BH")))

## Axis labels (% variance) – median across TPs
pct_tbl <- cap_df %>%
  dplyr::distinct(timepoint, pct_CAP1, pct_CAP2) %>%
  mutate(timepoint = factor(timepoint, levels = tp_order))

lab_x <- sprintf("CAP1 (%.1f%%)", median(pct_tbl$pct_CAP1, na.rm = TRUE))
lab_y <- sprintf("CAP2 (%.1f%%)", median(pct_tbl$pct_CAP2, na.rm = TRUE))

## Label positions – MAKE SURE timepoint stays in the result
lab_pos <- cap_df %>%
  dplyr::group_by(timepoint) %>%
  dplyr::summarise(
    timepoint = dplyr::first(timepoint),   # <- keep the key explicitly
    x = max(CAP1, na.rm = TRUE),
    y = max(CAP2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::left_join(p_tbl, by = "timepoint") %>%
  dplyr::mutate(
    label = ifelse(
      is.finite(p.adj),
      sprintf("FDR = %.3g%s", p.adj, ifelse(p.adj < 0.05, " *", "")),
      "FDR = NA"
    )
  )

## Centroids (and group sizes) — qualify n()
centroids <- cap_df %>%
  dplyr::group_by(timepoint, Dx.Status) %>%
  dplyr::summarise(
    cx = mean(CAP1),
    cy = mean(CAP2),
    n  = dplyr::n(),
    .groups = "drop"
  )

## Which timepoints have >=3 per Dx group for ellipses
tmp_counts <- cap_df %>% dplyr::count(timepoint, Dx.Status, name = "n")
tp_for_ellipse <- tmp_counts %>%
  dplyr::group_by(timepoint) %>%
  dplyr::summarise(ok = all(n >= 3), .groups = "drop") %>%
  dplyr::filter(ok) %>%
  dplyr::pull(timepoint)

## Plot
total_orf_beta_contrast <- ggplot(cap_df, aes(CAP1, CAP2, color = Dx.Status)) +
  geom_point(size = 2, alpha = 0.9) +
  stat_ellipse(
  data = subset(cap_df, timepoint %in% tp_for_ellipse),
  aes(group = Dx.Status),
  type = "t",
  level = 0.68,
  linewidth = 0.5,
  alpha = 0.18
) +
  geom_point(
    data = centroids, aes(cx, cy, color = Dx.Status),
    shape = 4, stroke = 1.2, size = 3, inherit.aes = FALSE
  ) +
  facet_wrap(~ timepoint, scales = "free") +
  geom_text(
    data = lab_pos, aes(x = x, y = y, label = label),
    inherit.aes = FALSE, hjust = 1, vjust = 1, size = 3.3
  ) +
  labs(
    title = "CAP (distance-based RDA) by timepoint",
    subtitle = "Dx.Status constrained; covariates conditioned. FDR = BH across timepoints",
    x = lab_x, y = lab_y, color = "Dx.Status"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")


ggsave(total_orf_beta_contrast,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/Orf_orf_Phrog_compositional/orf_orf_results_figures/total_orf_beta_contrast.pdf",dpi = 600,width = 10,height=10)



```


# total - all terms
```{r}


```


# US - per timepoint contrast
```{r}

# Packages
library(tidyverse)
library(vegan)
library(MDMR)

# ---- Inputs ----
counts <- US.orf.abundance.clean_0.75_0.03 %>% as.data.frame() %>% as.matrix()
meta   <- US.orf.metadata.clean_0.75_0.03 %>% as.data.frame() %>% tibble::rownames_to_column("sample_id")

# Align samples
common <- intersect(colnames(counts), meta$sample_id)
counts <- counts[, common, drop = FALSE]
meta   <- meta  %>% filter(sample_id %in% common)

# Bray–Curtis on relative abundances
X        <- t(counts)  # samples x ORFs
row_sums <- rowSums(X)
X_rel    <- sweep(X, 1, pmax(row_sums, 1e-12), "/")
D_bray   <- vegdist(X_rel, method = "bray")

# Reorder metadata to match distance matrix
ord_ids <- rownames(as.matrix(D_bray))
meta    <- meta %>% slice(match(ord_ids, sample_id))

# Coerce factors (important)
meta <- meta %>%
  mutate(
    Dx.Status = forcats::fct_relevel(as.factor(Dx.Status), "CONTROL"),
    Country = forcats::fct_relevel(as.factor(Country), "ITALY"),
    onset_timeline_combined = as.factor(onset_timeline_combined),
    patientID = as.factor(patientID),
    Sex = as.factor(Sex),
    HLA.Category = as.factor(HLA.Category),
    feeding_first_year = as.factor(feeding_first_year),
    Delivery.Mode = as.factor(Delivery.Mode),
    Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.)
  )

# ---- mixed MDMR: fixed effects + random intercept for patientID ----
fmla_mixed <- ~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. +
               HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID)

# Tip for Bray (non-Euclidean): use.ssd ~ 0.99 can speed up
fit_mmdmr <- mixed.mdmr(
  fmla = fmla_mixed,
  data = meta,
  D    = D_bray,
  use.ssd = 0.99,   # set to 1.00 if you prefer full SSD (slower)
  ncores = 1        # >1 uses forking; keep 1 on Windows
)

# Results
res_tab <- summary(fit_mmdmr)           # Statistic, Numer DF, p-value per term
res_tab <- data.frame(res_tab) %>%
  rownames_to_column("term") %>%
  mutate(p.adj = p.adjust(`p.value`, method = "BH"))



ord_ids <- rownames(as.matrix(D_bray))
stopifnot(identical(meta$sample_id, ord_ids))

# Helper that runs MDMR for ONE timepoint, adjusting for covariates
run_mdmr_tp <- function(tp, meta, D_full, ord_ids, use_perm_p = FALSE, nperm = 4999, seed = 1L) {
  print(tp)
  df <- meta %>% filter(onset_timeline_combined == tp) %>% droplevels()
  if (nrow(df) < 3 || n_distinct(df$Dx.Status) < 2) {
    return(tibble(timepoint = tp, term = "Dx.Status", Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }
  idx <- match(df$sample_id, ord_ids)
  if (any(is.na(idx))) {
    return(tibble(timepoint = tp, term = "Dx.Status", Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }
  D_tp <- as.dist(as.matrix(D_full)[idx, idx, drop = FALSE])


  # Quick degeneracy check (all pairwise distances == 0)
  M <- as.matrix(D_tp); U <- upper.tri(M, diag = FALSE)
  if (!any(U) || all(M[U] == 0)) {
    print("test")
    return(tibble(timepoint = tp, term = "Dx.Status",
                  Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }

  # Build X as a DATA FRAME with factors kept as factors
  X_df <- df %>%
    transmute(
      Dx.Status = factor(Dx.Status),
      Country = factor(Country),
      Sex = factor(Sex),
      Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.),
      HLA.Category = factor(HLA.Category),
      feeding_first_year = factor(feeding_first_year),
      Delivery.Mode = factor(Delivery.Mode)
    )

  # mdmr() expects X= (matrix or data.frame) and D=/G=
  # Use analytic p-values by default; or set use_perm_p = TRUE for permutation p's
  set.seed(seed)
  fit <- tryCatch(
    MDMR::mdmr(
      X = X_df,
      D = D_tp,
      perm.p = use_perm_p,
      nperm = if (use_perm_p) nperm else 500  # ignored if perm.p=FALSE
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    print("fit is still null")
    return(tibble(timepoint = tp, term = "Dx.Status",
                  Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }

  
  sm <- data.frame(summary(fit)) %>%
        rownames_to_column("term") %>%
    filter(term == "Dx.Status1") %>%
    mutate(timepoint = tp) %>%
    select(timepoint,Statistic,Numer.DF,Pseudo.R2,Analytic.p.value)


}


# Run across timepoints (no random effect needed within TP)
tp_levels <- levels(meta$onset_timeline_combined)

per_tp_dx <- purrr::map_dfr(
  tp_levels,
  ~ run_mdmr_tp(.x, meta = meta, D_full = D_bray, ord_ids = ord_ids, use_perm_p = FALSE)
) %>%
  dplyr::mutate(p.adj = ifelse(all(is.na(Analytic.p.value)), NA_real_, p.adjust(Analytic.p.value, method = "BH")))

per_tp_dx
#   timepoint  Statistic Numer.DF  Pseudo.R2 Analytic.p.value     p.adj
# 1        t0 0.01882650        1 0.01591231        0.6606958 0.7550809
# 2      t0-6 0.04802794        1 0.03108970        0.8658763 0.7550809
# 3     t0-12 0.04507872        1 0.03591940        0.1087494 0.7550809
# 4     t0-18 0.09820395        1 0.05283506        0.4670523 0.7550809
# 5     t0-24 0.09111564        1 0.05661518        0.2078437 0.7550809
# 6     t0-30 0.06125077        1 0.03941115        0.4701198 0.7550809
# 7     t0-36 0.06864050        1 0.04807131        0.2207682 0.7550809
# 8 t0-over42 0.06607252        1 0.04063349        0.0364879 0.7550809


# make plot
ord_ids <- rownames(as.matrix(D_bray))
stopifnot(identical(meta$sample_id, ord_ids))
meta <- droplevels(meta)

cap_one_tp <- function(tp, meta, D_full, ord_ids) {
  df <- meta %>% filter(onset_timeline_combined == tp) %>% droplevels()
  if (nrow(df) < 3 || n_distinct(df$Dx.Status) < 2) return(NULL)

  idx <- match(df$sample_id, ord_ids)
  if (any(is.na(idx))) return(NULL)

  subM <- as.matrix(D_full)[idx, idx, drop = FALSE]
  U <- upper.tri(subM, diag = FALSE)
  if (!any(U) || all(subM[U] == 0)) return(NULL)  # degenerate

  D_tp <- as.dist(subM)

  # CAP: Dx as constrained term, adjust (Condition) for covariates
  cap <- capscale(
    D_tp ~ Dx.Status +
      Condition(Sex + Age.at.Gluten.Introduction..months. +
                HLA.Category + feeding_first_year + Delivery.Mode),
    data = df, add = TRUE # add=TRUE (Cailliez) guards non-Euclidean cases
  )

  # Permutation test for Dx
  aov_dx <- anova(cap, by = "margin", permutations = 9999)
  p_dx <- tryCatch(aov_dx["Dx.Status", "Pr(>F)"], error = function(e) NA_real_)
  p_dx_adj <- p_dx  # (we’ll BH-adjust across TPs after mapping)

  # Scores on constrained axes (CAP1/CAP2)
  scr <- scores(cap, display = "sites", choices = 1:2)
  df_out <- tibble(
    sample_id = df$sample_id,
    timepoint = as.character(tp),
    Dx.Status = df$Dx.Status,
    CAP1 = scr[,1],
    CAP2 = if (ncol(scr) >= 2) scr[,2] else 0
  )

  # percent variance for constrained axes
  eig <- cap$CCA$eig
  varex <- if (length(eig)) 100 * eig / sum(eig) else c()
  attr(df_out, "p_dx") <- as.numeric(p_dx)
  attr(df_out, "%CAP1") <- if (length(varex) >= 1) varex[1] else NA_real_
  attr(df_out, "%CAP2") <- if (length(varex) >= 2) varex[2] else NA_real_
  df_out
}

tp_levels <- levels(meta$onset_timeline_combined)
cap_list <- lapply(tp_levels, cap_one_tp, meta = meta, D_full = D_bray, ord_ids = ord_ids)
cap_df <- bind_rows(cap_list)

# Collect per-TP p-values and BH adjust
p_tbl <- tibble(
  timepoint = map_chr(cap_list, ~ if (is.null(.x)) NA_character_ else unique(.x$timepoint))) %>%
  filter(!is.na(timepoint)) %>%
  mutate(p.value = map_dbl(cap_list[!map_lgl(cap_list, is.null)], ~ attr(.x, "p_dx"))) %>%
  mutate(p.adj = ifelse(all(is.na(p.value)), NA_real_, p.adjust(p.value, method = "BH")))

# Axis labels: use median % across TPs for readability
pct_tbl <- tibble(
  timepoint = map_chr(cap_list, ~ if (is.null(.x)) NA_character_ else unique(.x$timepoint)),
  cap1 = map_dbl(cap_list, ~ if (is.null(.x)) NA_real_ else attr(.x, "%CAP1")),
  cap2 = map_dbl(cap_list, ~ if (is.null(.x)) NA_real_ else attr(.x, "%CAP2"))
) %>% filter(!is.na(timepoint))

lab_x <- sprintf("CAP1 (%.1f%%)", median(pct_tbl$cap1, na.rm = TRUE))
lab_y <- sprintf("CAP2 (%.1f%%)", median(pct_tbl$cap2, na.rm = TRUE))


# Label positions + FDR
lab_pos <- cap_df %>%
  group_by(timepoint) %>%
  #summarise(x = max(CAP1, na.rm = TRUE), y = max(CAP2, na.rm = TRUE), .groups = "drop") %>%
  left_join(p_tbl, by = "timepoint") %>%
  mutate(label = ifelse(is.finite(p.adj),
                        sprintf("FDR = %.3g%s", p.adj, ifelse(p.adj < 0.05, " *", "")),
                        "FDR = NA"))

# Plot (points, centroids, ellipses where n>=3 per group)
centroids <- cap_df %>%
  group_by(timepoint, Dx.Status) %>%
  mutate(cx = mean(CAP1), cy = mean(CAP2), n = n(), .groups = "drop")

tp_for_ellipse <- cap_df %>%
  count(timepoint, Dx.Status) %>%
  group_by(timepoint) %>%
  summarise(ok = all(n >= 3), .groups = "drop") %>%
  filter(ok) %>% pull(timepoint)

US_orf_beta_contrast <- ggplot(cap_df, aes(CAP1, CAP2, color = Dx.Status)) +
  geom_point(size = 2, alpha = 0.9) +
  geom_stat_ellipse(data = subset(cap_df, timepoint %in% tp_for_ellipse),
                    aes(group = Dx.Status), type = "t", level = 0.68, linewidth = 0.5, alpha = 0.18) +
  geom_point(data = centroids, aes(cx, cy, color = Dx.Status), shape = 4, stroke = 1.2, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ timepoint, scales = "free") +
  geom_text(data = lab_pos, aes(x = x, y = y, label = label),
            inherit.aes = FALSE, hjust = 1, vjust = 1, size = 3.3) +
  labs(title = "US CAP (distance-based RDA) by timepoint",
       subtitle = "Dx.Status constrained, covariates conditioned; FDR from per-TP PERMANOVA on CAP model",
       x = lab_x, y = lab_y, color = "Dx.Status") +
  theme_minimal() +
  theme(legend.position = "bottom")




```

# US - all terms
```{r}


```


# Italy - per timepoint contrast
```{r}

# Packages
library(tidyverse)
library(vegan)
library(MDMR)

# ---- Inputs ----
counts <- Italy.orf.abundance.clean_0.75_0.03 %>% as.data.frame() %>% as.matrix()
meta   <- Italy.orf.metadata.clean_0.75_0.03 %>% as.data.frame() %>% tibble::rownames_to_column("sample_id")

# Align samples
common <- intersect(colnames(counts), meta$sample_id)
counts <- counts[, common, drop = FALSE]
meta   <- meta  %>% filter(sample_id %in% common)

# Bray–Curtis on relative abundances
X        <- t(counts)  # samples x ORFs
row_sums <- rowSums(X)
X_rel    <- sweep(X, 1, pmax(row_sums, 1e-12), "/")
D_bray   <- vegdist(X_rel, method = "bray")

# Reorder metadata to match distance matrix
ord_ids <- rownames(as.matrix(D_bray))
meta    <- meta %>% slice(match(ord_ids, sample_id))

# Coerce factors (important)
meta <- meta %>%
  mutate(
    Dx.Status = forcats::fct_relevel(as.factor(Dx.Status), "CONTROL"),
    Country = forcats::fct_relevel(as.factor(Country), "ITALY"),
    onset_timeline_combined = as.factor(onset_timeline_combined),
    patientID = as.factor(patientID),
    Sex = as.factor(Sex),
    HLA.Category = as.factor(HLA.Category),
    feeding_first_year = as.factor(feeding_first_year),
    Delivery.Mode = as.factor(Delivery.Mode),
    Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.)
  )

# ---- mixed MDMR: fixed effects + random intercept for patientID ----
fmla_mixed <- ~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. +
               HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID)

# Tip for Bray (non-Euclidean): use.ssd ~ 0.99 can speed up
fit_mmdmr <- mixed.mdmr(
  fmla = fmla_mixed,
  data = meta,
  D    = D_bray,
  use.ssd = 0.99,   # set to 1.00 if you prefer full SSD (slower)
  ncores = 1        # >1 uses forking; keep 1 on Windows
)

# Results
res_tab <- summary(fit_mmdmr)           # Statistic, Numer DF, p-value per term
res_tab <- data.frame(res_tab) %>%
  rownames_to_column("term") %>%
  mutate(p.adj = p.adjust(`p.value`, method = "BH"))



ord_ids <- rownames(as.matrix(D_bray))
stopifnot(identical(meta$sample_id, ord_ids))

# Helper that runs MDMR for ONE timepoint, adjusting for covariates
run_mdmr_tp <- function(tp, meta, D_full, ord_ids, use_perm_p = FALSE, nperm = 4999, seed = 1L) {
  print(tp)
  df <- meta %>% filter(onset_timeline_combined == tp) %>% droplevels()
  if (nrow(df) < 3 || n_distinct(df$Dx.Status) < 2) {
    return(tibble(timepoint = tp, term = "Dx.Status", Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }
  idx <- match(df$sample_id, ord_ids)
  if (any(is.na(idx))) {
    return(tibble(timepoint = tp, term = "Dx.Status", Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }
  D_tp <- as.dist(as.matrix(D_full)[idx, idx, drop = FALSE])


  # Quick degeneracy check (all pairwise distances == 0)
  M <- as.matrix(D_tp); U <- upper.tri(M, diag = FALSE)
  if (!any(U) || all(M[U] == 0)) {
    print("test")
    return(tibble(timepoint = tp, term = "Dx.Status",
                  Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }

  # Build X as a DATA FRAME with factors kept as factors
  X_df <- df %>%
    transmute(
      Dx.Status = factor(Dx.Status),
      Country = factor(Country),
      Sex = factor(Sex),
      Age.at.Gluten.Introduction..months. = as.numeric(Age.at.Gluten.Introduction..months.),
      HLA.Category = factor(HLA.Category),
      feeding_first_year = factor(feeding_first_year),
      Delivery.Mode = factor(Delivery.Mode)
    )

  # mdmr() expects X= (matrix or data.frame) and D=/G=
  # Use analytic p-values by default; or set use_perm_p = TRUE for permutation p's
  set.seed(seed)
  fit <- tryCatch(
    MDMR::mdmr(
      X = X_df,
      D = D_tp,
      perm.p = use_perm_p,
      nperm = if (use_perm_p) nperm else 500  # ignored if perm.p=FALSE
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    print("fit is still null")
    return(tibble(timepoint = tp, term = "Dx.Status",
                  Statistic = NA_real_, Numer.DF = NA_real_, p.value = NA_real_))
  }

  
  sm <- data.frame(summary(fit)) %>%
        rownames_to_column("term") %>%
    filter(term == "Dx.Status1") %>%
    mutate(timepoint = tp) %>%
    select(timepoint,Statistic,Numer.DF,Pseudo.R2,Analytic.p.value)


}


# Run across timepoints (no random effect needed within TP)
tp_levels <- levels(meta$onset_timeline_combined)

per_tp_dx <- purrr::map_dfr(
  tp_levels,
  ~ run_mdmr_tp(.x, meta = meta, D_full = D_bray, ord_ids = ord_ids, use_perm_p = FALSE)
) %>%
  dplyr::mutate(p.adj = ifelse(all(is.na(Analytic.p.value)), NA_real_, p.adjust(Analytic.p.value, method = "BH")))

per_tp_dx
#   timepoint  Statistic Numer.DF  Pseudo.R2 Analytic.p.value     p.adj
# 1        t0 0.01882650        1 0.01591231        0.6606958 0.7550809
# 2      t0-6 0.04802794        1 0.03108970        0.8658763 0.7550809
# 3     t0-12 0.04507872        1 0.03591940        0.1087494 0.7550809
# 4     t0-18 0.09820395        1 0.05283506        0.4670523 0.7550809
# 5     t0-24 0.09111564        1 0.05661518        0.2078437 0.7550809
# 6     t0-30 0.06125077        1 0.03941115        0.4701198 0.7550809
# 7     t0-36 0.06864050        1 0.04807131        0.2207682 0.7550809
# 8 t0-over42 0.06607252        1 0.04063349        0.0364879 0.7550809


# make plot
ord_ids <- rownames(as.matrix(D_bray))
stopifnot(identical(meta$sample_id, ord_ids))
meta <- droplevels(meta)

cap_one_tp <- function(tp, meta, D_full, ord_ids) {
  df <- meta %>% filter(onset_timeline_combined == tp) %>% droplevels()
  if (nrow(df) < 3 || n_distinct(df$Dx.Status) < 2) return(NULL)

  idx <- match(df$sample_id, ord_ids)
  if (any(is.na(idx))) return(NULL)

  subM <- as.matrix(D_full)[idx, idx, drop = FALSE]
  U <- upper.tri(subM, diag = FALSE)
  if (!any(U) || all(subM[U] == 0)) return(NULL)  # degenerate

  D_tp <- as.dist(subM)

  # CAP: Dx as constrained term, adjust (Condition) for covariates
  cap <- capscale(
    D_tp ~ Dx.Status +
      Condition(Sex + Age.at.Gluten.Introduction..months. +
                HLA.Category + feeding_first_year + Delivery.Mode),
    data = df, add = TRUE # add=TRUE (Cailliez) guards non-Euclidean cases
  )

  # Permutation test for Dx
  aov_dx <- anova(cap, by = "margin", permutations = 9999)
  p_dx <- tryCatch(aov_dx["Dx.Status", "Pr(>F)"], error = function(e) NA_real_)
  p_dx_adj <- p_dx  # (we’ll BH-adjust across TPs after mapping)

  # Scores on constrained axes (CAP1/CAP2)
  scr <- scores(cap, display = "sites", choices = 1:2)
  df_out <- tibble(
    sample_id = df$sample_id,
    timepoint = as.character(tp),
    Dx.Status = df$Dx.Status,
    CAP1 = scr[,1],
    CAP2 = if (ncol(scr) >= 2) scr[,2] else 0
  )

  # percent variance for constrained axes
  eig <- cap$CCA$eig
  varex <- if (length(eig)) 100 * eig / sum(eig) else c()
  attr(df_out, "p_dx") <- as.numeric(p_dx)
  attr(df_out, "%CAP1") <- if (length(varex) >= 1) varex[1] else NA_real_
  attr(df_out, "%CAP2") <- if (length(varex) >= 2) varex[2] else NA_real_
  df_out
}

tp_levels <- levels(meta$onset_timeline_combined)
cap_list <- lapply(tp_levels, cap_one_tp, meta = meta, D_full = D_bray, ord_ids = ord_ids)
cap_df <- bind_rows(cap_list)

# Collect per-TP p-values and BH adjust
p_tbl <- tibble(
  timepoint = map_chr(cap_list, ~ if (is.null(.x)) NA_character_ else unique(.x$timepoint))) %>%
  filter(!is.na(timepoint)) %>%
  mutate(p.value = map_dbl(cap_list[!map_lgl(cap_list, is.null)], ~ attr(.x, "p_dx"))) %>%
  mutate(p.adj = ifelse(all(is.na(p.value)), NA_real_, p.adjust(p.value, method = "BH")))

# Axis labels: use median % across TPs for readability
pct_tbl <- tibble(
  timepoint = map_chr(cap_list, ~ if (is.null(.x)) NA_character_ else unique(.x$timepoint)),
  cap1 = map_dbl(cap_list, ~ if (is.null(.x)) NA_real_ else attr(.x, "%CAP1")),
  cap2 = map_dbl(cap_list, ~ if (is.null(.x)) NA_real_ else attr(.x, "%CAP2"))
) %>% filter(!is.na(timepoint))

lab_x <- sprintf("CAP1 (%.1f%%)", median(pct_tbl$cap1, na.rm = TRUE))
lab_y <- sprintf("CAP2 (%.1f%%)", median(pct_tbl$cap2, na.rm = TRUE))


# Label positions + FDR
lab_pos <- cap_df %>%
  group_by(timepoint) %>%
  #summarise(x = max(CAP1, na.rm = TRUE), y = max(CAP2, na.rm = TRUE), .groups = "drop") %>%
  left_join(p_tbl, by = "timepoint") %>%
  mutate(label = ifelse(is.finite(p.adj),
                        sprintf("FDR = %.3g%s", p.adj, ifelse(p.adj < 0.05, " *", "")),
                        "FDR = NA"))

# Plot (points, centroids, ellipses where n>=3 per group)
centroids <- cap_df %>%
  group_by(timepoint, Dx.Status) %>%
  mutate(cx = mean(CAP1), cy = mean(CAP2), n = n(), .groups = "drop")

tp_for_ellipse <- cap_df %>%
  count(timepoint, Dx.Status) %>%
  group_by(timepoint) %>%
  summarise(ok = all(n >= 3), .groups = "drop") %>%
  filter(ok) %>% pull(timepoint)

Italy_orf_beta_contrast <- ggplot(cap_df, aes(CAP1, CAP2, color = Dx.Status)) +
  geom_point(size = 2, alpha = 0.9) +
  geom_stat_ellipse(data = subset(cap_df, timepoint %in% tp_for_ellipse),
                    aes(group = Dx.Status), type = "t", level = 0.68, linewidth = 0.5, alpha = 0.18) +
  geom_point(data = centroids, aes(cx, cy, color = Dx.Status), shape = 4, stroke = 1.2, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ timepoint, scales = "free") +
  geom_text(data = lab_pos, aes(x = x, y = y, label = label),
            inherit.aes = FALSE, hjust = 1, vjust = 1, size = 3.3) +
  labs(title = "Italy CAP (distance-based RDA) by timepoint",
       subtitle = "Dx.Status constrained, covariates conditioned; FDR from per-TP PERMANOVA on CAP model",
       x = lab_x, y = lab_y, color = "Dx.Status") +
  theme_minimal() +
  theme(legend.position = "bottom")




```

# Italy - all terms
```{r}


```



# annotation files
```{r}

orf_annotation <- read.delim("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/orf/annotation_files/genomes_and_unresolved_edges_mmseqs_lca.tsv",header = FALSE)

View(orf_annotation)

```




