---
title: "Celiac_Phage_orfs_analysis_Limma_newAssembly_timeOnset_finalModel"
output: html_document
date: "2025-07-25"
editor_options: 
  chunk_output_type: console
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table);packageVersion("data.table")
library(tidyverse);packageVersion("tidyverse")
library(ggplot2);packageVersion("ggplot2")
library(broom);packageVersion("broom")
library(vroom);packageVersion("vroom")
library(limma);packageVersion("limma")
library(edgeR);packageVersion("edgeR")
library(pheatmap);packageVersion("pheatmap")
library(survival);packageVersion("survival")
```



# functions
```{r}

filter_and_save_wide_data <- function(genesCoverm, 
                                      covered_fraction_threshold = 0.75, 
                                      output_path = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/orf/data_correct/total/") {
  
  # Set column names to match the original structure
  colnames(genesCoverm) <- c("sampleID", "ORFID", "RPKM", "ReadCount", "Variance", "Mean", "covered_fraction", "covered_bases")
  
  # Apply coverage filter
  genesCoverm2 <- genesCoverm %>%
    mutate(
      ReadCount_modified = ifelse(covered_fraction >= covered_fraction_threshold, ReadCount, 0),
      Changed = ifelse(ReadCount != ReadCount_modified, 1, 0)
    ) %>%
    mutate(ReadCount = ReadCount_modified) %>%
    select(-ReadCount_modified)
  
  # Count the number of changes
  num_changes <- sum(genesCoverm2$Changed)
  print(paste("Number of changes:", num_changes))
  
  # Convert to wide format
  wide_rpkm_genes <- reshape2::dcast(genesCoverm2, ORFID ~ sampleID, value.var = "ReadCount")
  colnames(wide_rpkm_genes) <- gsub("_stats", "", colnames(wide_rpkm_genes))
  
  # Filter out rows with all zeroes
  filtered_wide_rpkm_genes <- wide_rpkm_genes %>%
    filter(rowSums(across(where(is.numeric))) != 0)
  
  # Save the filtered wide format data
  rds_filename_wide <- paste0("wide_rpkm_genes_", covered_fraction_threshold * 100, "Cov.rds")
  
  saveRDS(filtered_wide_rpkm_genes, file = file.path(output_path, rds_filename_wide))
  
  return(filtered_wide_rpkm_genes)
}



```


# load metadata
```{r}

metadata <- read.csv("~/Handley Lab Dropbox/16S/Celiac/metadata/Updated_Metadata_with_Onset_Timeline.csv") %>%
            select(-Sample.Name.External.ID) %>%
            select(-Classification) %>%
            #mutate(Onset_timepoint = ifelse(Onset.Dx < 30,"pre_30","post_30")) %>%
            column_to_rownames("X")
dim(metadata)
# 335  24


# check the data distributions for feeding types
metadata.clean.patient <- metadata.clean %>% distinct(patientID,.keep_all = TRUE)
table(metadata.clean.patient$Country,metadata.clean.patient$Delivery.Mode)
  #       Vaginal C-Section
  # ITALY      11        22
  # USA        31         2



# check US and Italy onset time difference
metadata.check.onset <- metadata %>%
                        group_by(patientID) %>%
                        mutate(time_length_to_onset = max(month)) %>%
                        distinct(patientID,Country,time_length_to_onset)

patient_counts <- metadata.check.onset %>%
                  group_by(Country) %>%
                  summarise(n = n(), .groups = "drop")


labels_with_n <- setNames(
  paste0(patient_counts$Country, " (n = ", patient_counts$n, ")"),
  patient_counts$Country
)


metadata.check.onset.plot <- ggplot(metadata.check.onset,aes(x = Country,y = as.numeric(time_length_to_onset))) +
                             geom_boxplot() +
                             geom_point(size = 3,position = position_jitter(width = 0.2), alpha = 0.6) +
                             scale_x_discrete(labels = labels_with_n) +
                              geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.2) +
                              stat_compare_means(
                                method = "t.test",
                                label  = "p.format",
                                label.y = 90,
                                label.x = 1.5
                              ) +
                             labs(x = "",y = "Time Length to Onset (Month)") +
                             theme_bw()

mean(metadata.check.onset %>% filter(Country == "ITALY") %>% pull(time_length_to_onset))

ggsave(metadata.check.onset.plot,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/metadata_onset_time_length_boxplot_plot.png", width = 5, height = 6, dpi = 300)        





metadata.check.gluten <- metadata %>%
                        group_by(patientID) %>%
                        distinct(patientID,Country,Age.at.Gluten.Introduction..months.)

patient_counts <- metadata.check.gluten %>%
                  group_by(Country) %>%
                  summarise(n = n(), .groups = "drop")


labels_with_n <- setNames(
  paste0(patient_counts$Country, " (n = ", patient_counts$n, ")"),
  patient_counts$Country
)


metadata.check.gluten.plot <- ggplot(metadata.check.gluten,aes(x = Country,y = as.numeric(Age.at.Gluten.Introduction..months.))) +
                             geom_boxplot() +
                             geom_point(size = 3,position = position_jitter(width = 0.2), alpha = 0.6) +
                             scale_x_discrete(labels = labels_with_n) +
                              geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.2) +
                              stat_compare_means(
                                method = "t.test",
                                label  = "p.format",
                                label.y = 90,
                                label.x = 1.5
                              ) +
                             labs(x = "",y = "Age of gluten food introduction (Month)") +
                             theme_bw()

mean(metadata.check.onset %>% filter(Country == "ITALY") %>% pull(time_length_to_onset))

ggsave(metadata.check.onset.plot,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/metadata_onset_time_length_boxplot_plot.png", width = 5, height = 6, dpi = 300)   

```



# 75% mapping rate filtering and 3% prevlance filtering on total data
```{r}

orf <- vroom("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/Orf_Coverm_concatenated.txt",col_names = FALSE)
# we have 340 samples and 46594 orfs in total after phage pipeline


filter_and_save_wide_data(orf, 
                          covered_fraction_threshold = 0.75, 
                          output_path = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/")
# Number of changes: 185957

orf.processed <- readRDS("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/wide_rpkm_genes_75Cov.rds") %>%
                 column_to_rownames("ORFID")

# then let's check colSums and ensure there is no sample have all 0s across orfs
sample_all_0 <- names(colSums(orf.processed)[colSums(orf.processed) == 0])
# "NovaSeq_N978_I13149_39572_Celiac_Leonard_Stool_01_GEMM_093_24M"       "NovaSeq_N978_I13150_39573_Celiac_Leonard_Stool_01_GEMM_093_30M"      
# "NovaSeq_N978_I13151_39574_Celiac_Leonard_Stool_01_GEMM_093_36M"       "NovaSeq_N978_I13152_39575_Celiac_Leonard_Stool_01_GEMM_093_48M"      
# "NovaSeq_N982_I13323_39837_Celiac_Leonard_Stool_02_GEMM_077_12M"       "NovaSeq_N983_I13417_39799_Leonard_Human_stool_Celiac_02_GEMM_006_60M"


# let's remove the 6 0 counts samples
orf.processed <- orf.processed %>%
                    select(-all_of(sample_all_0))
dim(orf.processed)
# 46102   334


# here we found two duplicated samples:
# "NovaSeq_N978_I13164_39587_Celiac_Leonard_Stool_01_GEMM_106_12M_stats" and  "NovaSeq_N983_I13413_39587_Celiac_Leonard_Stool_01_GEMM_106_12M_stats"
# after checking with my note, the N983 01_GEMM_106_12M sample should actually be water
orf.processed <- orf.processed %>%
                 select(-NovaSeq_N983_I13413_39587_Celiac_Leonard_Stool_01_GEMM_106_12M)
# 46102   333

orf.count <- orf.processed

colnames(orf.count) <- gsub("NovaSeq_N966_|NovaSeq_N978_|NovaSeq_N979_|NovaSeq_N980_|NovaSeq_N981_|NovaSeq_N982_|NovaSeq_N983_|_stats","",colnames(orf.count))

colnames(orf.count) <- str_replace_all(colnames(orf.count),"Leonard_Human_stool_Celiac","Celiac_Leonard_Stool")
colnames(orf.count) <- sapply(strsplit(colnames(orf.count),"_Stool_|_Stool_Celiac_"),"[",2)
colnames(orf.count) <- ifelse(colnames(orf.count) %like% "_6Y",gsub("_6Y","_72M",colnames(orf.count)),
                                           ifelse(colnames(orf.count) %like% "_7Y",gsub("_7Y","_84M",colnames(orf.count)),colnames(orf.count)))

orf.count.table_0.75 <- t(orf.count) 
# 333 46102


orf.abundance.table_0.75 <- merge(metadata,orf.count.table_0.75,by = 0) %>% # 329 samples left
                              filter(feeding_first_year != "Unknown") %>%  # 312 samples left
                              filter(Age.at.Gluten.Introduction..months. != "Unknown") # 308 sample left
# 308 samples x  46127 metadata+orfs

all_0_contigs <- names(colSums(orf.abundance.table_0.75[25:46127])[colSums(orf.abundance.table_0.75[25:46127]) == 0])
# 837 contigs become 0 through all samples after the 25 samples were removed from the above step
# so we need to remove those contigs

orf.abundance.table_0.75 <- orf.abundance.table_0.75 %>%
                            select(-all_of(all_0_contigs))
# 308 samples x 45290 metadata+orfs
# 45265 orfs

orf.abundance.table <- orf.abundance.table_0.75 %>%
                       select(c("Row.names",26:45290)) %>%
                       column_to_rownames("Row.names") %>% 
                       t(.) %>%
                       data.frame(.)
colnames(orf.abundance.table) <- gsub("X","",colnames(orf.abundance.table))


# check prevelence filtering
prevalence <- rowSums(orf.abundance.table > 0) / ncol(orf.abundance.table)
table(prevalence > 0.03)  # 5% threshold
# FALSE  TRUE 
# 42952  2313

# Filter ORFs with prevalence > 3%
orf.abundance.clean <- orf.abundance.table[prevalence > 0.03, ]
      
dim(orf.abundance.clean)
# 2313  308 (3% prev)

# check if there are any samples become 0 counts
names(colSums(orf.abundance.clean)[colSums(orf.abundance.clean) == 0])
  # "01_GEMM_154_30M" "02_GEMM_021_36M"

# remove the two 0 counts samples
# we will have to remove sample 02_GEMM_021_36M because it becomes 0 counts at this step
orf.abundance.clean <- orf.abundance.clean %>%
                       select(-c("01_GEMM_154_30M","02_GEMM_021_36M"))
# 2313  306
# 577 306

write.csv(orf.abundance.clean,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_orf.abundance.clean.csv",quote = FALSE,row.names = TRUE)


metadata.clean <- orf.abundance.table_0.75[1:25] %>%
                  filter(!Row.names %in% c("01_GEMM_154_30M","02_GEMM_021_36M")) %>%
                  column_to_rownames("Row.names")
# 306  24


# write.csv(metadata.clean,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_metadata.clean.csv",quote = FALSE,row.names = TRUE)

# check if the column names of abundance table match the order of the row names of metadata table
all(rownames(metadata.clean) == colnames(orf.abundance.clean))
# TRUE

count_matrix <- as.matrix(orf.abundance.clean[, sapply(orf.abundance.clean, is.numeric)])
sparsity <- sum(count_matrix == 0) / length(count_matrix)
# 0.9518756

```



# 75% mapping rate filtering and 3% prevlance filtering on US
```{r}


US.orf <- orf %>%
             filter(X1 %like% "_01_GEMM_")
# 208 samples x 46594 orfs


filter_and_save_wide_data(US.orf, 
                          covered_fraction_threshold = 0.75, 
                          output_path = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/")
# "Number of changes: 132493"


# here we found two duplicated samples:
# "NovaSeq_N978_I13164_39587_Celiac_Leonard_Stool_01_GEMM_106_12M_stats" and  "NovaSeq_N983_I13413_39587_Celiac_Leonard_Stool_01_GEMM_106_12M_stats"
# after checking with my note, the N983 01_GEMM_106_12M sample should actually be water
# so we will just remove it
US_orf.processed <- readRDS("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/wide_rpkm_genes_75Cov.rds") %>%
                   column_to_rownames("ORFID") %>%
                   select(-NovaSeq_N983_I13413_39587_Celiac_Leonard_Stool_01_GEMM_106_12M)


US_sample_all_0 <- names(colSums(US_orf.processed)[colSums(US_orf.processed) == 0])
#"NovaSeq_N978_I13149_39572_Celiac_Leonard_Stool_01_GEMM_093_24M" "NovaSeq_N978_I13150_39573_Celiac_Leonard_Stool_01_GEMM_093_30M"
#"NovaSeq_N978_I13151_39574_Celiac_Leonard_Stool_01_GEMM_093_36M" "NovaSeq_N978_I13152_39575_Celiac_Leonard_Stool_01_GEMM_093_48M"


US.orf.count <- US_orf.processed %>%
                   select(-all_of(c(US_sample_all_0))) 
# 35430   203

colnames(US.orf.count) <- gsub("NovaSeq_N966_|NovaSeq_N978_|NovaSeq_N979_|NovaSeq_N980_|NovaSeq_N981_|NovaSeq_N982_|NovaSeq_N983_|_stats","",colnames(US.orf.count))
colnames(US.orf.count) <- str_replace_all(colnames(US.orf.count),"Leonard_Human_stool_Celiac","Celiac_Leonard_Stool")
colnames(US.orf.count) <- sapply(strsplit(colnames(US.orf.count),"_Stool_|_Stool_Celiac_"),"[",2)
colnames(US.orf.count) <- ifelse(colnames(US.orf.count) %like% "_6Y",gsub("_6Y","_72M",colnames(US.orf.count)),
                                           ifelse(colnames(US.orf.count) %like% "_7Y",gsub("_7Y","_84M",colnames(US.orf.count)),colnames(US.orf.count)))

US.orf.count.table_0.75 <- t(US.orf.count) 
# 203 35430


US.orf.abundance.table_0.75 <- merge(metadata,US.orf.count.table_0.75,by = 0) %>% # 201 samples left
                              filter(feeding_first_year != "Unknown") %>%  # 201 samples left
                              filter(Age.at.Gluten.Introduction..months. != "Unknown") # 197 sample left
# 197 samples x  35455 metadata+orfs


US.all_0_contigs <- names(colSums(US.orf.abundance.table_0.75[25:35455])[colSums(US.orf.abundance.table_0.75[25:35455]) == 0])
length(US.all_0_contigs)
# 423 contigs become 0 through all samples after the 6 samples were removed from the above step
# so we need to remove those contigs

US.orf.abundance.table_0.75 <- US.orf.abundance.table_0.75 %>%
                            select(-all_of(US.all_0_contigs))
# 197 samples x 35032 metadata+orfs
# 35007 orfs

US.orf.abundance.table <- US.orf.abundance.table_0.75 %>%
                       select(c("Row.names",26:35032)) %>%
                       column_to_rownames("Row.names") %>% 
                       t(.) %>%
                       data.frame(.)
colnames(US.orf.abundance.table) <- gsub("X","",colnames(US.orf.abundance.table))


# check prevelence filtering
prevalence <- rowSums(US.orf.abundance.table > 0) / ncol(US.orf.abundance.table)
table(prevalence > 0.03)  
# FALSE  TRUE 
# 31098  3909 

# Filter ORFs with prevalence > 3%
US.orf.abundance.table <- US.orf.abundance.table[prevalence > 0.03, ]
      
dim(US.orf.abundance.table)
# 3909  197

# check if there are any samples become 0 counts
names(colSums(US.orf.abundance.table)[colSums(US.orf.abundance.table) == 0])
# 0

US.orf.abundance.clean <- US.orf.abundance.table
# 3909  197


US.metadata.clean <- US.orf.abundance.table_0.75[1:25] %>%
                     mutate(Onset_timepoint = ifelse(Onset.Dx < 30,"pre_30","post_30")) %>%
                     column_to_rownames("Row.names")
# 197  25


US.metadata.celiac.clean <- US.metadata.clean %>% filter(Dx.Status == "CELIAC")
# 98 25

US.orf.abundance.celiac.clean <- US.orf.abundance.table %>%
                                 select(rownames(US.metadata.celiac.clean))

# check prevelence filtering
prevalence <- rowSums(US.orf.abundance.celiac.clean > 0) / ncol(US.orf.abundance.celiac.clean)
table(prevalence > 0.03)
# FALSE  TRUE 
# 1336  2573

US.orf.abundance.celiac.clean <- US.orf.abundance.celiac.clean[prevalence > 0.03, ]
dim(US.orf.abundance.celiac.clean)
# 2573   98

US.metadata.clean.patient <- US.metadata.clean %>% distinct(patientID,Onset_timepoint)

table(US.metadata.clean.patient$Onset_timepoint)
# post_30  pre_30 
#      31       2
table(US.metadata.clean$Onset_timepoint)
# post_30  pre_30 
#     193       4



# check if the column names of abundance table match the order of the row names of metadata table
all(rownames(US.metadata.clean) == colnames(US.orf.abundance.clean))
# TRUE

# check total sparsity
count_matrix <- as.matrix(US.orf.abundance.clean[, sapply(US.orf.abundance.clean, is.numeric)])
sparsity <- sum(count_matrix == 0) / length(count_matrix)
# 0.9518877

```



# 75% mapping rate filtering and 3% prevlance filtering on Italy
```{r}

Italy.orf <- orf %>%
             filter(X1 %like% "_02_GEMM_")
# 132 samples x 46594 orfs


filter_and_save_wide_data(Italy.orf, 
                          covered_fraction_threshold = 0.75, 
                          output_path = "~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/")
# "Number of changes: 53464"


Italy_orf.processed <- readRDS("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/wide_rpkm_genes_75Cov.rds") %>%
                 column_to_rownames("ORFID") 
dim(Italy_orf.processed)
# 23524   132


Italy_sample_all_0 <- names(colSums(Italy_orf.processed)[colSums(Italy_orf.processed) == 0])
# "NovaSeq_N982_I13323_39837_Celiac_Leonard_Stool_02_GEMM_077_12M"       "NovaSeq_N983_I13417_39799_Leonard_Human_stool_Celiac_02_GEMM_006_60M"


Italy.orf.count <- Italy_orf.processed %>%
                   select(-all_of(Italy_sample_all_0)) 


colnames(Italy.orf.count) <- gsub("NovaSeq_N966_|NovaSeq_N978_|NovaSeq_N979_|NovaSeq_N980_|NovaSeq_N981_|NovaSeq_N982_|NovaSeq_N983_|_stats","",colnames(Italy.orf.count))
colnames(Italy.orf.count) <- str_replace_all(colnames(Italy.orf.count),"Leonard_Human_stool_Celiac","Celiac_Leonard_Stool")
colnames(Italy.orf.count) <- sapply(strsplit(colnames(Italy.orf.count),"_Stool_|_Stool_Celiac_"),"[",2)
colnames(Italy.orf.count) <- ifelse(colnames(Italy.orf.count) %like% "_6Y",gsub("_6Y","_72M",colnames(Italy.orf.count)),
                                           ifelse(colnames(Italy.orf.count) %like% "_7Y",gsub("_7Y","_84M",colnames(Italy.orf.count)),colnames(Italy.orf.count)))

Italy.orf.count.table_0.75 <- t(Italy.orf.count) 
# 130 23524


Italy.orf.abundance.table_0.75 <- merge(metadata,Italy.orf.count.table_0.75,by = 0) %>% # 128 samples left
                              filter(feeding_first_year != "Unknown") %>%  # 111 samples left
                              filter(Age.at.Gluten.Introduction..months. != "Unknown") # 111 sample left
# 111 samples x  23549 metadata+orfs


Italy.all_0_contigs <- names(colSums(Italy.orf.abundance.table_0.75[25:23549])[colSums(Italy.orf.abundance.table_0.75[25:23549]) == 0])
# 1076 contigs become 0 through all samples after the 19 samples were removed from the above step
# so we need to remove those contigs


Italy.orf.abundance.table_0.75 <- Italy.orf.abundance.table_0.75 %>%
                            select(-all_of(Italy.all_0_contigs))
# 111 samples x 22473 metadata+orfs
# 22448 orfs


Italy.orf.abundance.table <- Italy.orf.abundance.table_0.75 %>%
                       select(c("Row.names",26:22473)) %>%
                       column_to_rownames("Row.names") %>% 
                       t(.) %>%
                       data.frame(.)
colnames(Italy.orf.abundance.table) <- gsub("X","",colnames(Italy.orf.abundance.table))


# check prevelence filtering
prevalence <- rowSums(Italy.orf.abundance.table > 0) / ncol(Italy.orf.abundance.table)
table(prevalence > 0.03)  # 5% threshold
# FALSE  TRUE 
# 19606  2842 

# Filter ORFs with prevalence > 3%
Italy.orf.abundance.clean <- Italy.orf.abundance.table[prevalence > 0.03, ]
      
dim(Italy.orf.abundance.clean)
# 2842  111

# check if there are any samples become 0 counts
names(colSums(Italy.orf.abundance.clean)[colSums(Italy.orf.abundance.clean) == 0])
# 0

dim(Italy.orf.abundance.clean)
# 2842  111

Italy.metadata.clean <- Italy.orf.abundance.table_0.75[1:25] %>%
                  mutate(Onset_timepoint = ifelse(Onset.Dx < 30,"pre_30","post_30")) %>%
                  column_to_rownames("Row.names")
# 111  25
# 111  25


Italy.metadata.celiac.clean <- Italy.metadata.clean %>% filter(Dx.Status == "CELIAC")
# 48 25

Italy.orf.abundance.celiac.clean <- Italy.orf.abundance.clean %>%
                                 select(rownames(Italy.metadata.celiac.clean))
# check prevelence filtering
prevalence <- rowSums(Italy.orf.abundance.celiac.clean > 0) / ncol(Italy.orf.abundance.celiac.clean)
table(prevalence > 0.03)
# FALSE  TRUE 
#  1487  1355 


Italy.orf.abundance.celiac.clean <- Italy.orf.abundance.celiac.clean[prevalence > 0.03, ]
# 1355   48


Italy.metadata.clean.patient <- Italy.metadata.clean %>% distinct(patientID,Onset_timepoint)

table(Italy.metadata.clean.patient$Onset_timepoint)
# post_30  pre_30 
#      25       8
table(Italy.metadata.clean$Onset_timepoint)
# post_30  pre_30 
#      95      16 

# check if the column names of abundance table match the order of the row names of metadata table
all(rownames(Italy.metadata.clean) == colnames(Italy.orf.abundance.clean))
# TRUE

count_matrix <- as.matrix(Italy.orf.abundance.clean[, sapply(Italy.orf.abundance.clean, is.numeric)])
sparsity <- sum(count_matrix == 0) / length(count_matrix)
# 0.9506502

```




########################################################################## Entire dataset ###############################################################       

Model 1: treat onset_timeline as a numerical variable: create a numeric format of onset_timeline and make it onset_timeline_numeric assume it is linear in the model
~ Dx.Status * onset_timeline_numeric + RunNumber + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode, block = patientID
```{r}



# prepare correct variable format
metadata.clean$feeding_first_year <- factor(metadata.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
table(metadata.clean$feeding_first_year)
# Breast_fed      Formula    Breastmilk_and_formula 
#   119              58                 129 
metadata.clean$HLA.Category <- factor(metadata.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
metadata.clean$Country <- factor(metadata.clean$Country,levels = c("ITALY","USA"))
# ITALY   USA 
#   110   196 
metadata.clean$Sex <- factor(metadata.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   237   69
metadata.clean$Delivery.Mode <- factor(metadata.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 216        90
metadata.clean$Age.at.Gluten.Introduction..months. <- as.numeric(metadata.clean$Age.at.Gluten.Introduction..months.)
 # 4  5   6  7  8   9 10 11 12 15 36 
 # 5  11 57  91 39 33 33 10  2 16  9 
metadata.clean$Dx.Status <- factor(metadata.clean$Dx.Status,levels = c("CELIAC","CONTROL"))
# CONTROL  CELIAC 
#    161     145 



# treat Age.at.Gluten.Introduction..months. as linear
# linear.model.design.runNumber <- model.matrix(~ Dx.Status * onset_timeline_numeric + RunNumber + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,metadata.clean)
# linear.model.fit.runNumber <- voomLmFit(orf.abundance.clean,linear.model.design.runNumber,block= metadata.clean$patientID,plot=FALSE)
# linear.model.fit.runNumber <- eBayes(linear.model.fit.runNumber)


linear.model.design <- model.matrix(~ Dx.Status * onset_timeline_numeric + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,metadata.clean)
linear.model.fit <- voomLmFit(orf.abundance.clean,linear.model.design,block= metadata.clean$patientID,plot=FALSE)
linear.model.fit <- eBayes(linear.model.fit)


# Run voom transformation manually
voom.transformed <- voom(orf.abundance.clean, design = linear.model.design,
                         block = metadata.clean$patientID, plot = TRUE)


# Extract the log2-CPM matrix
voom_logCPM <- voom.transformed$E

top_var_genes <- head(order(apply(voom_logCPM, 1, var), decreasing = TRUE), 500)
voom_top <- voom_logCPM[top_var_genes, ]

p.variability.all <- pheatmap(voom_logCPM,
         scale = "row",                # z-score per gene (optional, but helps)
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         show_colnames = TRUE,         # show sample names
         show_rownames = FALSE)        # hide gene names (if too many)

ggsave(p.variability.all,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/p.variability.all.pdf",dpi=600,width=40,height=30)


p.variability.top500 <- pheatmap(voom_top,
         scale = "row",                # z-score per gene (optional, but helps)
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         show_colnames = TRUE,         # show sample names
         show_rownames = FALSE)        # hide gene names (if too many)

ggsave(p.variability.top500,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/p.variability.top500.pdf",dpi=600,width=40,height=20)


p.pca.total <- prcomp(t(voom_logCPM), scale. = TRUE)

# Plot first two PCs
plot(p.pca.total$x[,1:2],
     main = "PCA of Samples",
     xlab = paste0("PC1 (", round(summary(p.pca.total)$importance[2,1]*100, 1), "%)"),
     ylab = paste0("PC2 (", round(summary(p.pca.total)$importance[2,2]*100, 1), "%)"),
     pch = 19)






## categorical vs numerical

fitted_runNumber <- t(linear.model.design.runNumber %*% t(linear.model.fit.runNumber$coefficients))
fitted_noRunNumber <- t(linear.model.design  %*% t(linear.model.fit$coefficients))


# Step 3: Compute residual sum of squares (RSS) for each gene
resid_runNumber <- US.orf.abundance.clean - fitted_runNumber
resid_noRunNumber  <- US.orf.abundance.clean - fitted_noRunNumber


rss_runNumber <- rowSums(resid_runNumber^2)
rss_noRunNumber     <- rowSums(resid_noRunNumber^2)

# Step 4: Get number of samples and parameters
n <- ncol(US.orf.abundance.clean)
k_cat <- ncol(linear.model.design.runNumber)
k_num <- ncol(linear.model.design)

# Step 5: Calculate AIC
aic_runNumber <- n * log(rss_runNumber / n) + 2 * k_cat
aic_noRunNumber     <- n * log(rss_noRunNumber / n) + 2 * k_num

# Step 6: Compare AIC
delta_aic <- aic_runNumber - aic_noRunNumber  # positive means noRunNumber model better
table(delta_aic > 0)  # how many genes favor noRunNumber model by AIC difference > 2
# FALSE  TRUE 
#  351   3558
table(delta_aic < 0) # how many genes favor runNumber model
# FALSE  TRUE 
#  3558   351
table(abs(delta_aic) <= 2)  # how many are roughly equivalent
# FALSE  TRUE 
#  3837    72

# Optional: Summarize
summary(delta_aic)

# based on the AIC scores, runNumber didn't largely improve the model prediction. So we decide to stay in the model without runNumber


# colnames(linear.model.fit)
#  [1] "(Intercept)"                              "Dx.StatusCONTROL"                         "onset_timeline_numeric"                  
#  [4] "CountryUSA"                               "SexMale"                                  "Age.at.Gluten.Introduction..months."     
#  [7] "HLA.CategoryHigh Risk"                    "HLA.CategoryLow/No Risk"                  "feeding_first_yearFormula"               
# [10] "feeding_first_yearBreastmilk_and_formula" "Delivery.ModeC-Section"                   "Dx.StatusCONTROL:onset_timeline_numeric"

model.res <- topTable(linear.model.fit,coef="Dx.StatusCONTROL:onset_timeline_numeric",number=Inf)
total.model.res <- model.res
sig.model.res <- topTable(linear.model.fit,coef="Dx.StatusCONTROL:onset_timeline_numeric",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 2009


# write.csv(model.res,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_limma_model_res_final.csv",quote = FALSE,row.names = TRUE)
# write.csv(sig.model.res,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_limma_model_Sig_res_final.csv",quote = FALSE,row.names = TRUE)

# Define gene of interest
#selected_gene <- rownames(sig.model.res)

# # Subset design matrix and model coefficients for prediction
# linear.model.design <- model.matrix(~ Dx.Status * onset_timeline_numeric + RunNumber + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,metadata.clean)
# 
# trajectory_plot_list <- list()
# 
# 
# for (i in selected_gene){
#   
#    # Extract observed abundance vector for gene i
#   gene_abundance <- orf.abundance.clean[i, ]
# 
#   # Create metadata + model prediction + observed abundance
#   predicted_df <- metadata.clean %>%
#     mutate(predicted_abundance = as.numeric(linear.model.design %*% linear.model.fit$coefficients[i, ]),
#            observed_abundance = as.numeric(gene_abundance))
# 
#   # Count non-zero samples and total abundance per group and time
#   count_df <- predicted_df %>%
#     group_by(onset_timeline_numeric, Dx.Status) %>%
#     summarise(
#       non_zero_n = sum(observed_abundance > 0, na.rm = TRUE),
#       total_reads = sum(observed_abundance, na.rm = TRUE),
#       .groups = "drop"
#     )
# 
#   # Get mean predicted values per group and time
#   plot_df_pred <- predicted_df %>%
#   left_join(count_df, by = c("onset_timeline_numeric", "Dx.Status")) %>%
#   group_by(onset_timeline_numeric, Dx.Status) %>%
#   summarise(
#     mean_pred = mean(predicted_abundance, na.rm = TRUE),
#     non_zero_n = first(non_zero_n),      # from count_df
#     total_reads = first(total_reads),    # from count_df
#     .groups = "drop"
#   ) %>%
#   mutate(mean_pred = ifelse(non_zero_n == 0 & total_reads == 0, NA, mean_pred))
#   
# 
#   # Label showing both counts
#   plot_df_pred <- plot_df_pred %>%
#     mutate(label_text = paste0("n=", non_zero_n, "\nreads=", round(total_reads)))
# 
#   # Plot with annotations for sample count and total reads
#   p <- ggplot(plot_df_pred, aes(x = onset_timeline_numeric, y = mean_pred, color = Dx.Status)) +
#     geom_line(size = 1.2) +
#     geom_point(size = 2) +
#     geom_text(aes(label = label_text), vjust = -0.5, size = 3) +
#     labs(title = i,
#          x = "Months before onset", y = "Predicted Abundance (logCPM)") +
#     theme(legend.position = "none")
# 
#   # Store in list
#   trajectory_plot_list[[i]] <- p
# }
# 

# # Pick 10 plots
# plot_subset <- trajectory_plot_list[1:500]  # or use names: trajectory_plot_list[c("gene1", "gene2", ...)]
# # Combine into grid: 2 rows × 5 columns
# grid_plot <- wrap_plots(plot_subset)
# ggsave(grid_plot,file=paste0("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Total_trajectory_batchRemoved_clean.pdf"),dpi = 600,width = 75,height=65,limitsize = FALSE)


# hist(sig.model.res$adj.P.Val, breaks = 50) # the shape looks valid
# hist(model.res$adj.P.Val, breaks = 50, col = "gray", main = "All Adjusted P-Values") # the shape looks valid as well

###########################################################################################################################

model.res <- topTable(linear.model.fit,coef="onset_timeline_numeric",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="onset_timeline_numeric",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 13

#################################################################################################

model.res <- topTable(linear.model.fit,coef="Dx.StatusCONTROL",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="Dx.StatusCONTROL",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 2206

#################################################################################################

model.res <- topTable(linear.model.fit,coef="CountryUSA",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="CountryUSA",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 2109
model.res <- topTable(linear.model.fit,coef="SexMale",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="SexMale",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 1541
model.res <- topTable(linear.model.fit,coef="HLA.CategoryHigh Risk",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="HLA.CategoryHigh Risk",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 0 6
model.res <- topTable(linear.model.fit,coef="HLA.CategoryLow/No Risk",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="HLA.CategoryLow/No Risk",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 0 6
model.res <- topTable(linear.model.fit,coef="feeding_first_yearFormula",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="feeding_first_yearFormula",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 0 6
model.res <- topTable(linear.model.fit,coef="feeding_first_yearBreastmilk_and_formula",number=Inf)
sig.model.res <- topTable(linear.model.fit,coef="feeding_first_yearBreastmilk_and_formula",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(sig.model.res)
# 0 6

```



# make volcano plots and coloring by function
```{r}


significant_gene_function_table <- read_delim("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/phold_per_cds_predictions.tsv") %>%
          select(`cds_id`,`phrog`,`function`,`product`) %>%
          column_to_rownames("cds_id") %>%
          merge(model.res,by = 0)


# make volcano plots for the significant genes

# Add significance categories
significant_gene_function_table$significant <- significant_gene_function_table$adj.P.Val < 0.05
significant_gene_function_table$highly_significant <- significant_gene_function_table$adj.P.Val < 0.001
significant_gene_function_table$neg_log10_padj <- -log10(significant_gene_function_table$adj.P.Val)


# FLIP THE SIGN of logFC to switch axis orientation
significant_gene_function_table$logFC_flipped <- -significant_gene_function_table$logFC


# Create significance labels
significant_gene_function_table$significance_label <- "Not Significant"
significant_gene_function_table$significance_label[significant_gene_function_table$significant] <- "Significant (adj.p < 0.05)"
significant_gene_function_table$significance_label[significant_gene_function_table$highly_significant] <- "Highly Significant (adj.p < 0.001)"



significant_gene_function_table$significance_label <- factor(significant_gene_function_table$significance_label,
                                          levels = c("Not Significant",
                                                    "Significant (adj.p < 0.05)",
                                                    "Highly Significant (adj.p < 0.001)"))


# Define colors
colors <- c("Not Significant" = "grey70",
           "Significant (adj.p < 0.05)" = "#E31A1C",
           "Highly Significant (adj.p < 0.001)" = "#B2182B")

# Create volcano plot
total_dataset_volcano_plot <- ggplot(significant_gene_function_table, aes(x = logFC_flipped, y = neg_log10_padj)) +
  geom_point(aes(color = significance_label), size = 0.8, alpha = 0.7) +
  scale_color_manual(values = colors, name = "Significance") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "darkred", alpha = 0.7) +
  labs(
    title = "Viral ORF Differential Abundance: CONTROL vs CELIAC",
    x = "Log2 Fold Change (CONTROL vs CELIAC)",
    y = "-Log10(Adjusted P-value)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )


# Save plot
ggsave(total_dataset_volcano_plot, file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_dataset_volcano_plot.png", width = 8, height = 8, dpi = 300)


######################### color the points based on gene functions ##############################

# Clean up function names and create simplified categories
significant_gene_function_table$`function`[is.na(significant_gene_function_table$`function`)] <- "unknown function"
significant_gene_function_table$`function`[significant_gene_function_table$`function` == ""] <- "unknown function"


# Create simplified functional categories for better visualization
significant_gene_function_table$func_category <- significant_gene_function_table$`function`
significant_gene_function_table$func_category[grepl("DNA|RNA|nucleotide|replication",
significant_gene_function_table$`function`, ignore.case = TRUE)] <- "DNA/RNA metabolism"
significant_gene_function_table$func_category[grepl("head|packaging|capsid",
significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Head & packaging"
significant_gene_function_table$func_category[grepl("tail|fiber", significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Tail"
significant_gene_function_table$func_category[grepl("lysis|holin", significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Lysis"
significant_gene_function_table$func_category[grepl("transcription|regulation",
significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Transcription"
significant_gene_function_table$func_category[grepl("integration|excision|recombination",
significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Integration/excision"
significant_gene_function_table$func_category[grepl("connector", significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Connector"
significant_gene_function_table$func_category[grepl("unknown|hypothetical",significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Unknown function"
significant_gene_function_table$func_category[grepl("other", significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Other"

# Count functions and keep only top categories, group rare ones as 
"Other"
func_counts <- table(significant_gene_function_table$func_category)
top_functions <- names(sort(func_counts, decreasing = TRUE))[1:8]  # Keep top 8
significant_gene_function_table$func_category[!significant_gene_function_table$func_category %in% top_functions] <- "Other/Rare"

# Create color palette for functions
func_colors <- c(
"DNA/RNA metabolism" = "#1f77b4",
"Head & packaging" = "#ff7f0e",
"Tail" = "#2ca02c",
"Unknown function" = "#7f7f7f",
"Lysis" = "#9467bd",
"Transcription" = "#8c564b",
"Integration/excision" = "#e377c2",
"Connector" = "#d62728",
"Other" = "#bcbd22",
"Other/Rare" = "#17becf"
)


significant_gene_function_table.known <- significant_gene_function_table %>% filter(func_category != "Unknown function")

# Create volcano plot colored by function
total_dataset_volcano_plot_color_by_function <- ggplot(significant_gene_function_table.known, aes(x = logFC_flipped, y = neg_log10_padj)) +
                              geom_point(data = significant_gene_function_table.known %>% filter(adj.P.Val < 0.05), aes(color = func_category), size = 0.8, alpha = 0.7) +
                              geom_point(data = significant_gene_function_table.known %>% filter(adj.P.Val >= 0.05),color = "gray",size = 0.8, alpha = 0.7)+ 
                              scale_color_manual(values = func_colors, name = "Gene Function") +
                              geom_hline(yintercept = -log10(0.05), linetype = "dashed", color =
                              "black", alpha = 0.5) +
                              geom_hline(yintercept = -log10(0.001), linetype = "dashed", color =
                              "black", alpha = 0.7) +
                              labs(
                                title = "Viral ORF Differential Abundance by Gene Function (Unknown functions removed)",
                                subtitle = paste("Total:", nrow(significant_gene_function_table.known), "ORFs | Significant:",
                              sum(significant_gene_function_table.known$significant), "ORFs (adj.p < 0.05)"),
                                x = "Log2 Fold Change (CONTROL vs CELIAC)",
                                y = "-Log10(Adjusted P-value)",
                                caption = "Negative logFC = Higher in CONTROL, Lower in CELIAC"
                              ) +
                              theme_minimal() +
                              theme(
                                plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
                                plot.subtitle = element_text(size = 11, hjust = 0.5),
                                legend.position = "right",
                                legend.title = element_text(size = 11, face = "bold"),
                                panel.grid.minor = element_blank()
                              )

ggsave(total_dataset_volcano_plot_color_by_function,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_dataset_volcano_plot_color_by_function_known.png", width = 10, height = 8, dpi = 300)


function.table <- significant_gene_function_table %>% filter(adj.P.Val < 0.05) %>% select(c("Row.names","phrog","function","product"))
write.csv(function.table,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_cohort_function_table.csv",quote = FALSE,row.names = FALSE)


View(significant_gene_function_table %>% group_by(func_category) %>% mutate(number_of_Orfs = n()) %>% distinct(func_category,.keep_all = TRUE) %>% select(func_category,number_of_Orfs))

```



# heatmap was made by claude code:create_temporal_heatmap_from_tables.R
<!-- This script takes your three input files: -->
<!--   - total_limma_model_Sig_res_final.csv (topTable results) -->
<!--   - total_orf.abundance.clean.csv (abundance data) -->
<!--   - total_metadata.clean.csv (metadata) -->

<!-- And generates the two heatmap files: -->
<!--   - all_significant_orfs_temporal_heatmap_from_tables.png -->
<!--   - top_variable_significant_orfs_heatmap_from_tables.png -->

# prepare dataframe for coefficient plot
```{r}

# create the table that combines all the model results
# Extract complete results for ALL coefficients and ALL ORFs

complete_results <- data.frame()


# Get coefficient names
coef_names <- colnames(linear.model.fit$coefficients)
print("Model coefficients:")
print(coef_names)

# Loop through each coefficient to get results
for(i in 1:length(coef_names)) {
  coef_name <- coef_names[i]

  # Get topTable for this coefficient (number = Inf gets ALL ORFs)
  coef_results <- topTable(linear.model.fit, coef = i, number = Inf, sort.by = "none")

  # Add coefficient name prefix to column names (except ORF names)
  colnames(coef_results) <- paste0(coef_name, "_", colnames(coef_results))


  if(i == 1) {
    # First coefficient - initialize the complete results
    complete_results <- coef_results
  } else {
    # Add additional coefficients as new columns
    complete_results <- cbind(complete_results, coef_results)
  }
}

# Add ORF IDs as first column
complete_results <- data.frame(
  orf_id = rownames(complete_results),
  complete_results,
  stringsAsFactors = FALSE
)

colnames(complete_results) <- gsub("X.","",colnames(complete_results))
colnames(complete_results) <- gsub("Dx.StatusCONTROL","Dx.Status(CONTROL vs CELIAC)",colnames(complete_results))
colnames(complete_results) <- gsub("CountryUSA","Country(USA vs Italy)",colnames(complete_results))
colnames(complete_results) <- gsub("SexMale","Sex(Male vs Female)",colnames(complete_results))
colnames(complete_results) <- gsub("HLA.CategoryHigh.Risk","HLA.Category(High.Risk vs Standard Risk)",colnames(complete_results))
colnames(complete_results) <- gsub("HLA.CategoryLow.No.Risk","HLA.Category(Low.No.Risk vs Standard Risk)",colnames(complete_results))
colnames(complete_results) <- gsub("feeding_first_yearFormula","feeding_first_year(Formula vs Breast_fed)",colnames(complete_results))
colnames(complete_results) <- gsub("Delivery.ModeC.Section","Delivery.Mode(C.Section vs Vaginal)",colnames(complete_results))

# Save to file
write.csv(complete_results,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/Total_complete_model_results.csv",quote = FALSE,row.names = TRUE)



```



Model 2: treat onset_timeline as a numerical variable: create a numeric format of onset_timeline and make it onset_timeline_numeric assume it is linear in the model
~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode, block = patientID
```{r}


# prepare correct variable format
metadata.clean$feeding_first_year <- factor(metadata.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
table(metadata.clean$feeding_first_year)
# Breast_fed      Formula    Breastmilk_and_formula 
#   119              58                 129 
metadata.clean$HLA.Category <- factor(metadata.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
metadata.clean$Country <- factor(metadata.clean$Country,levels = c("ITALY","USA"))
# ITALY   USA 
#   110   196 
metadata.clean$Sex <- factor(metadata.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   237   69
metadata.clean$Delivery.Mode <- factor(metadata.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 216        90
metadata.clean$Age.at.Gluten.Introduction..months. <- as.numeric(metadata.clean$Age.at.Gluten.Introduction..months.)
 # 4  5   6  7  8   9 10 11 12 15 36 
 # 5  11 57  91 39 33 33 10  2 16  9 
metadata.clean$Dx.Status <- factor(metadata.clean$Dx.Status,levels = c("CELIAC","CONTROL"))
# CONTROL  CELIAC 
#    161     145 
metadata.clean$onset_timeline_combined <- factor(metadata.clean$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))
 # t0      t0-6     t0-12     t0-18     t0-24     t0-30     t0-36 t0-over42 
 # 64        30        46        27        32        29        33        46 


categorical.model.design <- model.matrix(~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,metadata.clean)
colnames(categorical.model.design) <- make.names(colnames(categorical.model.design))
categorical.model.fit <- voomLmFit(orf.abundance.clean,categorical.model.design,block= metadata.clean$patientID,plot=FALSE)
categorical.model.fit <- eBayes(categorical.model.fit)

colnames(categorical.model.fit)
#  [1] "X.Intercept."                                      "Dx.StatusCONTROL"                                 
#  [3] "onset_timeline_combinedt0.6"                       "onset_timeline_combinedt0.12"                     
#  [5] "onset_timeline_combinedt0.18"                      "onset_timeline_combinedt0.24"                     
#  [7] "onset_timeline_combinedt0.30"                      "onset_timeline_combinedt0.36"                     
#  [9] "onset_timeline_combinedt0.over42"                  "CountryUSA"                                       
# [11] "SexMale"                                           "Age.at.Gluten.Introduction..months."              
# [13] "HLA.CategoryHigh.Risk"                             "HLA.CategoryLow.No.Risk"                          
# [15] "feeding_first_yearFormula"                         "feeding_first_yearBreastmilk_and_formula"         
# [17] "Delivery.ModeC.Section"                            "Dx.StatusCONTROL.onset_timeline_combinedt0.6"     
# [19] "Dx.StatusCONTROL.onset_timeline_combinedt0.12"     "Dx.StatusCONTROL.onset_timeline_combinedt0.18"    
# [21] "Dx.StatusCONTROL.onset_timeline_combinedt0.24"     "Dx.StatusCONTROL.onset_timeline_combinedt0.30"    
# [23] "Dx.StatusCONTROL.onset_timeline_combinedt0.36"     "Dx.StatusCONTROL.onset_timeline_combinedt0.over42"



############# are there any baseline difference in orf abundance between celiac and control group at T0 ##############

categorical.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL",number=Inf)
categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0


############# are there significantly different Orfs between Celiac and Control groups at each time point ###############

contrast_matrix <- makeContrasts(
    "CONTROL_vs_CELIAC_at_t0_6" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.6,
    "CONTROL_vs_CELIAC_at_t0_12" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.12,
    "CONTROL_vs_CELIAC_at_t0_18" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.18,
    "CONTROL_vs_CELIAC_at_t0_24" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.24,
    "CONTROL_vs_CELIAC_at_t0_30" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.30,
    "CONTROL_vs_CELIAC_at_t0_36" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.36,
    "CONTROL_vs_CELIAC_at_t0_over42" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.over42,
    levels = categorical.model.design
  )


# Fit contrasts
contrast_fit <- contrasts.fit(categorical.model.fit, contrast_matrix)
contrast_fit <- eBayes(contrast_fit)

colnames(contrast_fit)
# [1] "CONTROL_vs_CELIAC_at_t0_6"      "CONTROL_vs_CELIAC_at_t0_12"     "CONTROL_vs_CELIAC_at_t0_18"    
# [4] "CONTROL_vs_CELIAC_at_t0_24"     "CONTROL_vs_CELIAC_at_t0_30"     "CONTROL_vs_CELIAC_at_t0_36"    
# [7] "CONTROL_vs_CELIAC_at_t0_over42"

# extract results
results_t0_6 <- topTable(contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_6",number=Inf) %>% filter(adj.P.Val < 0.05)
# 2025 

results_t0_12 <- topTable(contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_12",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

results_t0_18 <- topTable(contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_18",number=Inf) %>% filter(adj.P.Val < 0.05)
# 2050


results_t0_24 <- topTable(contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_24",number=Inf) %>% filter(adj.P.Val < 0.05)
# 1

results_t0_30 <- topTable(contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_30",number=Inf) %>% filter(adj.P.Val < 0.05)
# 1633

results_t0_36 <- topTable(contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_36",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

results_t0_over42 <- topTable(contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_over42",number=Inf) %>% filter(adj.P.Val < 0.05)
# 2



############# are there any orf has significantly changed group difference between each time point and T0 ###############


categorical.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.12",number=Inf)
categorical.total.model.res <- categorical.model.res
categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.12",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.6",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.18",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 1 6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.24",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6


categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.36",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 1565    6


categorical.sig.model.res <- topTable(categorical.model.fit,coef="Dx.StatusCONTROL.onset_timeline_combinedt0.over42",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 1898    6


########### Are there any significantly differential abundant Orfs between each time point and T0 for Celiac group ############

categorical.sig.model.res <- topTable(categorical.model.fit,coef="onset_timeline_combinedt0-6",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="onset_timeline_combinedt0-12",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 34  6


categorical.sig.model.res <- topTable(categorical.model.fit,coef="onset_timeline_combinedt0-18",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="onset_timeline_combinedt0-24",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="onset_timeline_combinedt0-30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 10  6

categorical.sig.model.res <- topTable(categorical.model.fit,coef="onset_timeline_combinedt0-36",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(categorical.sig.model.res)
# 0 6


```



# map orfs to contigs and get the contig annotations
# done by claude

##################################################### US ############################################################### 

Model 1: treat onset_timeline as a numerical variable: create a numeric format of onset_timeline and make it onset_timeline_numeric assume it is linear in the model
~ Dx.Status * onset_timeline_numeric + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode, block = patientID
```{r}

# prepare correct variable format
US.metadata.clean$feeding_first_year <- factor(US.metadata.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
# Breast_fed      Formula   Breastmilk_and_formula 
#    120             58              129 
US.metadata.clean$HLA.Category <- factor(US.metadata.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
US.metadata.clean$Sex <- factor(US.metadata.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   238   69
US.metadata.clean$Delivery.Mode <- factor(US.metadata.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 217        90
US.metadata.clean$Age.at.Gluten.Introduction..months. <- as.numeric(US.metadata.clean$Age.at.Gluten.Introduction..months.)
# #  # 4  5  6  7  8  9 10 11 12 15 36 
# #  # 5 11 57 91 40 33 33 10  2 16  9 
US.metadata.clean$Dx.Status <- factor(US.metadata.clean$Dx.Status,levels = c("CELIAC","CONTROL"))
# CONTROL  CELIAC 
#    162     145 

#write.csv(US.metadata.clean,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/final_results/US.metadata.clean.csv",quote = FALSE,row.names = TRUE)


# treat Age.at.Gluten.Introduction..months. as linear
US.linear.model.design <- model.matrix(~ Dx.Status * onset_timeline_numeric + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,US.metadata.clean)
US.linear.model.fit <- voomLmFit(US.orf.abundance.clean,US.linear.model.design,block= US.metadata.clean$patientID,plot=FALSE)
US.linear.model.fit <- eBayes(US.linear.model.fit)

#write.csv(US.orf.abundance.clean,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/final_results/US.orf.abundance.clean.csv",quote = FALSE,row.names = TRUE)

colnames(US.linear.model.fit)
#  [1] "(Intercept)"                              "Dx.StatusCONTROL"                         "onset_timeline_numeric"                  
#  [4] "SexMale"                                  "Age.at.Gluten.Introduction..months."      "HLA.CategoryHigh Risk"                   
#  [7] "HLA.CategoryLow/No Risk"                  "feeding_first_yearFormula"                "feeding_first_yearBreastmilk_and_formula"
# [10] "Delivery.ModeC-Section"                   "Dx.StatusCONTROL:onset_timeline_numeric" 


US.model.res <- topTable(US.linear.model.fit,coef="Dx.StatusCONTROL:onset_timeline_numeric",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="Dx.StatusCONTROL:onset_timeline_numeric",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 3176 

# write.csv(US.model.res,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/final_results/US.limma_model_res.csv",quote = FALSE,row.names = TRUE)

# # Define gene of interest
# selected_gene <- rownames(US.sig.model.res)
# 
# # Subset design matrix and model coefficients for prediction
# US.linear.model.design <- model.matrix(~ Dx.Status * onset_timeline_numeric + RunNumber + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,US.metadata.clean)
# 
# trajectory_plot_list <- list()
# 
# 
# for (i in selected_gene){
#   
#    # Extract observed abundance vector for gene i
#   gene_abundance <- orf.abundance.clean[i, ]
# 
#   # Create metadata + model prediction + observed abundance
#   predicted_df <- metadata.clean %>%
#     mutate(predicted_abundance = as.numeric(linear.model.design %*% linear.model.fit$coefficients[i, ]),
#            observed_abundance = as.numeric(gene_abundance))
# 
#   # Count non-zero samples and total abundance per group and time
#   count_df <- predicted_df %>%
#     group_by(onset_timeline_numeric, Dx.Status) %>%
#     summarise(
#       non_zero_n = sum(observed_abundance > 0, na.rm = TRUE),
#       total_reads = sum(observed_abundance, na.rm = TRUE),
#       .groups = "drop"
#     )
# 
#   # Get mean predicted values per group and time
#   plot_df_pred <- predicted_df %>%
#   left_join(count_df, by = c("onset_timeline_numeric", "Dx.Status")) %>%
#   group_by(onset_timeline_numeric, Dx.Status) %>%
#   summarise(
#     mean_pred = mean(predicted_abundance, na.rm = TRUE),
#     non_zero_n = first(non_zero_n),      # from count_df
#     total_reads = first(total_reads),    # from count_df
#     .groups = "drop"
#   ) %>%
#   mutate(mean_pred = ifelse(non_zero_n == 0 & total_reads == 0, NA, mean_pred))
#   
# 
#   # Label showing both counts
#   plot_df_pred <- plot_df_pred %>%
#     mutate(label_text = paste0("n=", non_zero_n, "\nreads=", round(total_reads)))
# 
#   # Plot with annotations for sample count and total reads
#   p <- ggplot(plot_df_pred, aes(x = onset_timeline_numeric, y = mean_pred, color = Dx.Status)) +
#     geom_line(size = 1.2) +
#     geom_point(size = 2) +
#     geom_text(aes(label = label_text), vjust = -0.5, size = 3) +
#     labs(title = i,
#          x = "Months before onset", y = "Predicted Abundance (logCPM)") +
#     theme(legend.position = "none")
# 
#   # Store in list
#   trajectory_plot_list[[i]] <- p
# }
# 

# # Pick 10 plots
# plot_subset <- trajectory_plot_list[1:500]  # or use names: trajectory_plot_list[c("gene1", "gene2", ...)]
# # Combine into grid: 2 rows × 5 columns
# grid_plot <- wrap_plots(plot_subset)
# ggsave(grid_plot,file=paste0("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Total_trajectory_batchRemoved_clean.pdf"),dpi = 600,width = 75,height=65,limitsize = FALSE)


# hist(sig.model.res$adj.P.Val, breaks = 50) # the shape looks valid
# hist(model.res$adj.P.Val, breaks = 50, col = "gray", main = "All Adjusted P-Values") # the shape looks valid as well

###########################################################################################################################

US.model.res <- topTable(US.linear.model.fit,coef="onset_timeline_numeric",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="onset_timeline_numeric",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 117

#################################################################################################

US.model.res <- topTable(US.linear.model.fit,coef="Dx.StatusCONTROL",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="Dx.StatusCONTROL",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 3

#################################################################################################

US.model.res <- topTable(US.linear.model.fit,coef="SexMale",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="SexMale",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 1
US.model.res <- topTable(US.linear.model.fit,coef="HLA.CategoryHigh Risk",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="HLA.CategoryHigh Risk",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 0 6
US.model.res <- topTable(US.linear.model.fit,coef="HLA.CategoryLow/No Risk",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="HLA.CategoryLow/No Risk",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 0 6
US.model.res <- topTable(US.linear.model.fit,coef="feeding_first_yearFormula",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="feeding_first_yearFormula",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 0 6
US.model.res <- topTable(US.linear.model.fit,coef="feeding_first_yearBreastmilk_and_formula",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="feeding_first_yearBreastmilk_and_formula",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 0 6
US.model.res <- topTable(US.linear.model.fit,coef="Delivery.ModeC-Section",number=Inf)
US.sig.model.res <- topTable(US.linear.model.fit,coef="Delivery.ModeC-Section",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.sig.model.res)
# 1623    6

```


# make volcano plots and coloring by function
```{r}

View(US.significant_gene_function_table %>% group_by(func_category) %>% mutate(number_of_Orfs = n()) %>% distinct(func_category,.keep_all = TRUE) %>% select(func_category,number_of_Orfs))


US.significant_gene_function_table <- read_delim("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/phold_per_cds_predictions.tsv") %>%
          select(`cds_id`,`phrog`,`function`,`product`) %>%
          column_to_rownames("cds_id") %>%
          merge(US.model.res,by = 0)


# make volcano plots for the significant genes

# Add significance categories
US.significant_gene_function_table$significant <- US.significant_gene_function_table$adj.P.Val < 0.05
US.significant_gene_function_table$highly_significant <- US.significant_gene_function_table$adj.P.Val < 0.001
US.significant_gene_function_table$neg_log10_padj <- -log10(US.significant_gene_function_table$adj.P.Val)


# FLIP THE SIGN of logFC to switch axis orientation
US.significant_gene_function_table$logFC_flipped <- -US.significant_gene_function_table$logFC


# Create significance labels
US.significant_gene_function_table$significance_label <- "Not Significant"
US.significant_gene_function_table$significance_label[US.significant_gene_function_table$significant] <- "Significant (adj.p < 0.05)"
US.significant_gene_function_table$significance_label[US.significant_gene_function_table$highly_significant] <- "Highly Significant (adj.p < 0.001)"



US.significant_gene_function_table$significance_label <- factor(US.significant_gene_function_table$significance_label,
                                          levels = c("Not Significant",
                                                    "Significant (adj.p < 0.05)",
                                                    "Highly Significant (adj.p < 0.001)"))


# Define colors
colors <- c("Not Significant" = "grey70",
           "Significant (adj.p < 0.05)" = "#E31A1C",
           "Highly Significant (adj.p < 0.001)" = "#B2182B")

# Create volcano plot
US.total_dataset_volcano_plot <- ggplot(US.significant_gene_function_table, aes(x = logFC_flipped, y = neg_log10_padj)) +
  geom_point(aes(color = significance_label), size = 0.8, alpha = 0.7) +
  scale_color_manual(values = colors, name = "Significance") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "darkred", alpha = 0.7) +
  labs(
    title = "US Viral ORF Differential Abundance: CONTROL vs CELIAC",
    x = "Log2 Fold Change (CONTROL vs CELIAC)",
    y = "-Log10(Adjusted P-value)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )


# Save plot
ggsave(US.total_dataset_volcano_plot, file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/US.total_dataset_volcano_plot.png", width = 8, height = 8, dpi = 300)


######################### color the points based on gene functions ##############################

# Clean up function names and create simplified categories
US.significant_gene_function_table$`function`[is.na(US.significant_gene_function_table$`function`)] <- "unknown function"
US.significant_gene_function_table$`function`[US.significant_gene_function_table$`function` == ""] <- "unknown function"


# Create simplified functional categories for better visualization
US.significant_gene_function_table$func_category <- US.significant_gene_function_table$`function`
US.significant_gene_function_table$func_category[grepl("DNA|RNA|nucleotide|replication",
US.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "DNA/RNA metabolism"
US.significant_gene_function_table$func_category[grepl("head|packaging|capsid",
US.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Head & packaging"
US.significant_gene_function_table$func_category[grepl("tail|fiber", US.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Tail"
US.significant_gene_function_table$func_category[grepl("lysis|holin", US.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Lysis"
US.significant_gene_function_table$func_category[grepl("transcription|regulation",
US.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Transcription"
US.significant_gene_function_table$func_category[grepl("integration|excision|recombination",
US.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Integration/excision"
US.significant_gene_function_table$func_category[grepl("connector", US.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Connector"
US.significant_gene_function_table$func_category[grepl("unknown|hypothetical",US.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Unknown function"
US.significant_gene_function_table$func_category[grepl("other", US.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Other"

# Count functions and keep only top categories, group rare ones as 
"Other"
func_counts <- table(US.significant_gene_function_table$func_category)
top_functions <- names(sort(func_counts, decreasing = TRUE))[1:8]  # Keep top 8
US.significant_gene_function_table$func_category[!US.significant_gene_function_table$func_category %in% top_functions] <- "Other/Rare"

# Create color palette for functions
func_colors <- c(
"DNA/RNA metabolism" = "#1f77b4",
"Head & packaging" = "#ff7f0e",
"Tail" = "#2ca02c",
"Unknown function" = "#7f7f7f",
"Lysis" = "#9467bd",
"Transcription" = "#8c564b",
"Integration/excision" = "#e377c2",
"Connector" = "#d62728",
"Other" = "#bcbd22",
"Other/Rare" = "#17becf"
)


US.significant_gene_function_table.known <- US.significant_gene_function_table %>% filter(func_category != "Unknown function")


# Create volcano plot colored by function
US.total_dataset_volcano_plot_color_by_function <- ggplot(US.significant_gene_function_table.known, aes(x = logFC_flipped, y = neg_log10_padj)) +
                              geom_point(data = US.significant_gene_function_table.known %>% filter(adj.P.Val < 0.05), aes(color = func_category), size = 0.8, alpha = 0.7) +
                              geom_point(data = US.significant_gene_function_table.known %>% filter(adj.P.Val >= 0.05),color = "gray",size = 0.8, alpha = 0.7)+ 
                              scale_color_manual(values = func_colors, name = "Gene Function") +
                              geom_hline(yintercept = -log10(0.05), linetype = "dashed", color =
                              "black", alpha = 0.5) +
                              geom_hline(yintercept = -log10(0.001), linetype = "dashed", color =
                              "black", alpha = 0.7) +
                              labs(
                                title = "US Viral ORF Differential Abundance by Gene Function (Unknown functions removed)",
                                subtitle = paste("Total:", nrow(US.significant_gene_function_table.known), "ORFs | Significant:",
                              sum(US.significant_gene_function_table.known$significant), "ORFs (adj.p < 0.05)"),
                                x = "Log2 Fold Change (CONTROL vs CELIAC)",
                                y = "-Log10(Adjusted P-value)",
                                caption = "Negative logFC = Higher in CONTROL, Lower in CELIAC"
                              ) +
                              theme_minimal() +
                              theme(
                                plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
                                plot.subtitle = element_text(size = 11, hjust = 0.5),
                                legend.position = "right",
                                legend.title = element_text(size = 11, face = "bold"),
                                panel.grid.minor = element_blank()
                              )

ggsave(US.total_dataset_volcano_plot_color_by_function,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/final_results/US_dataset_volcano_plot_color_by_function_known.png", width = 10, height = 8, dpi = 300)


function.table <- significant_gene_function_table %>% filter(adj.P.Val < 0.05) %>% select(c("Row.names","phrog","function","product"))
write.csv(function.table,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_cohort_function_table.csv",quote = FALSE,row.names = FALSE)


View(significant_gene_function_table %>% group_by(func_category) %>% mutate(number_of_Orfs = n()) %>% distinct(func_category,.keep_all = TRUE) %>% select(func_category,number_of_Orfs))

```


# prepare dataframe for coefficient plot
```{r}

# create the table that combines all the model results
# Extract complete results for ALL coefficients and ALL ORFs
complete_results <- data.frame()

# Get coefficient names
coef_names <- colnames(US.linear.model.fit$coefficients)
print("Model coefficients:")
print(coef_names)

# Loop through each coefficient to get results
for(i in 1:length(coef_names)) {
  coef_name <- coef_names[i]

  # Get topTable for this coefficient (number = Inf gets ALL ORFs)
  coef_results <- topTable(US.linear.model.fit, coef = i, number = Inf, sort.by = "none")

  # Add coefficient name prefix to column names (except ORF names)
  colnames(coef_results) <- paste0(coef_name, "_", colnames(coef_results))

  if(i == 1) {
    # First coefficient - initialize the complete results
    complete_results <- coef_results
  } else {
    # Add additional coefficients as new columns
    complete_results <- cbind(complete_results, coef_results)
  }
}

# Add ORF IDs as first column
complete_results <- data.frame(
  orf_id = rownames(complete_results),
  complete_results,
  stringsAsFactors = FALSE
)

colnames(complete_results) <- gsub("X.","",colnames(complete_results))


colnames(complete_results) <- gsub("X.","",colnames(complete_results))
colnames(complete_results) <- gsub("Dx.StatusCONTROL","Dx.Status(CONTROL vs CELIAC)",colnames(complete_results))
colnames(complete_results) <- gsub("CountryUSA","Country(USA vs Italy)",colnames(complete_results))
colnames(complete_results) <- gsub("SexMale","Sex(Male vs Female)",colnames(complete_results))
colnames(complete_results) <- gsub("HLA.CategoryHigh.Risk","HLA.Category(High.Risk vs Standard Risk)",colnames(complete_results))
colnames(complete_results) <- gsub("HLA.CategoryLow.No.Risk","HLA.Category(Low.No.Risk vs Standard Risk)",colnames(complete_results))
colnames(complete_results) <- gsub("feeding_first_yearFormula","feeding_first_year(Formula vs Breast_fed)",colnames(complete_results))
colnames(complete_results) <- gsub("Delivery.ModeC.Section","Delivery.Mode(C.Section vs Vaginal)",colnames(complete_results))


# Save to file
write.csv(complete_results,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/final_results/US_complete_model_results.csv",quote = FALSE,row.names = TRUE)

```



Model 2: treat onset_timeline as a numerical variable: create a numeric format of onset_timeline and make it onset_timeline_combine assume it is linear in the model
~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode, block = patientID
```{r}


# prepare correct variable format
US.metadata.clean$feeding_first_year <- factor(US.metadata.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
# Breast_fed      Formula   Breastmilk_and_formula 
#    120             58              129 
US.metadata.clean$HLA.Category <- factor(US.metadata.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
US.metadata.clean$Sex <- factor(US.metadata.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   238   69
US.metadata.clean$Delivery.Mode <- factor(US.metadata.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 217        90
US.metadata.clean$Age.at.Gluten.Introduction..months. <- as.numeric(US.metadata.clean$Age.at.Gluten.Introduction..months.)
# #  # 4  5  6  7  8  9 10 11 12 15 36 
# #  # 5 11 57 91 40 33 33 10  2 16  9 
US.metadata.clean$Dx.Status <- factor(US.metadata.clean$Dx.Status,levels = c("CELIAC","CONTROL"))
# CONTROL  CELIAC 
#    162     145  
US.metadata.clean$onset_timeline_combined <- factor(US.metadata.clean$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))
 # t0      t0-6     t0-12     t0-18     t0-24     t0-30     t0-36 t0-over42 
 # 64        30        46        27        32        29        33        46 


US.categorical.model.design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,US.metadata.clean)
colnames(US.categorical.model.design) <- make.names(colnames(US.categorical.model.design))
US.categorical.model.fit <- voomLmFit(US.orf.abundance.clean,US.categorical.model.design,block= US.metadata.clean$patientID,plot=FALSE)
US.categorical.model.fit <- eBayes(US.categorical.model.fit)

colnames(US.categorical.model.fit)
#  [1] "X.Intercept."                                      "Dx.StatusCONTROL"                                 
#  [3] "onset_timeline_combinedt0.6"                       "onset_timeline_combinedt0.12"                     
#  [5] "onset_timeline_combinedt0.18"                      "onset_timeline_combinedt0.24"                     
#  [7] "onset_timeline_combinedt0.30"                      "onset_timeline_combinedt0.36"                     
#  [9] "onset_timeline_combinedt0.over42"                  "SexMale"                                          
# [11] "Age.at.Gluten.Introduction..months."               "HLA.CategoryHigh.Risk"                            
# [13] "HLA.CategoryLow.No.Risk"                           "feeding_first_yearFormula"                        
# [15] "feeding_first_yearBreastmilk_and_formula"          "Delivery.ModeC.Section"                           
# [17] "Dx.StatusCONTROL.onset_timeline_combinedt0.6"      "Dx.StatusCONTROL.onset_timeline_combinedt0.12"    
# [19] "Dx.StatusCONTROL.onset_timeline_combinedt0.18"     "Dx.StatusCONTROL.onset_timeline_combinedt0.24"    
# [21] "Dx.StatusCONTROL.onset_timeline_combinedt0.30"     "Dx.StatusCONTROL.onset_timeline_combinedt0.36"    
# [23] "Dx.StatusCONTROL.onset_timeline_combinedt0.over42"


############# are there any baseline difference in orf abundance between celiac and control group at T0 ##############

US.categorical.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL",number=Inf)
US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0


############# are there significantly different Orfs between Celiac and Control groups at each time point ###############

US.contrast_matrix <- makeContrasts(
    "CONTROL_vs_CELIAC_at_t0_6" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.6,
    "CONTROL_vs_CELIAC_at_t0_12" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.12,
    "CONTROL_vs_CELIAC_at_t0_18" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.18,
    "CONTROL_vs_CELIAC_at_t0_24" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.24,
    "CONTROL_vs_CELIAC_at_t0_30" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.30,
    "CONTROL_vs_CELIAC_at_t0_36" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.36,
    "CONTROL_vs_CELIAC_at_t0_over42" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.over42,
    levels = US.categorical.model.design
  )


# Fit contrasts
US.contrast_fit <- contrasts.fit(US.categorical.model.fit, US.contrast_matrix)
US.contrast_fit <- eBayes(US.contrast_fit)

colnames(US.contrast_fit)
# [1] "CONTROL_vs_CELIAC_at_t0_6"      "CONTROL_vs_CELIAC_at_t0_12"     "CONTROL_vs_CELIAC_at_t0_18"    
# [4] "CONTROL_vs_CELIAC_at_t0_24"     "CONTROL_vs_CELIAC_at_t0_30"     "CONTROL_vs_CELIAC_at_t0_36"    
# [7] "CONTROL_vs_CELIAC_at_t0_over42"

# extract results
US.results_t0_6 <- topTable(US.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_6",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

US.results_t0_12 <- topTable(US.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_12",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

US.results_t0_18 <- topTable(US.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_18",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

US.results_t0_24 <- topTable(US.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_24",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0


US.results_t0_30 <- topTable(US.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_30",number=Inf) %>% filter(adj.P.Val < 0.05)
# 23
rownames(US.results_t0_30)

US.results_t0_36 <- topTable(US.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_36",number=Inf) %>% filter(adj.P.Val < 0.05)
# 18

US.results_t0_over42 <- topTable(US.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_over42",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0



############# are there any orf has significantly changed group difference between each time point and T0 ###############


US.categorical.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-12",number=Inf)
US.categorical.total.model.res <- US.categorical.model.res
US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-12",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-6",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-18",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-24",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 610 6
View(US.categorical.sig.model.res)

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6


US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-36",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 2623    6


US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-over42",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 2270    6


###############################################

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-6",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 3374 6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-12",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 3493  6


US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-18",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 2869 6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-24",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0  6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-36",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6

```


check the t0-24, t0-30, and t0-36 abundance
```{r}


US.metadata.clean.24.30.36 <- US.metadata.clean %>%
                               filter(onset_timeline_combined %in% c("t0-24","t0-30","t0-36"))
US.24.30.36.id <- rownames(US.metadata.clean.24.30.36)
dim(US.metadata.clean.24.30.36)
# 73 25


# extract the count
US.orf.abundance.clean.24.30.36 <- US.orf.abundance.clean %>%
                                   select(all_of(US.24.30.36.id))
dim(US.orf.abundance.clean.24.30.36)
# 3909   73


US.meta.abundance.24.30.36 <- merge(US.metadata.clean.24.30.36,t(US.orf.abundance.clean.24.30.36),by = 0)


US.meta.abundance.24 <- US.meta.abundance.24.30.36 %>%
                        filter(onset_timeline_combined == "t0-24") 

US.abundance.24 <- US.meta.abundance.24[]




US.meta.abundance.30 <- US.meta.abundance.24.30.36 %>%
                        filter(onset_timeline_combined == "t0-30")




US.meta.abundance.36 <- US.meta.abundance.24.30.36 %>%
                        filter(onset_timeline_combined == "t0-36")






```



# pre-30 vs post-30
```{r}

# prepare correct variable format
US.metadata.celiac.clean$feeding_first_year <- factor(US.metadata.celiac.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
# Breast_fed      Formula   Breastmilk_and_formula 
#    120             58              129 
US.metadata.celiac.clean$HLA.Category <- factor(US.metadata.celiac.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
US.metadata.celiac.clean$Sex <- factor(US.metadata.celiac.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   238   69
US.metadata.celiac.clean$Delivery.Mode <- factor(US.metadata.celiac.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 217        90
US.metadata.celiac.clean$Age.at.Gluten.Introduction..months. <- as.numeric(US.metadata.celiac.clean$Age.at.Gluten.Introduction..months.)
# #  # 4  5  6  7  8  9 10 11 12 15 36 
# #  # 5 11 57 91 40 33 33 10  2 16  9 
US.metadata.celiac.clean$Onset_timepoint <- factor(US.metadata.celiac.clean$Onset_timepoint,levels = c("pre_30","post_30"))
# pre_30 post_30


US.Celiac.categorical.model.design <- model.matrix(~ Onset_timepoint + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year,US.metadata.celiac.clean)
US.Celiac.categorical.model.fit <- voomLmFit(US.orf.abundance.celiac.clean,US.Celiac.categorical.model.design,block= US.metadata.celiac.clean$patientID,plot=FALSE)
US.Celiac.categorical.model.fit <- eBayes(US.Celiac.categorical.model.fit)

colnames(US.Celiac.categorical.model.fit)
# [1] "(Intercept)"                              "Onset_timepointpost_30"                   "SexMale"                                 
# [4] "Age.at.Gluten.Introduction..months."      "HLA.CategoryHigh Risk"                    "HLA.CategoryLow/No Risk"                 
# [7] "feeding_first_yearFormula"                "feeding_first_yearBreastmilk_and_formula"


US.Celiac.categorical.sig.model.res <- topTable(US.Celiac.categorical.model.fit,coef="Onset_timepointpost_30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.Celiac.categorical.sig.model.res %>% filter(logFC > 0))
# 2290    6


View(US.Celiac.categorical.sig.model.res$logFC)


```



```{r}

Italy.significant_gene_function_table <- read_delim("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/phold_per_cds_predictions.tsv") %>%
          select(`cds_id`,`phrog`,`function`,`product`) %>%
          column_to_rownames("cds_id") %>%
          merge(Italy.sig.model.res,by = 0)


# make volcano plots for the significant genes

# Add significance categories
Italy.significant_gene_function_table$significant <- Italy.significant_gene_function_table$adj.P.Val < 0.05
Italy.significant_gene_function_table$highly_significant <- Italy.significant_gene_function_table$adj.P.Val < 0.001
Italy.significant_gene_function_table$neg_log10_padj <- -log10(Italy.significant_gene_function_table$adj.P.Val)


# FLIP THE SIGN of logFC to switch axis orientation
Italy.significant_gene_function_table$logFC_flipped <- -Italy.significant_gene_function_table$logFC


# Create significance labels
Italy.significant_gene_function_table$significance_label <- "Not Significant"
Italy.significant_gene_function_table$significance_label[Italy.significant_gene_function_table$significant] <- "Significant (adj.p < 0.05)"
Italy.significant_gene_function_table$significance_label[Italy.significant_gene_function_table$highly_significant] <- "Highly Significant (adj.p < 0.001)"



Italy.significant_gene_function_table$significance_label <- factor(Italy.significant_gene_function_table$significance_label,
                                          levels = c("Not Significant",
                                                    "Significant (adj.p < 0.05)",
                                                    "Highly Significant (adj.p < 0.001)"))


# Define colors
colors <- c("Not Significant" = "grey70",
           "Significant (adj.p < 0.05)" = "#E31A1C",
           "Highly Significant (adj.p < 0.001)" = "#B2182B")

# Create volcano plot
Italy.total_dataset_volcano_plot <- ggplot(Italy.significant_gene_function_table, aes(x = logFC_flipped, y = neg_log10_padj)) +
  geom_point(aes(color = significance_label), size = 0.8, alpha = 0.7) +
  scale_color_manual(values = colors, name = "Significance") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "darkred", alpha = 0.7) +
  labs(
    title = "US Viral ORF Differential Abundance: CONTROL vs CELIAC",
    x = "Log2 Fold Change (CONTROL vs CELIAC)",
    y = "-Log10(Adjusted P-value)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )


# Save plot
ggsave(Italy.total_dataset_volcano_plot, file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy.total_dataset_volcano_plot.png", width = 8, height = 8, dpi = 300)


######################### color the points based on gene functions ##############################

# Clean up function names and create simplified categories
Italy.significant_gene_function_table$`function`[is.na(Italy.significant_gene_function_table$`function`)] <- "unknown function"
Italy.significant_gene_function_table$`function`[Italy.significant_gene_function_table$`function` == ""] <- "unknown function"


# Create simplified functional categories for better visualization
Italy.significant_gene_function_table$func_category <- Italy.significant_gene_function_table$`function`
Italy.significant_gene_function_table$func_category[grepl("DNA|RNA|nucleotide|replication",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "DNA/RNA metabolism"
Italy.significant_gene_function_table$func_category[grepl("head|packaging|capsid",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Head & packaging"
Italy.significant_gene_function_table$func_category[grepl("tail|fiber", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Tail"
Italy.significant_gene_function_table$func_category[grepl("lysis|holin", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Lysis"
Italy.significant_gene_function_table$func_category[grepl("transcription|regulation",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Transcription"
Italy.significant_gene_function_table$func_category[grepl("integration|excision|recombination",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Integration/excision"
Italy.significant_gene_function_table$func_category[grepl("connector", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Connector"
Italy.significant_gene_function_table$func_category[grepl("unknown|hypothetical",Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Unknown function"
Italy.significant_gene_function_table$func_category[grepl("other", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Other"

# Count functions and keep only top categories, group rare ones as 
"Other"
func_counts <- table(Italy.significant_gene_function_table$func_category)
top_functions <- names(sort(func_counts, decreasing = TRUE))[1:8]  # Keep top 8
Italy.significant_gene_function_table$func_category[!Italy.significant_gene_function_table$func_category %in% top_functions] <- "Other/Rare"

# Create color palette for functions
func_colors <- c(
"DNA/RNA metabolism" = "#1f77b4",
"Head & packaging" = "#ff7f0e",
"Tail" = "#2ca02c",
"Unknown function" = "#7f7f7f",
"Lysis" = "#9467bd",
"Transcription" = "#8c564b",
"Integration/excision" = "#e377c2",
"Connector" = "#d62728",
"Other" = "#bcbd22",
"Other/Rare" = "#17becf"
)


Italy.significant_gene_function_table.known <- Italy.significant_gene_function_table %>% filter(func_category != "Unknown function")
# 0

# Create volcano plot colored by function
Italy.total_dataset_volcano_plot_color_by_function <- ggplot(Italy.significant_gene_function_table.known, aes(x = logFC_flipped, y = neg_log10_padj)) +
                              geom_point(data = Italy.significant_gene_function_table.known %>% filter(adj.P.Val < 0.05), aes(color = func_category), size = 0.8, alpha = 0.7) +
                              geom_point(data = Italy.significant_gene_function_table.known %>% filter(adj.P.Val >= 0.05),color = "gray",size = 0.8, alpha = 0.7)+ 
                              scale_color_manual(values = func_colors, name = "Gene Function") +
                              geom_hline(yintercept = -log10(0.05), linetype = "dashed", color =
                              "black", alpha = 0.5) +
                              geom_hline(yintercept = -log10(0.001), linetype = "dashed", color =
                              "black", alpha = 0.7) +
                              labs(
                                title = "US Viral ORF Differential Abundance by Gene Function (Unknown functions removed)",
                                subtitle = paste("Italy:", nrow(Italy.significant_gene_function_table.known), "ORFs | Significant:",
                              sum(Italy.significant_gene_function_table.known$significant), "ORFs (adj.p < 0.05)"),
                                x = "Log2 Fold Change (CONTROL vs CELIAC)",
                                y = "-Log10(Adjusted P-value)",
                                caption = "Negative logFC = Higher in CONTROL, Lower in CELIAC"
                              ) +
                              theme_minimal() +
                              theme(
                                plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
                                plot.subtitle = element_text(size = 11, hjust = 0.5),
                                legend.position = "right",
                                legend.title = element_text(size = 11, face = "bold"),
                                panel.grid.minor = element_blank()
                              )

ggsave(Italy.total_dataset_volcano_plot_color_by_function,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy_dataset_volcano_plot_color_by_function_known.png", width = 10, height = 8, dpi = 300)


function.table <- Italy.significant_gene_function_table %>% filter(adj.P.Val < 0.05) %>% select(c("Row.names","phrog","function","product"))
write.csv(function.table,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy_cohort_function_table.csv",quote = FALSE,row.names = FALSE)


View(Italy.significant_gene_function_table %>% group_by(func_category) %>% mutate(number_of_Orfs = n()) %>% distinct(func_category,.keep_all = TRUE) %>% select(func_category,number_of_Orfs))





```


########################################################################## Italy ############################################################### 

Model 1: treat onset_timeline as a numerical variable: create a numeric format of onset_timeline and make it onset_timeline_numeric assume it is linear in the model
~ Dx.Status * onset_timeline_numeric + RunNumber + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode, block = patientID
```{r}

# prepare correct variable format
Italy.metadata.clean$feeding_first_year <- factor(Italy.metadata.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
# Breast_fed      Formula   Breastmilk_and_formula 
#    120             58              129 
Italy.metadata.clean$HLA.Category <- factor(Italy.metadata.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
Italy.metadata.clean$Sex <- factor(Italy.metadata.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   238   69
Italy.metadata.clean$Delivery.Mode <- factor(Italy.metadata.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 217        90
Italy.metadata.clean$Age.at.Gluten.Introduction..months. <- as.numeric(Italy.metadata.clean$Age.at.Gluten.Introduction..months.)
# 4  5  6  7  8  9 10 11 12 15 36 
# 5 11 57 91 40 33 33 10  2 16  9 
Italy.metadata.clean$Dx.Status <- factor(Italy.metadata.clean$Dx.Status,levels = c("CELIAC","CONTROL"))
# CONTROL  CELIAC 
#    162     145 

# write.csv(Italy.metadata.clean,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy.metadata.clean.csv",quote = FALSE,row.names = TRUE)
# 
# 
# write.csv(Italy.orf.abundance.clean,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy.orf.abundance.clean.csv",quote = FALSE,row.names = TRUE)



# treat Age.at.Gluten.Introduction..months. as linear
Italy.linear.model.design <- model.matrix(~ Dx.Status * onset_timeline_numeric + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,Italy.metadata.clean)
Italy.linear.model.fit <- voomLmFit(Italy.orf.abundance.clean,Italy.linear.model.design,block= Italy.metadata.clean$patientID,plot=FALSE)
Italy.linear.model.fit <- eBayes(Italy.linear.model.fit)

colnames(Italy.linear.model.fit)
#  [1] "(Intercept)"                              "Dx.StatusCONTROL"                         "onset_timeline_numeric"                  
#  [4] "SexMale"                                  "Age.at.Gluten.Introduction..months."      "HLA.CategoryHigh Risk"                   
#  [7] "HLA.CategoryLow/No Risk"                  "feeding_first_yearFormula"                "feeding_first_yearBreastmilk_and_formula"
# [10] "Delivery.ModeC-Section"                   "Dx.StatusCONTROL:onset_timeline_numeric" 


Italy.model.res <- topTable(Italy.linear.model.fit,coef="Dx.StatusCONTROL:onset_timeline_numeric",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="Dx.StatusCONTROL:onset_timeline_numeric",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 2


# write.csv(Italy.model.res,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy.limma_model_res.csv",quote = FALSE,row.names = TRUE)

#################################################################################################

Italy.model.res <- topTable(Italy.linear.model.fit,coef="onset_timeline_numeric",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="onset_timeline_numeric",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 43

#################################################################################################

Italy.model.res <- topTable(Italy.linear.model.fit,coef="Dx.StatusCONTROL",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="Dx.StatusCONTROL",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 1933

#################################################################################################

Italy.model.res <- topTable(Italy.linear.model.fit,coef="SexMale",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="SexMale",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 0
Italy.model.res <- topTable(Italy.linear.model.fit,coef="HLA.CategoryHigh Risk",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="HLA.CategoryHigh Risk",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 0 6
Italy.model.res <- topTable(Italy.linear.model.fit,coef="HLA.CategoryLow/No Risk",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="HLA.CategoryLow/No Risk",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 0 6
Italy.model.res <- topTable(Italy.linear.model.fit,coef="feeding_first_yearFormula",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="feeding_first_yearFormula",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 0 6
Italy.model.res <- topTable(Italy.linear.model.fit,coef="feeding_first_yearBreastmilk_and_formula",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="feeding_first_yearBreastmilk_and_formula",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 0 6
Italy.model.res <- topTable(Italy.linear.model.fit,coef="Delivery.ModeC-Section",number=Inf)
Italy.sig.model.res <- topTable(Italy.linear.model.fit,coef="Delivery.ModeC-Section",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.sig.model.res)
# 1609  6

```


# make volcano plots and coloring by function
```{r}

Italy.significant_gene_function_table <- read_delim("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/phold_per_cds_predictions.tsv") %>%
          select(`cds_id`,`phrog`,`function`,`product`) %>%
          column_to_rownames("cds_id") %>%
          merge(Italy.sig.model.res,by = 0)


# make volcano plots for the significant genes

# Add significance categories
Italy.significant_gene_function_table$significant <- Italy.significant_gene_function_table$adj.P.Val < 0.05
Italy.significant_gene_function_table$highly_significant <- Italy.significant_gene_function_table$adj.P.Val < 0.001
Italy.significant_gene_function_table$neg_log10_padj <- -log10(Italy.significant_gene_function_table$adj.P.Val)


# FLIP THE SIGN of logFC to switch axis orientation
Italy.significant_gene_function_table$logFC_flipped <- -Italy.significant_gene_function_table$logFC


# Create significance labels
Italy.significant_gene_function_table$significance_label <- "Not Significant"
Italy.significant_gene_function_table$significance_label[Italy.significant_gene_function_table$significant] <- "Significant (adj.p < 0.05)"
Italy.significant_gene_function_table$significance_label[Italy.significant_gene_function_table$highly_significant] <- "Highly Significant (adj.p < 0.001)"



Italy.significant_gene_function_table$significance_label <- factor(Italy.significant_gene_function_table$significance_label,
                                          levels = c("Not Significant",
                                                    "Significant (adj.p < 0.05)",
                                                    "Highly Significant (adj.p < 0.001)"))


# Define colors
colors <- c("Not Significant" = "grey70",
           "Significant (adj.p < 0.05)" = "#E31A1C",
           "Highly Significant (adj.p < 0.001)" = "#B2182B")

# Create volcano plot
Italy.total_dataset_volcano_plot <- ggplot(Italy.significant_gene_function_table, aes(x = logFC_flipped, y = neg_log10_padj)) +
  geom_point(aes(color = significance_label), size = 0.8, alpha = 0.7) +
  scale_color_manual(values = colors, name = "Significance") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "darkred", alpha = 0.7) +
  labs(
    title = "US Viral ORF Differential Abundance: CONTROL vs CELIAC",
    x = "Log2 Fold Change (CONTROL vs CELIAC)",
    y = "-Log10(Adjusted P-value)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )


# Save plot
ggsave(Italy.total_dataset_volcano_plot, file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy.total_dataset_volcano_plot.png", width = 8, height = 8, dpi = 300)


######################### color the points based on gene functions ##############################

# Clean up function names and create simplified categories
Italy.significant_gene_function_table$`function`[is.na(Italy.significant_gene_function_table$`function`)] <- "unknown function"
Italy.significant_gene_function_table$`function`[Italy.significant_gene_function_table$`function` == ""] <- "unknown function"


# Create simplified functional categories for better visualization
Italy.significant_gene_function_table$func_category <- Italy.significant_gene_function_table$`function`
Italy.significant_gene_function_table$func_category[grepl("DNA|RNA|nucleotide|replication",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "DNA/RNA metabolism"
Italy.significant_gene_function_table$func_category[grepl("head|packaging|capsid",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Head & packaging"
Italy.significant_gene_function_table$func_category[grepl("tail|fiber", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Tail"
Italy.significant_gene_function_table$func_category[grepl("lysis|holin", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Lysis"
Italy.significant_gene_function_table$func_category[grepl("transcription|regulation",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Transcription"
Italy.significant_gene_function_table$func_category[grepl("integration|excision|recombination",
Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Integration/excision"
Italy.significant_gene_function_table$func_category[grepl("connector", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Connector"
Italy.significant_gene_function_table$func_category[grepl("unknown|hypothetical",Italy.significant_gene_function_table$`function`, ignore.case = TRUE)] <- "Unknown function"
Italy.significant_gene_function_table$func_category[grepl("other", Italy.significant_gene_function_table$`function`,ignore.case = TRUE)] <- "Other"

# Count functions and keep only top categories, group rare ones as 
"Other"
func_counts <- table(Italy.significant_gene_function_table$func_category)
top_functions <- names(sort(func_counts, decreasing = TRUE))[1:8]  # Keep top 8
Italy.significant_gene_function_table$func_category[!Italy.significant_gene_function_table$func_category %in% top_functions] <- "Other/Rare"

# Create color palette for functions
func_colors <- c(
"DNA/RNA metabolism" = "#1f77b4",
"Head & packaging" = "#ff7f0e",
"Tail" = "#2ca02c",
"Unknown function" = "#7f7f7f",
"Lysis" = "#9467bd",
"Transcription" = "#8c564b",
"Integration/excision" = "#e377c2",
"Connector" = "#d62728",
"Other" = "#bcbd22",
"Other/Rare" = "#17becf"
)


Italy.significant_gene_function_table.known <- Italy.significant_gene_function_table %>% filter(func_category != "Unknown function")
# 0

# Create volcano plot colored by function
Italy.total_dataset_volcano_plot_color_by_function <- ggplot(Italy.significant_gene_function_table.known, aes(x = logFC_flipped, y = neg_log10_padj)) +
                              geom_point(data = Italy.significant_gene_function_table.known %>% filter(adj.P.Val < 0.05), aes(color = func_category), size = 0.8, alpha = 0.7) +
                              geom_point(data = Italy.significant_gene_function_table.known %>% filter(adj.P.Val >= 0.05),color = "gray",size = 0.8, alpha = 0.7)+ 
                              scale_color_manual(values = func_colors, name = "Gene Function") +
                              geom_hline(yintercept = -log10(0.05), linetype = "dashed", color =
                              "black", alpha = 0.5) +
                              geom_hline(yintercept = -log10(0.001), linetype = "dashed", color =
                              "black", alpha = 0.7) +
                              labs(
                                title = "US Viral ORF Differential Abundance by Gene Function (Unknown functions removed)",
                                subtitle = paste("Italy:", nrow(Italy.significant_gene_function_table.known), "ORFs | Significant:",
                              sum(Italy.significant_gene_function_table.known$significant), "ORFs (adj.p < 0.05)"),
                                x = "Log2 Fold Change (CONTROL vs CELIAC)",
                                y = "-Log10(Adjusted P-value)",
                                caption = "Negative logFC = Higher in CONTROL, Lower in CELIAC"
                              ) +
                              theme_minimal() +
                              theme(
                                plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
                                plot.subtitle = element_text(size = 11, hjust = 0.5),
                                legend.position = "right",
                                legend.title = element_text(size = 11, face = "bold"),
                                panel.grid.minor = element_blank()
                              )

ggsave(Italy.total_dataset_volcano_plot_color_by_function,file="~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy_dataset_volcano_plot_color_by_function_known.png", width = 10, height = 8, dpi = 300)


function.table <- Italy.significant_gene_function_table %>% filter(adj.P.Val < 0.05) %>% select(c("Row.names","phrog","function","product"))
write.csv(function.table,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy_cohort_function_table.csv",quote = FALSE,row.names = FALSE)


View(Italy.significant_gene_function_table %>% group_by(func_category) %>% mutate(number_of_Orfs = n()) %>% distinct(func_category,.keep_all = TRUE) %>% select(func_category,number_of_Orfs))

```


# prepare dataframe for coefficient plot
```{r}

# create the table that combines all the model results
# Extract complete results for ALL coefficients and ALL ORFs

complete_results <- data.frame()

# Get coefficient names
coef_names <- colnames(Italy.linear.model.fit$coefficients)
print("Model coefficients:")
print(coef_names)

# Loop through each coefficient to get results
for(i in 1:length(coef_names)) {
  coef_name <- coef_names[i]

  # Get topTable for this coefficient (number = Inf gets ALL ORFs)
  coef_results <- topTable(Italy.linear.model.fit, coef = i, number = Inf, sort.by = "none")

  # Add coefficient name prefix to column names (except ORF names)
  colnames(coef_results) <- paste0(coef_name, "_", colnames(coef_results))

  if(i == 1) {
    # First coefficient - initialize the complete results
    complete_results <- coef_results
  } else {
    # Add additional coefficients as new columns
    complete_results <- cbind(complete_results, coef_results)
  }
}

# Add ORF IDs as first column
complete_results <- data.frame(
  orf_id = rownames(complete_results),
  complete_results,
  stringsAsFactors = FALSE
)

colnames(complete_results) <- gsub("X.","",colnames(complete_results))


colnames(complete_results) <- gsub("X.","",colnames(complete_results))
colnames(complete_results) <- gsub("Dx.StatusCONTROL","Dx.Status(CONTROL vs CELIAC)",colnames(complete_results))
colnames(complete_results) <- gsub("CountryUSA","Country(USA vs Italy)",colnames(complete_results))
colnames(complete_results) <- gsub("SexMale","Sex(Male vs Female)",colnames(complete_results))
colnames(complete_results) <- gsub("HLA.CategoryHigh.Risk","HLA.Category(High.Risk vs Standard Risk)",colnames(complete_results))
colnames(complete_results) <- gsub("HLA.CategoryLow.No.Risk","HLA.Category(Low.No.Risk vs Standard Risk)",colnames(complete_results))
colnames(complete_results) <- gsub("feeding_first_yearFormula","feeding_first_year(Formula vs Breast_fed)",colnames(complete_results))
colnames(complete_results) <- gsub("Delivery.ModeC.Section","Delivery.Mode(C.Section vs Vaginal)",colnames(complete_results))


# Save to file
write.csv(complete_results,"~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/Italy_complete_model_results.csv",quote = FALSE,row.names = TRUE)

```


Model 2: treat onset_timeline as a numerical variable: create a numeric format of onset_timeline and make it onset_timeline_combine assume it is linear in the model
~ Dx.Status * onset_timeline_combined + Country + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode, block = patientID
```{r}


# prepare correct variable format
Italy.metadata.clean$feeding_first_year <- factor(Italy.metadata.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
# Breast_fed      Formula   Breastmilk_and_formula 
#    120             58              129 
Italy.metadata.clean$HLA.Category <- factor(Italy.metadata.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
Italy.metadata.clean$Sex <- factor(Italy.metadata.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   238   69
Italy.metadata.clean$Delivery.Mode <- factor(Italy.metadata.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 217        90
Italy.metadata.clean$Age.at.Gluten.Introduction..months. <- as.numeric(Italy.metadata.clean$Age.at.Gluten.Introduction..months.)
# #  # 4  5  6  7  8  9 10 11 12 15 36 
# #  # 5 11 57 91 40 33 33 10  2 16  9 
Italy.metadata.clean$Dx.Status <- factor(Italy.metadata.clean$Dx.Status,levels = c("CELIAC","CONTROL"))
# CONTROL  CELIAC 
#    162     145  
Italy.metadata.clean$onset_timeline_combined <- factor(Italy.metadata.clean$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))
 # t0      t0-6     t0-12     t0-18     t0-24     t0-30     t0-36 t0-over42 
 # 64        30        46        27        32        29        33        46 


Italy.categorical.model.design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + HLA.Category + feeding_first_year + Delivery.Mode,Italy.metadata.clean)
colnames(Italy.categorical.model.design) <- make.names(colnames(Italy.categorical.model.design))
Italy.categorical.model.fit <- voomLmFit(Italy.orf.abundance.clean,Italy.categorical.model.design,block= Italy.metadata.clean$patientID,plot=FALSE)
Italy.categorical.model.fit <- eBayes(Italy.categorical.model.fit)

colnames(Italy.categorical.model.fit)
#  [1] "X.Intercept."                                      "Dx.StatusCONTROL"                                 
#  [3] "onset_timeline_combinedt0.6"                       "onset_timeline_combinedt0.12"                     
#  [5] "onset_timeline_combinedt0.18"                      "onset_timeline_combinedt0.24"                     
#  [7] "onset_timeline_combinedt0.30"                      "onset_timeline_combinedt0.36"                     
#  [9] "onset_timeline_combinedt0.over42"                  "SexMale"                                          
# [11] "Age.at.Gluten.Introduction..months."               "HLA.CategoryHigh.Risk"                            
# [13] "HLA.CategoryLow.No.Risk"                           "feeding_first_yearFormula"                        
# [15] "feeding_first_yearBreastmilk_and_formula"          "Delivery.ModeC.Section"                           
# [17] "Dx.StatusCONTROL.onset_timeline_combinedt0.6"      "Dx.StatusCONTROL.onset_timeline_combinedt0.12"    
# [19] "Dx.StatusCONTROL.onset_timeline_combinedt0.18"     "Dx.StatusCONTROL.onset_timeline_combinedt0.24"    
# [21] "Dx.StatusCONTROL.onset_timeline_combinedt0.30"     "Dx.StatusCONTROL.onset_timeline_combinedt0.36"    
# [23] "Dx.StatusCONTROL.onset_timeline_combinedt0.over42"



############# are there any baseline difference in orf abundance between celiac and control group at T0 ##############

Italy.categorical.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL",number=Inf)
Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 0

############# are there significantly different Orfs between Celiac and Control groups at each time point ###############

Italy.contrast_matrix <- makeContrasts(
    "CONTROL_vs_CELIAC_at_t0_6" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.6,
    "CONTROL_vs_CELIAC_at_t0_12" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.12,
    "CONTROL_vs_CELIAC_at_t0_18" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.18,
    "CONTROL_vs_CELIAC_at_t0_24" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.24,
    "CONTROL_vs_CELIAC_at_t0_30" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.30,
    "CONTROL_vs_CELIAC_at_t0_36" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.36,
    "CONTROL_vs_CELIAC_at_t0_over42" = Dx.StatusCONTROL + Dx.StatusCONTROL.onset_timeline_combinedt0.over42,
    levels = Italy.categorical.model.design
  )


# Fit contrasts
Italy.contrast_fit <- contrasts.fit(Italy.categorical.model.fit, Italy.contrast_matrix)
Italy.contrast_fit <- eBayes(Italy.contrast_fit)

colnames(Italy.contrast_fit)
# [1] "CONTROL_vs_CELIAC_at_t0_6"      "CONTROL_vs_CELIAC_at_t0_12"     "CONTROL_vs_CELIAC_at_t0_18"    
# [4] "CONTROL_vs_CELIAC_at_t0_24"     "CONTROL_vs_CELIAC_at_t0_30"     "CONTROL_vs_CELIAC_at_t0_36"    
# [7] "CONTROL_vs_CELIAC_at_t0_over42"

# extract results
Italy.results_t0_6 <- topTable(Italy.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_6",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

Italy.results_t0_12 <- topTable(Italy.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_12",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

Italy.results_t0_18 <- topTable(Italy.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_18",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

Italy.results_t0_24 <- topTable(Italy.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_24",number=Inf) %>% filter(adj.P.Val < 0.05)
# 2570


Italy.results_t0_30 <- topTable(Italy.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_30",number=Inf) %>% filter(adj.P.Val < 0.05)
# 2387


Italy.results_t0_36 <- topTable(Italy.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_36",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0

Italy.results_t0_over42 <- topTable(Italy.contrast_fit, coef = "CONTROL_vs_CELIAC_at_t0_over42",number=Inf) %>% filter(adj.P.Val < 0.05)
# 0



############# are there any orf has significantly changed group difference between each time point and T0 ###############




Italy.categorical.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-12",number=Inf)
Italy.categorical.total.model.res <- Italy.categorical.model.res
Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-12",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 0 6

Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-6",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 0 6

Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-18",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 0 6

Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-24",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 2743 6

View(Italy.categorical.sig.model.res)

Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 2167 6
View(Italy.categorical.sig.model.res)

Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-36",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 0    6


Italy.categorical.sig.model.res <- topTable(Italy.categorical.model.fit,coef="Dx.StatusCONTROL:onset_timeline_combinedt0-over42",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.categorical.sig.model.res)
# 0  6


###############################################

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-6",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 3374  6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-12",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 3493  6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-18",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 2869  6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-24",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0  6

US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-36",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6


US.categorical.sig.model.res <- topTable(US.categorical.model.fit,coef="onset_timeline_combinedt0-over42",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(US.categorical.sig.model.res)
# 0 6

```


# pre-30 vs post-30
```{r}

# prepare correct variable format
Italy.metadata.celiac.clean$feeding_first_year <- factor(Italy.metadata.celiac.clean$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
# Breast_fed      Formula   Breastmilk_and_formula 
#    120             58              129 
Italy.metadata.celiac.clean$HLA.Category <- factor(Italy.metadata.celiac.clean$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
# Standard Risk     High Risk   Low/No Risk 
#      169            94            44
Italy.metadata.celiac.clean$Sex <- factor(Italy.metadata.celiac.clean$Sex,levels = c("Female","Male"))
# Female  Male 
#   238   69
Italy.metadata.celiac.clean$Delivery.Mode <- factor(Italy.metadata.celiac.clean$Delivery.Mode,levels = c("Vaginal","C-Section"))
# Vaginal C-Section 
# 217        90
Italy.metadata.celiac.clean$Age.at.Gluten.Introduction..months. <- as.numeric(Italy.metadata.celiac.clean$Age.at.Gluten.Introduction..months.)
# #  # 4  5  6  7  8  9 10 11 12 15 36 
# #  # 5 11 57 91 40 33 33 10  2 16  9 
Italy.metadata.celiac.clean$Onset_timepoint <- factor(Italy.metadata.celiac.clean$Onset_timepoint,levels = c("pre_30","post_30"))
# pre_30 post_30


Italy.Celiac.categorical.model.design <- model.matrix(~ Onset_timepoint + Sex + Age.at.Gluten.Introduction..months. + feeding_first_year + Delivery.Mode,Italy.metadata.celiac.clean)
Italy.Celiac.categorical.model.fit <- voomLmFit(Italy.orf.abundance.celiac.clean,Italy.Celiac.categorical.model.design,block= Italy.metadata.celiac.clean$patientID,plot=FALSE)
Italy.Celiac.categorical.model.fit <- eBayes(Italy.Celiac.categorical.model.fit)

colnames(Italy.Celiac.categorical.model.fit)
# [1] "(Intercept)"                              "Onset_timepointpost_30"                   "SexMale"                                 
# [4] "Age.at.Gluten.Introduction..months."      "feeding_first_yearFormula"                "feeding_first_yearBreastmilk_and_formula"
# [7] "Delivery.ModeC-Section" 


Italy.Celiac.categorical.sig.model.res <- topTable(Italy.Celiac.categorical.model.fit,coef="Onset_timepointpost_30",number=Inf) %>% filter(adj.P.Val < 0.05)
dim(Italy.Celiac.categorical.sig.model.res)
# 0   6


```




# make the prediction plot for all variables together
```{r}



# Load the complete model results
total_results <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_complete_model_results.csv")
us_results <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/final_results/US_complete_model_results.csv")
italy_results <- read.csv("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/italy_complete_model_results.csv")

# Function to process results for plotting
process_cohort_data <- function(results, cohort_name) {
  # Get coefficient columns (logFC) and their corresponding adj.P.Val columns
  coef_cols <- grep("_logFC$", colnames(results), value = TRUE)
  pval_cols <- gsub("_logFC$", "_adj.P.Val", coef_cols)

    # Create long format data
    plot_data <- data.frame()

    for(i in 1:length(coef_cols)) {
      coef_col <- coef_cols[i]
      pval_col <- pval_cols[i]
      variable_name <- gsub("_logFC$", "", coef_col)

      # Skip intercept
      if(variable_name == "Intercept.") next 

      temp_data <- data.frame(
        cohort = cohort_name,
        variable = variable_name,
        coefficient = -results[[coef_col]],
        adj_p_val = results[[pval_col]],
        significant = results[[pval_col]] < 0.05
      )

      plot_data <- rbind(plot_data, temp_data)
    }

    return(plot_data)
}



# Process all three cohorts
total_data <- process_cohort_data(total_results, "Total")
us_data <- process_cohort_data(us_results, "US")
italy_data <- process_cohort_data(italy_results, "Italy")


# Combine all data
all_data <- rbind(total_data, us_data, italy_data)

# Create the plot with visible boxes and annotations
create_coef_plot <- function(data, cohort_name) {
  # Calculate significance counts for annotations
  sig_counts <- data %>%
    group_by(variable) %>%
    summarise(
      n_significant = sum(significant, na.rm = TRUE),
      n_total = n(),
      .groups = "drop"
    ) %>%
    mutate(
      annotation = paste0("Sig: ", n_significant, "/", n_total),
      max_coef = max(data$coefficient, na.rm = TRUE)
    )

    p <- ggplot(data, aes(x = coefficient, y = variable)) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.7) +

      # Add points first (background layer)
      geom_point(aes(color = significant), alpha = 0.4, size = 0.6,
                 position = position_jitter(height = 0.25, width = 0)) +

      # Add boxplots on top (foreground layer)
      geom_boxplot(alpha = 0.7, outlier.shape = NA,
                   fill = "white", color = "black", linewidth = 0.8) +

      # Add significance count annotations
      geom_text(data = sig_counts,
                aes(x = max_coef * 1.1, y = variable, label = annotation),
                hjust = 0, vjust = 0.5, size = 3, color = "darkblue") +

      scale_color_manual(values = c("FALSE" = "gray60", "TRUE" = "red"),
                         labels = c("Non-significant", "Significant (adj.p < 0.05)"),
                         name = "Significance") +
      labs(title = paste("Coefficient Distribution -", cohort_name, "Cohort"),
           x = "Coefficient Value (logFC)",
           y = "Model Variables") +
      #theme_minimal() +
      theme(axis.text = element_text(size = 12,face = "bold"),
            plot.title = element_text(face = "bold"),
            axis.title.x = element_text(face = "bold"),
            axis.title.y = element_text(face = "bold"),
            legend.position = "bottom",
            plot.margin = margin(5.5, 50, 5.5, 5.5, "pt")) +  # Extra right margin for annotations
      # Expand x-axis to make room for annotations
      scale_x_continuous(expand = expansion(mult = c(0.05, 0.2)))

    return(p)
  }

# Create individual plots
total_plot <- create_coef_plot(total_data, "Total")
us_plot <- create_coef_plot(us_data, "US")
italy_plot <- create_coef_plot(italy_data, "Italy")


# Or combine all three
combined_plot <- grid.arrange(total_plot, us_plot, italy_plot, nrow = 3)

# Save plots
ggsave("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/total_coefficient_plot.png", total_plot, width = 12, height = 8)
ggsave("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/US/final_results/us_coefficient_plot.png", us_plot, width = 12, height = 8)
ggsave("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/Italy/final_results/italy_coefficient_plot.png", italy_plot, width = 12, height = 8)
ggsave("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Orf/data_correct/total/final_results/combined_coefficient_plot.png", combined_plot, width = 12, height = 18)


```







