---
title: "Celiac_phage_phrog_PA_abundance_analysis"
output: html_document
date: "2025-10-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(edgeR)
library(glmmTMB)
library(broom.mixed)  # tidy() for mixed models
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)  # for pivot_longer
library(lme4)  # for glmer
library(geepack)  # for GEE
library(brms)  # for Bayesian models
library(limma)
```





# load phrog file
```{r}

# remember to filter out NA but keep "NO_PHTOG"
phrog <- read_delim("~/Handley Lab Dropbox/16S/Celiac/Phage/phage_detection_pipeline_new_assembly/Virome/phrog/data_correct/total/phold_per_cds_predictions.tsv") %>%
         select(`cds_id`,`phrog`) %>%
         column_to_rownames("cds_id") %>%
         filter(!is.na(phrog))

```



# Load total data
```{r load and clean metadata}

# Total cohort
total.phrog.abundance.clean_0.75_0.03 <- read.csv("data/total/total.phrog.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(total.phrog.abundance.clean_0.75_0.03) <- gsub("X","",colnames(total.phrog.abundance.clean_0.75_0.03))

total.phrog.metadata.clean_0.75_0.03 <- read.csv("data/total/total.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")

# Check if column names of abundance table is consistent with row names of the metadata table
stopifnot(identical(colnames(total.phrog.abundance.clean_0.75_0.03), rownames(total.phrog.metadata.clean_0.75_0.03)))

# Clean metadata
# Prepare correct variable format
total.phrog.metadata.clean_0.75_0.03$feeding_first_year <- factor(total.phrog.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
table(total.phrog.metadata.clean_0.75_0.03$feeding_first_year)

total.phrog.metadata.clean_0.75_0.03$HLA.Category <- factor(total.phrog.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
table(total.phrog.metadata.clean_0.75_0.03$HLA.Category)

total.phrog.metadata.clean_0.75_0.03$Sex <- factor(total.phrog.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
table(total.phrog.metadata.clean_0.75_0.03$Sex)

total.phrog.metadata.clean_0.75_0.03$Delivery.Mode <- factor(total.phrog.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
table(total.phrog.metadata.clean_0.75_0.03$Delivery.Mode)

total.phrog.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(total.phrog.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
table(total.phrog.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)

total.phrog.metadata.clean_0.75_0.03$Dx.Status <- factor(total.phrog.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
table(total.phrog.metadata.clean_0.75_0.03$Dx.Status)

total.phrog.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(total.phrog.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))
table(total.phrog.metadata.clean_0.75_0.03$onset_timeline_combined)

```

# Load US data
```{r create clean abundance tables for three cohorts}

# US cohort
US.phrog.abundance.clean_0.75_0.03 <- read.csv("data/US/US.phrog.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(US.phrog.abundance.clean_0.75_0.03) <- gsub("X","",colnames(US.phrog.abundance.clean_0.75_0.03))

US.phrog.metadata.clean_0.75_0.03 <- read.csv("data/US/US.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")

# Set factor levels for US cohort to match total cohort
US.phrog.metadata.clean_0.75_0.03$feeding_first_year <- factor(US.phrog.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
US.phrog.metadata.clean_0.75_0.03$HLA.Category <- factor(US.phrog.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
US.phrog.metadata.clean_0.75_0.03$Sex <- factor(US.phrog.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
US.phrog.metadata.clean_0.75_0.03$Delivery.Mode <- factor(US.phrog.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
US.phrog.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(US.phrog.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
US.phrog.metadata.clean_0.75_0.03$Dx.Status <- factor(US.phrog.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
US.phrog.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(US.phrog.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))

# Italy cohort
Italy.phrog.abundance.clean_0.75_0.03 <- read.csv("data/Italy/Italy.phrog.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(Italy.phrog.abundance.clean_0.75_0.03) <- gsub("X","",colnames(Italy.phrog.abundance.clean_0.75_0.03))

Italy.phrog.metadata.clean_0.75_0.03 <- read.csv("data/Italy/Italy.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")

# Set factor levels for Italy cohort to match total cohort
Italy.phrog.metadata.clean_0.75_0.03$feeding_first_year <- factor(Italy.phrog.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
Italy.phrog.metadata.clean_0.75_0.03$HLA.Category <- factor(Italy.phrog.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
Italy.phrog.metadata.clean_0.75_0.03$Sex <- factor(Italy.phrog.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
Italy.phrog.metadata.clean_0.75_0.03$Delivery.Mode <- factor(Italy.phrog.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
Italy.phrog.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(Italy.phrog.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
Italy.phrog.metadata.clean_0.75_0.03$Dx.Status <- factor(Italy.phrog.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
Italy.phrog.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(Italy.phrog.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))

```

# Create PA (Presence/Absence) data from abundance data
```{r create PA data}

# Create PA data for all cohorts
total.PA <- (total.phrog.abundance.clean_0.75_0.03 > 0) * 1
US.PA <- (US.phrog.abundance.clean_0.75_0.03 > 0) * 1  
Italy.PA <- (Italy.phrog.abundance.clean_0.75_0.03 > 0) * 1

# Convert to long format for modeling
total_pa_long <- total.PA %>%
  rownames_to_column("phrog_id") %>%
  pivot_longer(cols = -phrog_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(total.phrog.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

US_pa_long <- US.PA %>%
  rownames_to_column("phrog_id") %>%
  pivot_longer(cols = -phrog_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(US.phrog.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

Italy_pa_long <- Italy.PA %>%
  rownames_to_column("phrog_id") %>%
  pivot_longer(cols = -phrog_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(Italy.phrog.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

```

# PA Model 1: glmmTMB with Logit Link

## Total cohort - glmmTMB logit
```{r PA_glmmTMB_logit_total}
# Fit model for each phrog using total cohort data
total_glmmTMB_logit_results <- total_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_glmmTMB_logit_results, "../phrog_Contig_phrog_compositional/phrog_results/total_PA_glmmTMB_logit_results.csv", row.names = FALSE)
```

## US cohort - glmmTMB logit
```{r PA_glmmTMB_logit_US}
# Fit model for each phrog using US cohort data
US_glmmTMB_logit_results <- US_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_glmmTMB_logit_results, "../phrog_Contig_phrog_compositional/phrog_results/US_PA_glmmTMB_logit_results.csv", row.names = FALSE)
```

## Italy cohort - glmmTMB logit
```{r PA_glmmTMB_logit_Italy}
# Fit model for each phrog using Italy cohort data
Italy_glmmTMB_logit_results <- Italy_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_glmmTMB_logit_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_PA_glmmTMB_logit_results.csv", row.names = FALSE)
```


# PA Model 2: glmmTMB with Complementary Log-Log Link

## Total cohort - glmmTMB cloglog
```{r PA_glmmTMB_cloglog_total}
# Fit model for each phrog using total cohort data
total_glmmTMB_cloglog_results <- total_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "cloglog"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_glmmTMB_cloglog_results, "../phrog_Contig_phrog_compositional/phrog_results/total_PA_glmmTMB_cloglog_results.csv", row.names = FALSE)
```

## US cohort - glmmTMB cloglog
```{r PA_glmmTMB_cloglog_US}
# Fit model for each phrog using US cohort data
US_glmmTMB_cloglog_results <- US_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "cloglog"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_glmmTMB_cloglog_results, "../phrog_Contig_phrog_compositional/phrog_results/US_PA_glmmTMB_cloglog_results.csv", row.names = FALSE)
```

## Italy cohort - glmmTMB cloglog
```{r PA_glmmTMB_cloglog_Italy}
# Fit model for each phrog using Italy cohort data
Italy_glmmTMB_cloglog_results <- Italy_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "cloglog"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_glmmTMB_cloglog_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_PA_glmmTMB_cloglog_results.csv", row.names = FALSE)
```


# PA Model 3: glmer (Mixed-effects logistic)
When: You want a second mixed-model engine (very standard).
Why: Solid optimizer, good diagnostics; similar results to glmmTMB on many datasets.
Pros: Widely used; good default.
Cons: No alternative links beyond logit; no zero-inflation (not needed for binary anyway).

## Total cohort - glmer
```{r PA_glmer_total}
library(lme4)

# Fit model for each phrog using total cohort data
total_glmer_results <- total_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmer(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
              HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
              family = binomial(link = "logit"), data = .x,
              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_glmer_results, "../phrog_Contig_phrog_compositional/phrog_results/total_PA_glmer_results.csv", row.names = FALSE)
```

## US cohort - glmer
```{r PA_glmer_US}
# Fit model for each phrog using US cohort data
US_glmer_results <- US_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmer(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
              HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
              family = binomial(link = "logit"), data = .x,
              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_glmer_results, "../phrog_Contig_phrog_compositional/phrog_results/US_PA_glmer_results.csv", row.names = FALSE)
```

## Italy cohort - glmer
```{r PA_glmer_Italy}
# Fit model for each phrog using Italy cohort data
Italy_glmer_results <- Italy_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmer(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
              HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
              family = binomial(link = "logit"), data = .x,
              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_glmer_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_PA_glmer_results.csv", row.names = FALSE)
```


# PA Model 4: GEE (population-averaged logistic)
When: Repeated measures (patientID), convergence issues with random effects, want robust SEs.
Pros: Fewer failures than GLMM; interprets as average effect.
Pros: Robust sandwich SEs; simple to fit.
Cons: No subject-specific effects (interpretation is population-average); no random slopes.

## Total cohort - GEE
```{r PA_GEE_total}
library(geepack)

# Fit model for each phrog using total cohort data
total_gee_results <- total_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        geeglm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
               HLA.Category + feeding_first_year + Delivery.Mode,
               id = patientID, family = binomial, corstr = "exchangeable", data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_gee_results, "../phrog_Contig_phrog_compositional/phrog_results/total_PA_GEE_results.csv", row.names = FALSE)
```

## US cohort - GEE
```{r PA_GEE_US}
# Fit model for each phrog using US cohort data
US_gee_results <- US_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        geeglm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
               HLA.Category + feeding_first_year + Delivery.Mode,
               id = patientID, family = binomial, corstr = "exchangeable", data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_gee_results, "../phrog_Contig_phrog_compositional/phrog_results/US_PA_GEE_results.csv", row.names = FALSE)
```

## Italy cohort - GEE
```{r PA_GEE_Italy}
# Fit model for each phrog using Italy cohort data
Italy_gee_results <- Italy_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        geeglm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
               HLA.Category + feeding_first_year + Delivery.Mode,
               id = patientID, family = binomial, corstr = "exchangeable", data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_gee_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_PA_GEE_results.csv", row.names = FALSE)
```


# PA Model 5: Bayesian mixed-effects logistic (brms/Stan)
When: Ultra-sparse, separation, or you want partial pooling with priors (more stable).
Why: Priors stabilize estimates; rich inference (CIs/ORs); handles complex structures.
Pros: Very stable with sparsity; full posteriors; can change link (e.g., cloglog).
Cons: Slower; requires MCMC checks (R̂, ESS, pp-checks).

## Total cohort - Bayesian brms
```{r PA_brms_total}
library(brms)

# Fit model for each phrog using total cohort data
total_brms_results <- total_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        brm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
            HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
            family = bernoulli(link = "logit"), data = .x,
            prior = c(
              set_prior("normal(0, 1)", class = "b"),     # fixed effects
              set_prior("exponential(2)", class = "sd")   # random-effect SDs
            ),
            cores = 4, chains = 4, iter = 2000, refresh = 0)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_brms_results, "../phrog_Contig_phrog_compositional/phrog_results/total_PA_brms_results.csv", row.names = FALSE)
```

## US cohort - Bayesian brms
```{r PA_brms_US}
# Fit model for each phrog using US cohort data
US_brms_results <- US_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        brm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
            HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
            family = bernoulli(link = "logit"), data = .x,
            prior = c(
              set_prior("normal(0, 1)", class = "b"),     # fixed effects
              set_prior("exponential(2)", class = "sd")   # random-effect SDs
            ),
            cores = 4, chains = 4, iter = 2000, refresh = 0)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_brms_results, "../phrog_Contig_phrog_compositional/phrog_results/US_PA_brms_results.csv", row.names = FALSE)
```

## Italy cohort - Bayesian brms
```{r PA_brms_Italy}
# Fit model for each phrog using Italy cohort data
Italy_brms_results <- Italy_pa_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        brm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
            HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
            family = bernoulli(link = "logit"), data = .x,
            prior = c(
              set_prior("normal(0, 1)", class = "b"),     # fixed effects
              set_prior("exponential(2)", class = "sd")   # random-effect SDs
            ),
            cores = 4, chains = 4, iter = 2000, refresh = 0)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_brms_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_PA_brms_results.csv", row.names = FALSE)
```


# A) Count GLMMs (per-phrog), with random effects

# Abundance Model 1: Negative Binomial GLMM — workhorse

## Total cohort - Negative Binomial GLMM
```{r abundance_NB_total}
library(edgeR)
library(glmmTMB)

# Prepare DGEList and normalization for total cohort
total.y <- DGEList(counts = total.phrog.abundance.clean_0.75_0.03, samples = total.phrog.metadata.clean_0.75_0.03)
total.y <- calcNormFactors(total.y, method = "TMMwsp")
total.lib_eff <- with(total.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
total_abundance_long <- total.phrog.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("phrog_id") %>%
  pivot_longer(cols = -phrog_id, names_to = "sample_id", values_to = "count") %>%
  left_join(total.phrog.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(offset = log(total.lib_eff[sample_id]))

# Fit NB model for each phrog
total_nb_results <- total_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_nb_results, "../phrog_Contig_phrog_compositional/phrog_results/total_abundance_NB_results.csv", row.names = FALSE)
```

## US cohort - Negative Binomial GLMM
```{r abundance_NB_US}
# Prepare DGEList and normalization for US cohort
US.y <- DGEList(counts = US.phrog.abundance.clean_0.75_0.03, samples = US.phrog.metadata.clean_0.75_0.03)
US.y <- calcNormFactors(US.y, method = "TMMwsp")
US.lib_eff <- with(US.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
US_abundance_long <- US.phrog.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("phrog_id") %>%
  pivot_longer(cols = -phrog_id, names_to = "sample_id", values_to = "count") %>%
  left_join(US.phrog.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(offset = log(US.lib_eff[sample_id]))

# Fit NB model for each phrog
US_nb_results <- US_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_nb_results, "../phrog_Contig_phrog_compositional/phrog_results/US_abundance_NB_results.csv", row.names = FALSE)
```

## Italy cohort - Negative Binomial GLMM
```{r abundance_NB_Italy}
# Prepare DGEList and normalization for Italy cohort
Italy.y <- DGEList(counts = Italy.phrog.abundance.clean_0.75_0.03, samples = Italy.phrog.metadata.clean_0.75_0.03)
Italy.y <- calcNormFactors(Italy.y, method = "TMMwsp")
Italy.lib_eff <- with(Italy.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
Italy_abundance_long <- Italy.phrog.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("phrog_id") %>%
  pivot_longer(cols = -phrog_id, names_to = "sample_id", values_to = "count") %>%
  left_join(Italy.phrog.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(offset = log(Italy.lib_eff[sample_id]))

# Fit NB model for each phrog
Italy_nb_results <- Italy_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_nb_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_abundance_NB_results.csv", row.names = FALSE)
```


# Abundance Model 2: Hurdle Negative Binomial GLMM — excess zeros, two-part

## Total cohort - Hurdle Negative Binomial GLMM
```{r abundance_hurdle_total}
# Fit Hurdle NB model for each phrog using total cohort data
total_hurdle_results <- total_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                           HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = truncated_nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_hurdle_results, "../phrog_Contig_phrog_compositional/phrog_results/total_abundance_hurdle_results.csv", row.names = FALSE)
```

## US cohort - Hurdle Negative Binomial GLMM
```{r abundance_hurdle_US}
# Fit Hurdle NB model for each phrog using US cohort data
US_hurdle_results <- US_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                           HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = truncated_nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_hurdle_results, "../phrog_Contig_phrog_compositional/phrog_results/US_abundance_hurdle_results.csv", row.names = FALSE)
```

## Italy cohort - Hurdle Negative Binomial GLMM
```{r abundance_hurdle_Italy}
# Fit Hurdle NB model for each phrog using Italy cohort data
Italy_hurdle_results <- Italy_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                           HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = truncated_nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_hurdle_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_abundance_hurdle_results.csv", row.names = FALSE)
```


# Abundance Model 3: Zero-Inflated NB GLMM — structural zeros

## Total cohort - Zero-Inflated NB GLMM
```{r abundance_ZINB_total}
# Fit Zero-Inflated NB model for each phrog using total cohort data
total_zinb_results <- total_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                           HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_zinb_results, "../phrog_Contig_phrog_compositional/phrog_results/total_abundance_ZINB_results.csv", row.names = FALSE)
```

## US cohort - Zero-Inflated NB GLMM
```{r abundance_ZINB_US}
# Fit Zero-Inflated NB model for each phrog using US cohort data
US_zinb_results <- US_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                           HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_zinb_results, "../phrog_Contig_phrog_compositional/phrog_results/US_abundance_ZINB_results.csv", row.names = FALSE)
```

## Italy cohort - Zero-Inflated NB GLMM
```{r abundance_ZINB_Italy}
# Fit Zero-Inflated NB model for each phrog using Italy cohort data
Italy_zinb_results <- Italy_abundance_long %>%
  group_by(phrog_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                           HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(phrog_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_zinb_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_abundance_ZINB_results.csv", row.names = FALSE)
```


# B) Voom / Linear Mixed Models on log-counts (rate scale)

# Abundance Model 4: limma-voom + duplicateCorrelation — fast, empirical Bayes

## Total cohort - limma-voom + duplicateCorrelation
```{r abundance_voom_total}
library(limma)
library(edgeR)

# Prepare design matrix for total cohort
total_design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                            HLA.Category + feeding_first_year + Delivery.Mode, 
                            data = total.phrog.metadata.clean_0.75_0.03)

# Apply voom transformation
total_v <- voom(total.y, total_design, plot = FALSE)

# Calculate duplicate correlation
total_cphrogit <- duplicateCorrelation(total_v, total_design, block = total.phrog.metadata.clean_0.75_0.03$patientID)

# Re-apply voom with correlation
total_v2 <- voom(total.y, total_design, plot = FALSE, 
                 block = total.phrog.metadata.clean_0.75_0.03$patientID, 
                 correlation = total_cphrogit$consensus)

# Fit linear model
total_voom_fit <- lmFit(total_v2, total_design, 
                       block = total.phrog.metadata.clean_0.75_0.03$patientID, 
                       correlation = total_cphrogit$consensus)
total_voom_fit <- eBayes(total_voom_fit)

# Extract results for all coefficients
total_voom_results <- topTable(total_voom_fit, number = Inf) %>%
  rownames_to_column("phrog_id") %>%
  mutate(cohort = "total")

# Save results
write.csv(total_voom_results, "../phrog_Contig_phrog_compositional/phrog_results/total_abundance_voom_results.csv", row.names = FALSE)
```

## US cohort - limma-voom + duplicateCorrelation
```{r abundance_voom_US}
# Prepare design matrix for US cohort
US_design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                         HLA.Category + feeding_first_year + Delivery.Mode, 
                         data = US.phrog.metadata.clean_0.75_0.03)

# Apply voom transformation
US_v <- voom(US.y, US_design, plot = FALSE)

# Calculate duplicate correlation
US_cphrogit <- duplicateCorrelation(US_v, US_design, block = US.phrog.metadata.clean_0.75_0.03$patientID)

# Re-apply voom with correlation
US_v2 <- voom(US.y, US_design, plot = FALSE, 
              block = US.phrog.metadata.clean_0.75_0.03$patientID, 
              correlation = US_cphrogit$consensus)

# Fit linear model
US_voom_fit <- lmFit(US_v2, US_design, 
                     block = US.phrog.metadata.clean_0.75_0.03$patientID, 
                     correlation = US_cphrogit$consensus)
US_voom_fit <- eBayes(US_voom_fit)

# Extract results for all coefficients
US_voom_results <- topTable(US_voom_fit, number = Inf) %>%
  rownames_to_column("phrog_id") %>%
  mutate(cohort = "US")

# Save results
write.csv(US_voom_results, "../phrog_Contig_phrog_compositional/phrog_results/US_abundance_voom_results.csv", row.names = FALSE)
```

## Italy cohort - limma-voom + duplicateCorrelation
```{r abundance_voom_Italy}
# Prepare design matrix for Italy cohort
Italy_design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                            HLA.Category + feeding_first_year + Delivery.Mode, 
                            data = Italy.phrog.metadata.clean_0.75_0.03)

# Apply voom transformation
Italy_v <- voom(Italy.y, Italy_design, plot = FALSE)

# Calculate duplicate correlation
Italy_cphrogit <- duplicateCorrelation(Italy_v, Italy_design, block = Italy.phrog.metadata.clean_0.75_0.03$patientID)

# Re-apply voom with correlation
Italy_v2 <- voom(Italy.y, Italy_design, plot = FALSE, 
                 block = Italy.phrog.metadata.clean_0.75_0.03$patientID, 
                 correlation = Italy_cphrogit$consensus)

# Fit linear model
Italy_voom_fit <- lmFit(Italy_v2, Italy_design, 
                       block = Italy.phrog.metadata.clean_0.75_0.03$patientID, 
                       correlation = Italy_cphrogit$consensus)
Italy_voom_fit <- eBayes(Italy_voom_fit)

# Extract results for all coefficients
Italy_voom_results <- topTable(Italy_voom_fit, number = Inf) %>%
  rownames_to_column("phrog_id") %>%
  mutate(cohort = "Italy")

# Save results
write.csv(Italy_voom_results, "../phrog_Contig_phrog_compositional/phrog_results/Italy_abundance_voom_results.csv", row.names = FALSE)
```




