---
title: "Celiac_phage_orf_PA_abundance_analysis"
output: html_document
date: "2025-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(edgeR)
library(glmmTMB)
library(broom.mixed)  # tidy() for mixed models
library(dplyr)
library(purrr)
library(tibble)
```




# load clean abundant table
```{r create clean abundance tables for three cohorts}

# total cohort
total.orf.abundance.clean_0.75_0.03 <- read.csv("../Orf/data_correct/total/total.orf.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(total.orf.abundance.clean_0.75_0.03) <- gsub("X","",colnames(total.orf.abundance.clean_0.75_0.03))

total.orf.metadata.clean_0.75_0.03 <- read.csv("../Orf/data_correct/total/total.orf.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")



# US cohort
US.orf.abundance.clean_0.75_0.03 <- read.csv("../Orf/data_correct/US/US.orf.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(US.orf.abundance.clean_0.75_0.03) <- gsub("X","",colnames(US.orf.abundance.clean_0.75_0.03))


US.orf.metadata.clean_0.75_0.03 <- read.csv("../Orf/data_correct/US/US.orf.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")

# Set factor levels for US cohort to match total cohort
US.orf.metadata.clean_0.75_0.03$feeding_first_year <- factor(US.orf.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
US.orf.metadata.clean_0.75_0.03$HLA.Category <- factor(US.orf.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
US.orf.metadata.clean_0.75_0.03$Sex <- factor(US.orf.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
US.orf.metadata.clean_0.75_0.03$Delivery.Mode <- factor(US.orf.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
US.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(US.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
US.orf.metadata.clean_0.75_0.03$Dx.Status <- factor(US.orf.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
US.orf.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(US.orf.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))

# Italy cohort
Italy.orf.abundance.clean_0.75_0.03 <- read.csv("../Orf/data_correct/Italy/Italy.orf.abundance.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")
colnames(Italy.orf.abundance.clean_0.75_0.03) <- gsub("X","",colnames(Italy.orf.abundance.clean_0.75_0.03))


Italy.orf.metadata.clean_0.75_0.03 <- read.csv("../Orf/data_correct/Italy/Italy.orf.metadata.clean_0.75_0.03.csv") %>%
                                          column_to_rownames("X")

# Set factor levels for Italy cohort to match total cohort
Italy.orf.metadata.clean_0.75_0.03$feeding_first_year <- factor(Italy.orf.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
Italy.orf.metadata.clean_0.75_0.03$HLA.Category <- factor(Italy.orf.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
Italy.orf.metadata.clean_0.75_0.03$Sex <- factor(Italy.orf.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
Italy.orf.metadata.clean_0.75_0.03$Delivery.Mode <- factor(Italy.orf.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
Italy.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(Italy.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
Italy.orf.metadata.clean_0.75_0.03$Dx.Status <- factor(Italy.orf.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
Italy.orf.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(Italy.orf.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))

```



# load clean metadata
```{r load and clean metadata}

# check if column names of abundance table is consistent with row names of the metadata table
stopifnot(identical(colnames(total.orf.abundance.clean_0.75_0.03), rownames(total.orf.metadata.clean_0.75_0.03)))

# clean metadata
# prepare correct variable format
total.orf.metadata.clean_0.75_0.03$feeding_first_year <- factor(total.orf.metadata.clean_0.75_0.03$feeding_first_year,levels = c("Breast_fed","Formula","Breastmilk_and_formula"))
table(total.orf.metadata.clean_0.75_0.03$feeding_first_year)
# Breast_fed  Formula       Breastmilk_and_formula 
#  100            40                    103
total.orf.metadata.clean_0.75_0.03$HLA.Category <- factor(total.orf.metadata.clean_0.75_0.03$HLA.Category,levels = c("Standard Risk","High Risk","Low/No Risk"))
table(total.orf.metadata.clean_0.75_0.03$HLA.Category)
# Standard Risk     High Risk   Low/No Risk 
#      123            82            38
total.orf.metadata.clean_0.75_0.03$Sex <- factor(total.orf.metadata.clean_0.75_0.03$Sex,levels = c("Female","Male"))
table(total.orf.metadata.clean_0.75_0.03$Sex)
# Female   Male 
#    194     49 
total.orf.metadata.clean_0.75_0.03$Delivery.Mode <- factor(total.orf.metadata.clean_0.75_0.03$Delivery.Mode,levels = c("Vaginal","C-Section"))
table(total.orf.metadata.clean_0.75_0.03$Delivery.Mode)
  # Vaginal C-Section 
  #     180     63 
total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months. <- as.numeric(total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
table(total.orf.metadata.clean_0.75_0.03$Age.at.Gluten.Introduction..months.)
 # 4  5  6  7  8  9 10 11 12 15 36 
 # 4  9 46 64 32 27 29  7  1 15  9
total.orf.metadata.clean_0.75_0.03$Dx.Status <- factor(total.orf.metadata.clean_0.75_0.03$Dx.Status,levels = c("CONTROL","CELIAC"))
table(total.orf.metadata.clean_0.75_0.03$Dx.Status)
 # CELIAC CONTROL 
 #    114    129 
total.orf.metadata.clean_0.75_0.03$onset_timeline_combined <- factor(total.orf.metadata.clean_0.75_0.03$onset_timeline_combined,levels = c("t0","t0-6","t0-12","t0-18","t0-24","t0-30","t0-36","t0-over42"))
table(total.orf.metadata.clean_0.75_0.03$onset_timeline_combined)
# t0      t0-6     t0-12     t0-18     t0-24     t0-30     t0-36 t0-over42 
# 53        20        40        19        23        25        27        36 


```



# create total PA table
```{r generate total PA table}

total.target_cpm      <- 0.5   # e.g., 0.25–1.0 for very sparse data
total.min_reads_floor <- 3L    # guard against singletons

## ---- 1) Align samples ------------------------------------------------------
stopifnot(all(colnames(total.orf.abundance.clean_0.75_0.03) %in% rownames(total.orf.metadata.clean_0.75_0.03)))
# Reorder metadata to match counts columns
total.orf.metadata.clean_0.75_0.03 <- total.orf.metadata.clean_0.75_0.03[colnames(total.orf.abundance.clean_0.75_0.03), , drop = FALSE]

## ---- 2) Build DGEList & normalize (TMMwsp) --------------------------------

total.y <- DGEList(counts = total.orf.abundance.clean_0.75_0.03, samples = total.orf.metadata.clean_0.75_0.03)
# Optional: drop genes with all zeros (purely empty rows)
total.y <- total.y[rowSums(total.y$counts) > 0, , keep.lib.sizes = FALSE]

# Library-size normalization (robust for sparse data)
total.y <- calcNormFactors(total.y, method = "TMMwsp")

## ---- 3) CPM-anchored presence/absence -------------------------------------
# Normalized CPMs (uses effective library sizes = lib.size * norm.factors)
total.cpm_mat <- edgeR::cpm(total.y, normalized.lib.sizes = TRUE)

# Presence rule: CPM >= target_cpm AND raw count >= min_reads_floor
total.present_logical <- (total.cpm_mat >= total.target_cpm) & (total.y$counts >= total.min_reads_floor)

# Convert to 0/1 matrix with dimnames preserved
total.PA <- matrix(
  as.integer(total.present_logical),
  nrow = nrow(total.present_logical),
  ncol = ncol(total.present_logical),
  dimnames = dimnames(total.present_logical)
)

dim(total.PA)
# 109 243

## ---- 4) Quick sanity checks ------------------------------------------------
# How many “present” calls overall?
mean(total.PA)                      # fraction of 1s
summary(rowSums(total.PA))          # #samples present per gene
summary(colSums(total.PA))          # #genes present per sample

# Optional: keep genes present in at least M samples (global prevalence)
total.M <- 6L
total.keep <- rowSums(total.PA) >= M
total.PA_kept <- total.PA[total.keep, , drop = FALSE]
dim(total.PA_kept)
# 109 243


```


# create US PA table
```{r generate US PA table}

US.target_cpm  <- 0.5   # e.g., 0.25–1.0 for very sparse data
US.min_reads_floor <- 3L    # guard against singletons

## ---- 1) Align samples ------------------------------------------------------
stopifnot(all(colnames(US.orf.abundance.clean_0.75_0.03) %in% rownames(US.orf.metadata.clean_0.75_0.03)))
# Reorder metadata to match counts columns
US.orf.metadata.clean_0.75_0.03 <- US.orf.metadata.clean_0.75_0.03[colnames(US.orf.abundance.clean_0.75_0.03), , drop = FALSE]
sum(US.orf.abundance.clean_0.75_0.03 > 0)
#1547

## ---- 2) Build DGEList & normalize (TMMwsp) --------------------------------

US.y <- DGEList(counts = US.orf.abundance.clean_0.75_0.03, samples = US.orf.metadata.clean_0.75_0.03)
# Optional: drop genes with all zeros (purely empty rows)
US.y <- US.y[rowSums(US.y$counts) > 0, , keep.lib.sizes = FALSE]


# Library-size normalization (robust for sparse data)
US.y <- calcNormFactors(US.y, method = "TMMwsp")


## ---- 3) CPM-anchored presence/absence -------------------------------------
# Normalized CPMs (uses effective library sizes = lib.size * norm.factors)
US.cpm_mat <- edgeR::cpm(US.y, normalized.lib.sizes = TRUE)


# Presence rule: CPM >= target_cpm AND raw count >= min_reads_floor
US.present_logical <- (US.cpm_mat >= US.target_cpm) & (US.y$counts >= US.min_reads_floor)
# FALSE  TRUE 
# 24703  1547 

# every non-zero count passed

# Convert to 0/1 matrix with dimnames preserved
US.PA <- matrix(
  as.integer(US.present_logical),
  nrow = nrow(US.present_logical),
  ncol = ncol(US.present_logical),
  dimnames = dimnames(US.present_logical)
)

dim(US.PA)
# 150 175

## ---- 4) Quick sanity checks ------------------------------------------------
# How many “present” calls overall?
mean(US.PA) # 0.05893333  
summary(rowSums(US.PA))          # #samples present per gene
summary(colSums(US.PA))          # #genes present per sample

# Optional: keep genes present in at least M samples (global prevalence)
US.M <- 6L
US.keep <- rowSums(US.PA) >= US.M
US.PA_kept <- US.PA[US.keep, , drop = FALSE]
# 150 175


```


# create Italy PA table
```{r generate Italy PA table}


Italy.target_cpm      <- 0.5   # e.g., 0.25–1.0 for very sparse data
Italy.min_reads_floor <- 3L    # guard against singletons

## ---- 1) Align samples ------------------------------------------------------
stopifnot(all(colnames(Italy.orf.abundance.clean_0.75_0.03) %in% rownames(Italy.orf.metadata.clean_0.75_0.03)))
# Reorder metadata to match counts columns
Italy.orf.metadata.clean_0.75_0.03 <- Italy.orf.metadata.clean_0.75_0.03[colnames(Italy.orf.abundance.clean_0.75_0.03), , drop = FALSE]

## ---- 2) Build DGEList & normalize (TMMwsp) --------------------------------

Italy.y <- DGEList(counts = Italy.orf.abundance.clean_0.75_0.03, samples = Italy.orf.metadata.clean_0.75_0.03)
# Optional: drop genes with all zeros (purely empty rows)
Italy.y <- Italy.y[rowSums(Italy.y$counts) > 0, , keep.lib.sizes = FALSE]

# Library-size normalization (robItalyt for sparse data)
Italy.y <- calcNormFactors(Italy.y, method = "TMMwsp")

## ---- 3) CPM-anchored presence/absence -------------------------------------
# Normalized CPMs (Italyes effective library sizes = lib.size * norm.factors)
Italy.cpm_mat <- edgeR::cpm(Italy.y, normalized.lib.sizes = TRUE)

# Presence rule: CPM >= target_cpm AND raw count >= min_reads_floor
Italy.present_logical <- (Italy.cpm_mat >= Italy.target_cpm) & (Italy.y$counts >= Italy.min_reads_floor)

# Convert to 0/1 matrix with dimnames preserved
Italy.PA <- matrix(
  as.integer(Italy.present_logical),
  nrow = nrow(Italy.present_logical),
  ncol = ncol(Italy.present_logical),
  dimnames = dimnames(Italy.present_logical)
)

dim(Italy.PA)
# 125  83



```


# PA ANALYSIS MODELS

# PA model 1 - glmmTMB (Mixed-effects logistic) - logit link
When: Default choice for pooled-over-time models with within-patient correlation.
Why: True random intercept (and optional random slope), fast, flexible links.
Pros: Handles complex fixed effects; supports cloglog link (family = binomial("cloglog")) if events are very rare.
Cons: Can struggle with extreme sparsity/separation—try GEE or Bayesian if convergence issues persist.

## Total cohort - glmmTMB logit
```{r PA_glmmTMB_logit_total}
library(glmmTMB)
library(broom.mixed)
library(tidyr)

# Create long-format data for total cohort
total_pa_long <- total.PA %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(total.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

# Fit model for each ORF
total_glmmTMB_logit_results <- total_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_glmmTMB_logit_results, "../Orf_Contig_Phrog_compositional/orf_results/total_PA_model1_glmmTMB_logit.csv", row.names = FALSE)
```

## US cohort - glmmTMB logit
```{r PA_glmmTMB_logit_US}
# Create long-format data for US cohort
US_pa_long <- US.PA %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(US.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

# Fit model for each ORF
US_glmmTMB_logit_results <- US_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_glmmTMB_logit_results, "../Orf_Contig_Phrog_compositional/orf_results/US_PA_model1_glmmTMB_logit.csv", row.names = FALSE)
```

## Italy cohort - glmmTMB logit
```{r PA_glmmTMB_logit_Italy}
# Create long-format data for Italy cohort
Italy_pa_long <- Italy.PA %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "PA") %>%
  left_join(Italy.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id")

# Fit model for each ORF
Italy_glmmTMB_logit_results <- Italy_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "logit"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_glmmTMB_logit_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_PA_model1_glmmTMB_logit.csv", row.names = FALSE)
```


# PA model 2 - glmmTMB (Mixed-effects logistic) - cloglog link
When: Events are very rare (asymmetric). Sometimes converges better than logit.

## Total cohort - glmmTMB cloglog
```{r PA_glmmTMB_cloglog_total}
# Fit model for each ORF using total cohort data
total_glmmTMB_cloglog_results <- total_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "cloglog"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_glmmTMB_cloglog_results, "../Orf_Contig_Phrog_compositional/orf_results/total_PA_model2_glmmTMB_cloglog.csv", row.names = FALSE)
```

## US cohort - glmmTMB cloglog
```{r PA_glmmTMB_cloglog_US}
# Fit model for each ORF using US cohort data
US_glmmTMB_cloglog_results <- US_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "cloglog"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_glmmTMB_cloglog_results, "../Orf_Contig_Phrog_compositional/orf_results/US_PA_model2_glmmTMB_cloglog.csv", row.names = FALSE)
```

## Italy cohort - glmmTMB cloglog
```{r PA_glmmTMB_cloglog_Italy}
# Fit model for each ORF using Italy cohort data
Italy_glmmTMB_cloglog_results <- Italy_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
                family = binomial(link = "cloglog"), data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_glmmTMB_cloglog_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_PA_model2_glmmTMB_cloglog.csv", row.names = FALSE)
```


# PA model 3 - glmer (Mixed-effects logistic)
When: You want a second mixed-model engine (very standard).
Why: Solid optimizer, good diagnostics; similar results to glmmTMB on many datasets.
Pros: Widely used; good default.
Cons: No alternative links beyond logit; no zero-inflation (not needed for binary anyway).

## Total cohort - glmer
```{r PA_glmer_total}
library(lme4)

# Fit model for each ORF using total cohort data
total_glmer_results <- total_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmer(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
              HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
              family = binomial(link = "logit"), data = .x,
              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_glmer_results, "../Orf_Contig_Phrog_compositional/orf_results/total_PA_model3_glmer.csv", row.names = FALSE)
```

## US cohort - glmer
```{r PA_glmer_US}
# Fit model for each ORF using US cohort data
US_glmer_results <- US_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmer(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
              HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
              family = binomial(link = "logit"), data = .x,
              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_glmer_results, "../Orf_Contig_Phrog_compositional/orf_results/US_PA_model3_glmer.csv", row.names = FALSE)
```

## Italy cohort - glmer
```{r PA_glmer_Italy}
# Fit model for each ORF using Italy cohort data
Italy_glmer_results <- Italy_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmer(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
              HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
              family = binomial(link = "logit"), data = .x,
              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_glmer_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_PA_model3_glmer.csv", row.names = FALSE)
```


# PA model 4 - GEE (population-averaged logistic)
When: Repeated measures (patientID), convergence issues with random effects, want population-averaged effects.
Pros: Fewer failures than GLMM; interprets as average effect.
Pros: Robust sandwich SEs; simple to fit.
Cons: No subject-specific effects (interpretation is population-average); no random slopes.

## Total cohort - GEE
```{r PA_GEE_total}
library(geepack)

# Fit model for each ORF using total cohort data
total_gee_results <- total_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        geeglm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
               HLA.Category + feeding_first_year + Delivery.Mode,
               id = patientID, family = binomial, corstr = "exchangeable", data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_gee_results, "../Orf_Contig_Phrog_compositional/orf_results/total_PA_model4_GEE.csv", row.names = FALSE)
```

## US cohort - GEE
```{r PA_GEE_US}
# Fit model for each ORF using US cohort data
US_gee_results <- US_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        geeglm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
               HLA.Category + feeding_first_year + Delivery.Mode,
               id = patientID, family = binomial, corstr = "exchangeable", data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_gee_results, "../Orf_Contig_Phrog_compositional/orf_results/US_PA_model4_GEE.csv", row.names = FALSE)
```

## Italy cohort - GEE
```{r PA_GEE_Italy}
# Fit model for each ORF using Italy cohort data
Italy_gee_results <- Italy_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        geeglm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
               HLA.Category + feeding_first_year + Delivery.Mode,
               id = patientID, family = binomial, corstr = "exchangeable", data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_gee_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_PA_model4_GEE.csv", row.names = FALSE)
```


# PA model 5 - Bayesian mixed-effects logistic (brms/Stan)
When: Ultra-sparse, separation, or you want partial pooling with priors (more stable).
Why: Priors stabilize estimates; rich inference (CIs/ORs); handles complex structures.
Pros: Very stable with sparsity; full posteriors; can change link (e.g., cloglog).
Cons: Slower; requires MCMC checks (R̂, ESS, pp-checks).

## Total cohort - Bayesian brms
```{r PA_brms_total}
library(brms)

# Fit model for each ORF using total cohort data
total_brms_results <- total_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        brm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
            HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
            family = bernoulli(link = "logit"), data = .x,
            prior = c(
              set_prior("normal(0, 1)", class = "b"),     # fixed effects
              set_prior("exponential(2)", class = "sd")   # random-effect SDs
            ),
            cores = 4, chains = 4, iter = 2000, refresh = 0)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_brms_results, "../Orf_Contig_Phrog_compositional/orf_results/total_PA_model5_brms.csv", row.names = FALSE)
```

## US cohort - Bayesian brms
```{r PA_brms_US}
# Fit model for each ORF using US cohort data
US_brms_results <- US_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        brm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
            HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
            family = bernoulli(link = "logit"), data = .x,
            prior = c(
              set_prior("normal(0, 1)", class = "b"),     # fixed effects
              set_prior("exponential(2)", class = "sd")   # random-effect SDs
            ),
            cores = 4, chains = 4, iter = 2000, refresh = 0)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_brms_results, "../Orf_Contig_Phrog_compositional/orf_results/US_PA_model5_brms.csv", row.names = FALSE)
```

## Italy cohort - Bayesian brms
```{r PA_brms_Italy}
# Fit model for each ORF using Italy cohort data
Italy_brms_results <- Italy_pa_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        brm(PA ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
            HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID),
            family = bernoulli(link = "logit"), data = .x,
            prior = c(
              set_prior("normal(0, 1)", class = "b"),     # fixed effects
              set_prior("exponential(2)", class = "sd")   # random-effect SDs
            ),
            cores = 4, chains = 4, iter = 2000, refresh = 0)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_brms_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_PA_model5_brms.csv", row.names = FALSE)
```


# ABUNDANCE ANALYSIS MODELS

# Abundance model 1 - Negative Binomial GLMM — workhorse

## Total cohort - Negative Binomial GLMM
```{r abundance_NB_total}
library(edgeR)
library(glmmTMB)

# Prepare DGEList and normalization for total cohort
total.y <- DGEList(counts = total.orf.abundance.clean_0.75_0.03, samples = total.orf.metadata.clean_0.75_0.03)
total.y <- calcNormFactors(total.y, method = "TMMwsp")
total.lib_eff <- with(total.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
total_abundance_long <- total.orf.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "count") %>%
  left_join(total.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(offset = log(total.lib_eff[sample_id]))

# Fit NB model for each ORF
total_nb_results <- total_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_nb_results, "../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model1_nbinom.csv", row.names = FALSE)
```

## US cohort - Negative Binomial GLMM
```{r abundance_NB_US}
# Prepare DGEList and normalization for US cohort
US.y <- DGEList(counts = US.orf.abundance.clean_0.75_0.03, samples = US.orf.metadata.clean_0.75_0.03)
US.y <- calcNormFactors(US.y, method = "TMMwsp")
US.lib_eff <- with(US.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
US_abundance_long <- US.orf.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "count") %>%
  left_join(US.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(offset = log(US.lib_eff[sample_id]))

# Fit NB model for each ORF
US_nb_results <- US_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_nb_results, "../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model1_nbinom.csv", row.names = FALSE)
```

## Italy cohort - Negative Binomial GLMM
```{r abundance_NB_Italy}
# Prepare DGEList and normalization for Italy cohort
Italy.y <- DGEList(counts = Italy.orf.abundance.clean_0.75_0.03, samples = Italy.orf.metadata.clean_0.75_0.03)
Italy.y <- calcNormFactors(Italy.y, method = "TMMwsp")
Italy.lib_eff <- with(Italy.y$samples, lib.size * norm.factors)

# Create long-format data for abundance modeling
Italy_abundance_long <- Italy.orf.abundance.clean_0.75_0.03 %>%
  as.data.frame() %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "count") %>%
  left_join(Italy.orf.metadata.clean_0.75_0.03 %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(offset = log(Italy.lib_eff[sample_id]))

# Fit NB model for each ORF
Italy_nb_results <- Italy_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_nb_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_abundance_model1_nbinom.csv", row.names = FALSE)
```


# Abundance model 2 - Hurdle Negative Binomial GLMM
When: Many zeros that are structurally different from count zeros.

## Total cohort - Hurdle NB
```{r abundance_hurdle_total}
# Fit hurdle NB model for each ORF using total cohort data
total_hurdle_results <- total_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = truncated_nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_hurdle_results, "../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model2_hurdle.csv", row.names = FALSE)
```

## US cohort - Hurdle NB
```{r abundance_hurdle_US}
# Fit hurdle NB model for each ORF using US cohort data
US_hurdle_results <- US_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = truncated_nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_hurdle_results, "../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model2_hurdle.csv", row.names = FALSE)
```

## Italy cohort - Hurdle NB
```{r abundance_hurdle_Italy}
# Fit hurdle NB model for each ORF using Italy cohort data
Italy_hurdle_results <- Italy_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                family = truncated_nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_hurdle_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_abundance_model2_hurdle.csv", row.names = FALSE)
```


# Abundance model 3 - Zero-Inflated Negative Binomial GLMM
When: Excess zeros beyond what NB can handle, and you think some are "structural" zeros.

## Total cohort - Zero-Inflated NB
```{r abundance_zinb_total}
# Fit ZINB model for each ORF using total cohort data
total_zinb_results <- total_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ 1,
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(total_zinb_results, "../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model3_zinb.csv", row.names = FALSE)
```

## US cohort - Zero-Inflated NB
```{r abundance_zinb_US}
# Fit ZINB model for each ORF using US cohort data
US_zinb_results <- US_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ 1,
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(US_zinb_results, "../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model3_zinb.csv", row.names = FALSE)
```

## Italy cohort - Zero-Inflated NB
```{r abundance_zinb_Italy}
# Fit ZINB model for each ORF using Italy cohort data
Italy_zinb_results <- Italy_abundance_long %>%
  group_by(orf_id) %>%
  nest() %>%
  mutate(
    model = map(data, ~ {
      tryCatch({
        glmmTMB(count ~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                HLA.Category + feeding_first_year + Delivery.Mode + (1 | patientID) + offset(offset),
                ziformula = ~ 1,
                family = nbinom2, data = .x)
      }, error = function(e) NULL)
    }),
    tidy_results = map(model, ~ {
      if (!is.null(.x)) {
        tryCatch(tidy(.x, conf.int = TRUE), error = function(e) NULL)
      } else NULL
    })
  ) %>%
  select(orf_id, tidy_results) %>%
  unnest(tidy_results)

# Save results
write.csv(Italy_zinb_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_abundance_model3_zinb.csv", row.names = FALSE)
```


# Abundance model 4 - limma-voom + duplicateCorrelation
When: Gene expression-style analysis with precision weights and correlation structure.

## Total cohort - limma-voom
```{r abundance_limma_total}
library(limma)

# Prepare design matrix for total cohort
total_design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                            HLA.Category + feeding_first_year + Delivery.Mode, 
                            data = total.orf.metadata.clean_0.75_0.03)

# Apply voom transformation
total_voom <- voom(total.y, design = total_design, plot = FALSE)

# Estimate correlation for repeated measures
total_corfit <- duplicateCorrelation(total_voom, total_design, block = total.orf.metadata.clean_0.75_0.03$patientID)

# Fit linear model
total_fit <- lmFit(total_voom, total_design, block = total.orf.metadata.clean_0.75_0.03$patientID, 
                   correlation = total_corfit$consensus)
total_fit <- eBayes(total_fit)

# Extract results for all coefficients
total_limma_results <- topTable(total_fit, number = Inf, sort.by = "none") %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "coefficient", values_to = "value") %>%
  mutate(model = "limma_voom")

# Save results
write.csv(total_limma_results, "../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model4_limma_voom.csv", row.names = FALSE)
```

## US cohort - limma-voom
```{r abundance_limma_US}
# Prepare design matrix for US cohort
US_design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                         HLA.Category + feeding_first_year + Delivery.Mode, 
                         data = US.orf.metadata.clean_0.75_0.03)

# Apply voom transformation
US_voom <- voom(US.y, design = US_design, plot = FALSE)

# Estimate correlation for repeated measures
US_corfit <- duplicateCorrelation(US_voom, US_design, block = US.orf.metadata.clean_0.75_0.03$patientID)

# Fit linear model
US_fit <- lmFit(US_voom, US_design, block = US.orf.metadata.clean_0.75_0.03$patientID, 
                correlation = US_corfit$consensus)
US_fit <- eBayes(US_fit)

# Extract results for all coefficients
US_limma_results <- topTable(US_fit, number = Inf, sort.by = "none") %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "coefficient", values_to = "value") %>%
  mutate(model = "limma_voom")

# Save results
write.csv(US_limma_results, "../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model4_limma_voom.csv", row.names = FALSE)
```

## Italy cohort - limma-voom
```{r abundance_limma_Italy}
# Prepare design matrix for Italy cohort
Italy_design <- model.matrix(~ Dx.Status * onset_timeline_combined + Sex + Age.at.Gluten.Introduction..months. + 
                            HLA.Category + feeding_first_year + Delivery.Mode, 
                            data = Italy.orf.metadata.clean_0.75_0.03)

# Apply voom transformation
Italy_voom <- voom(Italy.y, design = Italy_design, plot = FALSE)

# Estimate correlation for repeated measures
Italy_corfit <- duplicateCorrelation(Italy_voom, Italy_design, block = Italy.orf.metadata.clean_0.75_0.03$patientID)

# Fit linear model
Italy_fit <- lmFit(Italy_voom, Italy_design, block = Italy.orf.metadata.clean_0.75_0.03$patientID, 
                   correlation = Italy_corfit$consensus)
Italy_fit <- eBayes(Italy_fit)

# Extract results for all coefficients
Italy_limma_results <- topTable(Italy_fit, number = Inf, sort.by = "none") %>%
  rownames_to_column("orf_id") %>%
  pivot_longer(cols = -orf_id, names_to = "coefficient", values_to = "value") %>%
  mutate(model = "limma_voom")

# Save results
write.csv(Italy_limma_results, "../Orf_Contig_Phrog_compositional/orf_results/Italy_abundance_model4_limma_voom.csv", row.names = FALSE)
```


# TIMEPOINT-SPECIFIC EXTRACTION FUNCTIONS

# Function to extract timepoint-specific differential abundance/PA results
```{r timepoint_extraction_functions}
# Function to extract timepoint-specific results from model output
extract_timepoint_results <- function(model_results, model_name, gene_col = "orf_id") {
  if(is.null(model_results) || nrow(model_results) == 0) {
    return(data.frame())
  }
  
  # Extract interaction terms for timepoints
  timepoint_results <- model_results %>%
    filter(grepl("Dx.StatusCELIAC:onset_timeline_combined", term)) %>%
    mutate(
      timepoint = gsub("Dx.StatusCELIAC:onset_timeline_combined", "", term),
      model = model_name,
      gene = get(gene_col)
    ) %>%
    select(gene, timepoint, estimate, std.error, p.value, conf.low, conf.high, model)
  
  return(timepoint_results)
}

# Function to apply extraction to all models for a cohort
extract_all_timepoints <- function(cohort_name) {
  results_list <- list()
  
  # PA models
  tryCatch({
    if(exists(paste0(cohort_name, "_glmmTMB_logit_results"))) {
      results_list[["PA_model1"]] <- extract_timepoint_results(
        get(paste0(cohort_name, "_glmmTMB_logit_results")), "PA_glmmTMB_logit")
    }
  }, error = function(e) NULL)
  
  # Abundance models
  tryCatch({
    if(exists(paste0(cohort_name, "_nb_results"))) {
      results_list[["abundance_model1"]] <- extract_timepoint_results(
        get(paste0(cohort_name, "_nb_results")), "abundance_nbinom")
    }
  }, error = function(e) NULL)
  
  # Combine all results
  if(length(results_list) > 0) {
    combined_results <- do.call(rbind, results_list)
    return(combined_results)
  } else {
    return(data.frame())
  }
}
```

# Extract timepoint-specific results for PA model 1 (main model)
```{r extract_PA_timepoints}
# Total cohort PA timepoints
if(exists("total_glmmTMB_logit_results")) {
  total_pa_timepoints <- extract_timepoint_results(total_glmmTMB_logit_results, "PA_glmmTMB_logit")
  write.csv(total_pa_timepoints, "../Orf_Contig_Phrog_compositional/orf_results/total_PA_model1_glmmTMB_logit_timepoint_specific_results.csv", row.names = FALSE)
}

# US cohort PA timepoints
if(exists("US_glmmTMB_logit_results")) {
  US_pa_timepoints <- extract_timepoint_results(US_glmmTMB_logit_results, "PA_glmmTMB_logit")
  write.csv(US_pa_timepoints, "../Orf_Contig_Phrog_compositional/orf_results/US_PA_model1_glmmTMB_logit_timepoint_specific_results.csv", row.names = FALSE)
}

# Italy cohort PA timepoints
if(exists("Italy_glmmTMB_logit_results")) {
  Italy_pa_timepoints <- extract_timepoint_results(Italy_glmmTMB_logit_results, "PA_glmmTMB_logit")
  write.csv(Italy_pa_timepoints, "../Orf_Contig_Phrog_compositional/orf_results/Italy_PA_model1_glmmTMB_logit_timepoint_specific_results.csv", row.names = FALSE)
}
```

# Extract timepoint-specific results for abundance model 1 (main model)
```{r extract_abundance_timepoints}
# Total cohort abundance timepoints
if(exists("total_nb_results")) {
  total_abundance_timepoints <- extract_timepoint_results(total_nb_results, "abundance_nbinom")
  write.csv(total_abundance_timepoints, "../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model1_nbinom_timepoint_specific_results.csv", row.names = FALSE)
}

# US cohort abundance timepoints
if(exists("US_nb_results")) {
  US_abundance_timepoints <- extract_timepoint_results(US_nb_results, "abundance_nbinom")
  write.csv(US_abundance_timepoints, "../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model1_nbinom_timepoint_specific_results.csv", row.names = FALSE)
}

# Italy cohort abundance timepoints
if(exists("Italy_nb_results")) {
  Italy_abundance_timepoints <- extract_timepoint_results(Italy_nb_results, "abundance_nbinom")
  write.csv(Italy_abundance_timepoints, "../Orf_Contig_Phrog_compositional/orf_results/Italy_abundance_model1_nbinom_timepoint_specific_results.csv", row.names = FALSE)
}
```


# COMPOSITIONAL ANALYSIS
# Trajectory-based analysis focusing on temporal patterns and individual slopes

```{r compositional_analysis_setup}
library(vegan)
library(ggplot2)
library(cowplot)

# Calculate diversity metrics for total cohort
calculate_diversity_metrics <- function(abundance_matrix, metadata) {
  # Transpose so samples are rows
  abundance_t <- t(abundance_matrix)
  
  # Calculate various diversity metrics
  diversity_metrics <- data.frame(
    sample_id = rownames(abundance_t),
    shannon = diversity(abundance_t, index = "shannon"),
    simpson = diversity(abundance_t, index = "simpson"),
    richness = specnumber(abundance_t),
    evenness = diversity(abundance_t, index = "shannon") / log(specnumber(abundance_t))
  ) %>%
    left_join(metadata %>% rownames_to_column("sample_id"), by = "sample_id")
  
  return(diversity_metrics)
}

# Total cohort diversity
total_diversity <- calculate_diversity_metrics(total.orf.abundance.clean_0.75_0.03, total.orf.metadata.clean_0.75_0.03)
write.csv(total_diversity, "../Orf_Contig_Phrog_compositional/orf_results/total_diversity_metrics.csv", row.names = FALSE)
```

# Individual trajectory analysis
```{r individual_trajectories}
# Calculate individual slopes for abundance changes over time
calculate_individual_slopes <- function(abundance_matrix, metadata, min_timepoints = 3) {
  # Convert to long format
  abundance_long <- abundance_matrix %>%
    as.data.frame() %>%
    rownames_to_column("orf_id") %>%
    pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "abundance") %>%
    left_join(metadata %>% rownames_to_column("sample_id"), by = "sample_id") %>%
    filter(!is.na(onset_timeline_combined))
  
  # Convert timepoint to numeric (months before diagnosis)
  abundance_long$timepoint_numeric <- case_when(
    abundance_long$onset_timeline_combined == "t0" ~ 0,
    abundance_long$onset_timeline_combined == "t0-6" ~ -6,
    abundance_long$onset_timeline_combined == "t0-12" ~ -12,
    abundance_long$onset_timeline_combined == "t0-18" ~ -18,
    abundance_long$onset_timeline_combined == "t0-24" ~ -24,
    abundance_long$onset_timeline_combined == "t0-30" ~ -30,
    abundance_long$onset_timeline_combined == "t0-36" ~ -36,
    abundance_long$onset_timeline_combined == "t0-over42" ~ -45,
    TRUE ~ NA_real_
  )
  
  # Calculate slopes for each patient-gene combination
  slopes <- abundance_long %>%
    group_by(patientID, orf_id, Dx.Status) %>%
    filter(n() >= min_timepoints) %>%
    do({
      model <- lm(log1p(abundance) ~ timepoint_numeric, data = .)
      data.frame(
        slope = coef(model)[2],
        r_squared = summary(model)$r.squared,
        p_value = summary(model)$coefficients[2, 4]
      )
    }) %>%
    ungroup()
  
  return(slopes)
}

# Calculate slopes for total cohort
total_slopes <- calculate_individual_slopes(total.orf.abundance.clean_0.75_0.03, total.orf.metadata.clean_0.75_0.03)
write.csv(total_slopes, "../Orf_Contig_Phrog_compositional/orf_results/total_individual_trajectories.csv", row.names = FALSE)
```

# Stability analysis
```{r stability_analysis}
# Calculate coefficient of variation (CV) for each gene across timepoints
calculate_stability_metrics <- function(abundance_matrix, metadata) {
  # Long format
  abundance_long <- abundance_matrix %>%
    as.data.frame() %>%
    rownames_to_column("orf_id") %>%
    pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "abundance") %>%
    left_join(metadata %>% rownames_to_column("sample_id"), by = "sample_id")
  
  # Calculate stability metrics by patient and gene
  stability <- abundance_long %>%
    group_by(patientID, orf_id, Dx.Status) %>%
    filter(n() >= 3) %>%  # At least 3 timepoints
    summarise(
      mean_abundance = mean(abundance, na.rm = TRUE),
      cv = sd(abundance, na.rm = TRUE) / mean(abundance, na.rm = TRUE),
      stability_score = 1 / (1 + cv),  # Higher = more stable
      .groups = 'drop'
    )
  
  return(stability)
}

# Calculate stability for total cohort
total_stability <- calculate_stability_metrics(total.orf.abundance.clean_0.75_0.03, total.orf.metadata.clean_0.75_0.03)
write.csv(total_stability, "../Orf_Contig_Phrog_compositional/orf_results/total_stability_metrics.csv", row.names = FALSE)
```

# Clustering analysis based on temporal patterns
```{r clustering_analysis}
# Cluster patients based on temporal abundance patterns
cluster_temporal_patterns <- function(abundance_matrix, metadata, n_clusters = 4) {
  # Calculate patient-level temporal summaries
  abundance_long <- abundance_matrix %>%
    as.data.frame() %>%
    rownames_to_column("orf_id") %>%
    pivot_longer(cols = -orf_id, names_to = "sample_id", values_to = "abundance") %>%
    left_join(metadata %>% rownames_to_column("sample_id"), by = "sample_id")
  
  # Create patient-timepoint matrix
  patient_timepoint <- abundance_long %>%
    group_by(patientID, onset_timeline_combined, Dx.Status) %>%
    summarise(total_abundance = sum(abundance, na.rm = TRUE), .groups = 'drop') %>%
    pivot_wider(names_from = onset_timeline_combined, values_from = total_abundance, values_fill = 0)
  
  # Perform clustering
  cluster_data <- patient_timepoint %>%
    select(-patientID, -Dx.Status) %>%
    as.matrix()
  
  # Handle missing values
  cluster_data[is.na(cluster_data)] <- 0
  
  # K-means clustering
  if(nrow(cluster_data) > n_clusters) {
    kmeans_result <- kmeans(cluster_data, centers = n_clusters, nstart = 25)
    
    # Add cluster assignments back
    patient_timepoint$cluster <- kmeans_result$cluster
    
    return(list(
      patient_clusters = patient_timepoint,
      cluster_centers = kmeans_result$centers,
      within_ss = kmeans_result$tot.withinss
    ))
  } else {
    return(NULL)
  }
}

# Perform clustering for total cohort
total_clusters <- cluster_temporal_patterns(total.orf.abundance.clean_0.75_0.03, total.orf.metadata.clean_0.75_0.03)
if(!is.null(total_clusters)) {
  write.csv(total_clusters$patient_clusters, "../Orf_Contig_Phrog_compositional/orf_results/total_patient_clusters.csv", row.names = FALSE)
}
```


# VISUALIZATION FUNCTIONS

```{r visualization_functions}
# Create visualizations for compositional analysis

# 1. Diversity by diagnosis and timepoint
create_diversity_plots <- function(diversity_data, output_prefix) {
  # By diagnosis
  p1 <- ggplot(diversity_data, aes(x = Dx.Status, y = shannon, fill = Dx.Status)) +
    geom_boxplot() +
    geom_jitter(width = 0.2, alpha = 0.6) +
    labs(title = "Shannon Diversity by Diagnosis",
         x = "Diagnosis Status", y = "Shannon Diversity") +
    theme_minimal() +
    theme(legend.position = "none")
  
  # By timepoint
  p2 <- ggplot(diversity_data, aes(x = onset_timeline_combined, y = shannon, fill = Dx.Status)) +
    geom_boxplot() +
    labs(title = "Shannon Diversity by Timepoint",
         x = "Timepoint", y = "Shannon Diversity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Save plots
  ggsave(paste0("../Orf_Contig_Phrog_compositional/orf_figures/", output_prefix, "_shannon_diversity_by_diagnosis.pdf"), p1, width = 8, height = 6)
  ggsave(paste0("../Orf_Contig_Phrog_compositional/orf_figures/", output_prefix, "_shannon_diversity_by_timepoint.pdf"), p2, width = 10, height = 6)
  
  return(list(by_diagnosis = p1, by_timepoint = p2))
}

# Create plots for total cohort
if(exists("total_diversity")) {
  diversity_plots <- create_diversity_plots(total_diversity, "total")
}
```


# CREATE ANALYSIS SUMMARY

```{r analysis_summary}
# Create summary of all model results
create_analysis_summary <- function() {
  model_files <- list.files("../Orf_Contig_Phrog_compositional/orf_results/", pattern = "\\.csv$", full.names = TRUE)
  
  summary_data <- data.frame()
  
  for(file in model_files) {
    if(grepl("timepoint_specific", file)) next  # Skip timepoint files for main summary
    
    tryCatch({
      data <- read.csv(file)
      
      if(nrow(data) > 0 && "p.value" %in% colnames(data)) {
        file_summary <- data.frame(
          file = basename(file),
          cohort = gsub("_.*", "", basename(file)),
          model_type = ifelse(grepl("PA", basename(file)), "PA", "Abundance"),
          total_results = nrow(data),
          significant_results = sum(data$p.value < 0.05, na.rm = TRUE),
          valid_pvalues = sum(!is.na(data$p.value)),
          convergence_rate = round(sum(!is.na(data$p.value)) / nrow(data) * 100, 1)
        )
        
        summary_data <- rbind(summary_data, file_summary)
      }
    }, error = function(e) {
      cat("Error processing file:", file, "\n")
    })
  }
  
  return(summary_data)
}

# Generate analysis summary
analysis_summary <- create_analysis_summary()
write.csv(analysis_summary, "../Orf_Contig_Phrog_compositional/orf_results/analysis_summary.csv", row.names = FALSE)

# Print summary
if(exists("analysis_summary") && nrow(analysis_summary) > 0) {
  print("Analysis Summary:")
  print(analysis_summary)
}
```


# COMPLETE VISUALIZATION PIPELINE 
# Adding all the comprehensive visualizations from the contig analysis

```{r basic_visualizations}
library(ggplot2)
library(cowplot)
library(viridis)

# Function to create effect size plots for all models
create_effect_size_plots <- function(timepoint_file, output_name, title_prefix) {
  if(!file.exists(timepoint_file)) {
    cat("File not found:", timepoint_file, "\n")
    return(NULL)
  }
  
  # Load timepoint results
  data <- read.csv(timepoint_file)
  
  if(nrow(data) == 0) {
    cat("No data in file:", timepoint_file, "\n")
    return(NULL)
  }
  
  cat("Creating effect size plot for", nrow(data), "results from", basename(timepoint_file), "\n")
  
  # Prepare data for plotting
  data$has_pvalue <- !is.na(data$p.value)
  data$neg_log10_p <- ifelse(data$has_pvalue, -log10(data$p.value), NA)
  data$is_significant <- data$has_pvalue & data$p.value < 0.05
  
  # Traditional volcano plot (only valid p-values)
  data_valid <- data[data$has_pvalue, ]
  
  if(nrow(data_valid) > 0) {
    p1 <- ggplot(data_valid, aes(x = estimate, y = neg_log10_p)) +
      geom_point(aes(color = is_significant, shape = timepoint), 
                 alpha = 0.7, size = 2) +
      scale_color_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#95A5A6"),
                         labels = c("TRUE" = "Significant (p<0.05)", "FALSE" = "Not significant"),
                         name = "Significance") +
      labs(title = paste(title_prefix, "- Effect Sizes"),
           subtitle = paste0("Showing ", nrow(data_valid), " out of ", nrow(data), " results with valid p-values"),
           x = "Effect Size (Log Odds/Fold Change)",
           y = "-log10(p-value)") +
      theme_minimal() +
      theme(legend.position = "bottom",
            plot.title = element_text(size = 11),
            plot.subtitle = element_text(size = 9, color = "gray60"))
  } else {
    # Create empty plot if no valid p-values
    p1 <- ggplot() + 
      annotate("text", x = 0, y = 0, 
               label = "No valid p-values found\n(All models failed to converge)", 
               size = 5, color = "red") +
      labs(title = paste(title_prefix, "- Effect Sizes"),
           subtitle = "No convergent models found",
           x = "Effect Size", y = "-log10(p-value)") +
      theme_void()
  }
  
  # Save effect size plot
  output_file <- paste0("../Orf_Contig_Phrog_compositional/orf_figures/", output_name, "_effect_sizes.pdf")
  ggsave(output_file, p1, width = 10, height = 8, device = "pdf")
  
  cat("Effect size plot saved:", basename(output_file), "\n")
  
  return(p1)
}

# Create effect size plots for main models (if timepoint results exist)
tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/total_PA_model1_glmmTMB_logit_timepoint_specific_results.csv")) {
    create_effect_size_plots("../Orf_Contig_Phrog_compositional/orf_results/total_PA_model1_glmmTMB_logit_timepoint_specific_results.csv", 
                           "total_PA_model1_glmmTMB_logit", "Total PA Model")
  }
}, error = function(e) cat("Error creating total PA effect size plot\n"))

tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model1_nbinom_timepoint_specific_results.csv")) {
    create_effect_size_plots("../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model1_nbinom_timepoint_specific_results.csv", 
                           "total_abundance_model1_nbinom", "Total Abundance Model")
  }
}, error = function(e) cat("Error creating total abundance effect size plot\n"))

tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/US_PA_model1_glmmTMB_logit_timepoint_specific_results.csv")) {
    create_effect_size_plots("../Orf_Contig_Phrog_compositional/orf_results/US_PA_model1_glmmTMB_logit_timepoint_specific_results.csv", 
                           "US_PA_model1_glmmTMB_logit", "US PA Model")
  }
}, error = function(e) cat("Error creating US PA effect size plot\n"))

tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model1_nbinom_timepoint_specific_results.csv")) {
    create_effect_size_plots("../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model1_nbinom_timepoint_specific_results.csv", 
                           "US_abundance_model1_nbinom", "US Abundance Model")
  }
}, error = function(e) cat("Error creating US abundance effect size plot\n"))
```

```{r heatmap_visualizations}
library(pheatmap)
library(RColorBrewer)
library(reshape2)

# Define correct timepoint order
timepoint_order <- c("t0", "t0-6", "t0-12", "t0-18", "t0-24", "t0-30", "t0-36", "t0-over42")

# Function to create heatmap
create_heatmap <- function(timepoint_file, output_name, title_prefix) {
  if(!file.exists(timepoint_file)) {
    cat("File not found:", timepoint_file, "\n")
    return(NULL)
  }
  
  # Load timepoint results
  data <- read.csv(timepoint_file)
  
  if(nrow(data) == 0) {
    cat("No data in file:", timepoint_file, "\n")
    return(NULL)
  }
  
  cat("Processing", nrow(data), "results from", basename(timepoint_file), "\n")
  
  # Filter for significant results (p < 0.05) and create matrix
  sig_data <- data %>%
    filter(p.value < 0.05) %>%
    select(gene, timepoint, estimate, p.value)
  
  if(nrow(sig_data) == 0) {
    cat("No significant results found for", output_name, "\n")
    return(NULL)
  }
  
  # Create matrix for heatmap (genes x timepoints)
  estimate_matrix <- sig_data %>%
    select(gene, timepoint, estimate) %>%
    dcast(gene ~ timepoint, value.var = "estimate", fill = 0)
  
  # Convert to matrix
  rownames(estimate_matrix) <- estimate_matrix$gene
  estimate_matrix <- estimate_matrix[, -1, drop = FALSE]
  estimate_matrix <- as.matrix(estimate_matrix)
  
  # Remove genes with all zeros
  estimate_matrix <- estimate_matrix[rowSums(abs(estimate_matrix)) > 0, , drop = FALSE]
  
  if(nrow(estimate_matrix) == 0) {
    cat("No genes with non-zero estimates for", output_name, "\n")
    return(NULL)
  }
  
  # Reorder columns to match timepoint order
  available_timepoints <- intersect(timepoint_order, colnames(estimate_matrix))
  if(length(available_timepoints) > 0) {
    estimate_matrix <- estimate_matrix[, available_timepoints, drop = FALSE]
  }
  
  cat("Creating heatmap with", nrow(estimate_matrix), "genes and", ncol(estimate_matrix), "timepoints\n")
  
  # Calculate dimensions
  n_genes <- nrow(estimate_matrix)
  n_timepoints <- ncol(estimate_matrix)
  
  cell_height <- 0.3
  cell_width <- 0.8
  
  fig_height <- max(4, min(12, n_genes * cell_height + 2))
  fig_width <- max(6, n_timepoints * cell_width + 3)
  
  # Create heatmap
  pdf(paste0("../Orf_Contig_Phrog_compositional/orf_figures/", output_name, "_heatmap.pdf"), 
      width = fig_width, height = fig_height)
  
  pheatmap(estimate_matrix,
           cluster_rows = TRUE,
           cluster_cols = FALSE,
           scale = "none",
           color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
           breaks = seq(-max(abs(estimate_matrix)), max(abs(estimate_matrix)), length.out = 101),
           main = paste(title_prefix, "- Log Odds/Fold Change (CELIAC vs CONTROL)"),
           fontsize = 10,
           fontsize_row = 8,
           fontsize_col = 10,
           show_rownames = TRUE,
           border_color = "white",
           cellwidth = 25,
           cellheight = 12,
           treeheight_row = 30,
           treeheight_col = 0,
           angle_col = 45)
  
  dev.off()
  
  return(estimate_matrix)
}

# Create heatmaps for main models
tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/total_PA_model1_glmmTMB_logit_timepoint_specific_results.csv")) {
    create_heatmap("../Orf_Contig_Phrog_compositional/orf_results/total_PA_model1_glmmTMB_logit_timepoint_specific_results.csv", 
                  "total_PA_model1_glmmTMB_logit", "Total PA Model")
  }
}, error = function(e) cat("Error creating total PA heatmap\n"))

tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model1_nbinom_timepoint_specific_results.csv")) {
    create_heatmap("../Orf_Contig_Phrog_compositional/orf_results/total_abundance_model1_nbinom_timepoint_specific_results.csv", 
                  "total_abundance_model1_nbinom", "Total Abundance Model")
  }
}, error = function(e) cat("Error creating total abundance heatmap\n"))

tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/US_PA_model1_glmmTMB_logit_timepoint_specific_results.csv")) {
    create_heatmap("../Orf_Contig_Phrog_compositional/orf_results/US_PA_model1_glmmTMB_logit_timepoint_specific_results.csv", 
                  "US_PA_model1_glmmTMB_logit", "US PA Model")
  }
}, error = function(e) cat("Error creating US PA heatmap\n"))

tryCatch({
  if(file.exists("../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model1_nbinom_timepoint_specific_results.csv")) {
    create_heatmap("../Orf_Contig_Phrog_compositional/orf_results/US_abundance_model1_nbinom_timepoint_specific_results.csv", 
                  "US_abundance_model1_nbinom", "US Abundance Model")
  }
}, error = function(e) cat("Error creating US abundance heatmap\n"))
```

```{r summary_visualizations}
# Create analysis summary barplot
create_summary_barplot <- function() {
  tryCatch({
    summary_data <- read.csv("../Orf_Contig_Phrog_compositional/orf_results/analysis_summary.csv")
    
    if(nrow(summary_data) > 0) {
      # Create barplot of model success rates
      p <- ggplot(summary_data, aes(x = file, y = convergence_rate, fill = model_type)) +
        geom_bar(stat = "identity") +
        geom_text(aes(label = paste0(convergence_rate, "%")), 
                  vjust = -0.5, size = 3) +
        labs(title = "Model Convergence Rates by Cohort and Model Type",
             subtitle = "ORF-based Analysis Results",
             x = "Model Files", 
             y = "Convergence Rate (%)",
             fill = "Model Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              plot.title = element_text(size = 14, face = "bold")) +
        scale_fill_viridis_d()
      
      ggsave("../Orf_Contig_Phrog_compositional/orf_figures/analysis_summary_barplot.pdf", 
             p, width = 12, height = 8)
      
      cat("Summary barplot created\n")
      return(p)
    }
  }, error = function(e) {
    cat("Error creating summary barplot:", e$message, "\n")
  })
}

# Create summary barplot
create_summary_barplot()
```

```{r improved_effect_plots}
# Create improved effect size visualizations that show the full picture
create_improved_effect_plots <- function(timepoint_file, output_name, title_prefix) {
  if(!file.exists(timepoint_file)) {
    cat("File not found:", timepoint_file, "\n")
    return(NULL)
  }
  
  # Load timepoint results
  data <- read.csv(timepoint_file)
  
  if(nrow(data) == 0) {
    cat("No data in file:", timepoint_file, "\n")
    return(NULL)
  }
  
  cat("Processing", nrow(data), "results from", basename(timepoint_file), "\n")
  
  # Prepare data for plotting
  data$has_pvalue <- !is.na(data$p.value)
  data$neg_log10_p <- ifelse(data$has_pvalue, -log10(data$p.value), NA)
  data$is_significant <- data$has_pvalue & data$p.value < 0.05
  
  # Summary statistics
  n_total <- nrow(data)
  n_with_pvalue <- sum(data$has_pvalue)
  n_significant <- sum(data$is_significant, na.rm = TRUE)
  
  cat("Data summary for", output_name, ":\n")
  cat("  Total results:", n_total, "\n")
  cat("  With valid p-values:", n_with_pvalue, "(", round(100*n_with_pvalue/n_total, 1), "%)\n")
  cat("  Significant (p<0.05):", n_significant, "\n")
  cat("  Effect size range:", round(min(data$estimate, na.rm=TRUE), 2), "to", round(max(data$estimate, na.rm=TRUE), 2), "\n")
  
  # Create multiple plots to show the full picture
  
  # Plot 1: All effect sizes (including NAs)
  p1 <- ggplot(data, aes(x = estimate, y = timepoint)) +
    geom_point(aes(color = has_pvalue, shape = is_significant), 
               alpha = 0.7, size = 2) +
    scale_color_manual(values = c("TRUE" = "#2E86AB", "FALSE" = "#F24236"),
                       labels = c("TRUE" = "Valid p-value", "FALSE" = "No p-value (NA)"),
                       name = "Statistical Status") +
    scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                       labels = c("TRUE" = "Significant", "FALSE" = "Not significant"),
                       name = "Significance",
                       na.value = 4) +
    labs(title = paste(title_prefix, "- All Effect Sizes"),
         subtitle = paste0("Total: ", n_total, " | Valid p-values: ", n_with_pvalue, 
                          " (", round(100*n_with_pvalue/n_total, 1), "%)"),
         x = "Effect Size (Log Odds/Fold Change)",
         y = "Timepoint") +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 11),
          plot.subtitle = element_text(size = 9, color = "gray60")) +
    guides(color = guide_legend(override.aes = list(size = 3)),
           shape = guide_legend(override.aes = list(size = 3)))
  
  # Plot 2: Distribution of effect sizes
  p2 <- ggplot(data, aes(x = estimate)) +
    geom_histogram(aes(fill = has_pvalue), bins = 30, alpha = 0.7, position = "stack") +
    scale_fill_manual(values = c("TRUE" = "#2E86AB", "FALSE" = "#F24236"),
                      labels = c("TRUE" = "Valid p-value", "FALSE" = "No p-value (NA)"),
                      name = "Statistical Status") +
    labs(title = "Distribution of Effect Sizes",
         subtitle = "Red bars show large effects with no p-values (model convergence issues)",
         x = "Effect Size (Log Odds/Fold Change)",
         y = "Count") +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 11),
          plot.subtitle = element_text(size = 9, color = "gray60"))
  
  # Plot 3: Traditional volcano plot (only valid p-values)
  data_valid <- data[data$has_pvalue, ]
  
  if(nrow(data_valid) > 0) {
    p3 <- ggplot(data_valid, aes(x = estimate, y = neg_log10_p)) +
      geom_point(aes(color = is_significant, shape = timepoint), 
                 alpha = 0.7, size = 2) +
      scale_color_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#95A5A6"),
                         labels = c("TRUE" = "Significant (p<0.05)", "FALSE" = "Not significant"),
                         name = "Significance") +
      labs(title = "Traditional Volcano Plot (Valid P-values Only)",
           subtitle = paste0("Showing ", nrow(data_valid), " out of ", n_total, " results"),
           x = "Effect Size (Log Odds/Fold Change)",
           y = "-log10(p-value)") +
      theme_minimal() +
      theme(legend.position = "bottom",
            plot.title = element_text(size = 11),
            plot.subtitle = element_text(size = 9, color = "gray60"))
  } else {
    # Create empty plot if no valid p-values
    p3 <- ggplot() + 
      annotate("text", x = 0, y = 0, 
               label = "No valid p-values found\n(All models failed to converge)", 
               size = 5, color = "red") +
      labs(title = "Traditional Volcano Plot (Valid P-values Only)",
           subtitle = "No convergent models found",
           x = "Effect Size", y = "-log10(p-value)") +
      theme_void()
  }
  
  # Combine plots
  combined_plot <- plot_grid(p1, p2, p3, 
                            ncol = 1, 
                            rel_heights = c(1, 0.8, 1),
                            align = "v")
  
  # Save improved effect size plot
  output_file <- paste0("../Orf_Contig_Phrog_compositional/orf_figures/", output_name, "_improved_effects.pdf")
  ggsave(output_file, combined_plot, width = 10, height = 12, device = "pdf")
  
  cat("Improved effect size plot saved:", basename(output_file), "\n")
  
  return(list(
    total_results = n_total,
    valid_pvalues = n_with_pvalue,
    significant = n_significant,
    effect_range = c(min(data$estimate, na.rm=TRUE), max(data$estimate, na.rm=TRUE))
  ))
}

# Create improved effect plots for main models (to be run after timepoint extraction)
cat("Improved effect size visualization functions loaded\n")
```


# CREATE DIRECTORIES AND FINAL SETUP

```{r create_directories}
# Create output directories if they don't exist
dir.create("../Orf_Contig_Phrog_compositional", showWarnings = FALSE, recursive = TRUE)
dir.create("../Orf_Contig_Phrog_compositional/orf_results", showWarnings = FALSE, recursive = TRUE)
dir.create("../Orf_Contig_Phrog_compositional/orf_figures", showWarnings = FALSE, recursive = TRUE)
dir.create("../Orf_Contig_Phrog_compositional/orf_report_figures", showWarnings = FALSE, recursive = TRUE)

cat("ORF-based PA and abundance analysis completed!\n")
cat("Results saved to: ../Orf_Contig_Phrog_compositional/orf_results/\n")
cat("Figures saved to: ../Orf_Contig_Phrog_compositional/orf_figures/\n")
```
